---
title: "rbt_1_5"
output: html_document
date: "2026-02-21"
---

Author: Wei (Adelyn) Tsai; [tsai.wei\@mayo.edu](mailto:tsai.wei@mayo.edu){.email} If you are using this script, please cite our study This script addresses comment 5 about genetic support from reviewer 1. Chunks with bash are done in linux.

```{r set up}
suppressPackageStartupMessages(library(tidyverse)) #v2.0.0
suppressPackageStartupMessages(library(openxlsx)) #v4.2.6.1
suppressPackageStartupMessages(library(data.table)) #v1.15.4
suppressPackageStartupMessages(library(pbapply)) #v1.7
suppressPackageStartupMessages(library(ggh4x)) #v0.2.8
suppressPackageStartupMessages(library(broom)) #v1.0.6
suppressPackageStartupMessages(library(coloc)) #5.2.3
suppressPackageStartupMessages(library(TwoSampleMR)) #0.6.22

# list of input/output directories
indir <- "./indir"
residir <- "./outdir/resids"
blwgcna_dir <- "./outdir/bl_wgcna"
nduf_assoc_dir <- "./outdir/nduf_assoc_de"

outdir <- "./Rebuttal/outdir"

# source commonly used functions
source("./Codes/fns.R")
```

# eQTL files prep

## 1. Generate residuals adjusting for technical covariates

```{r}
cqn_raw <- read_delim(paste0(indir, "/PaX108_R01resilience_gene_CQN_neg3_postQC_s105.txt"), delim = "\t") # MCSA cqn file for 18046 genes. Each row is a gene and each column is a donor
adni_mc <- readRDS(paste0(indir,"/ADNI_s91_microarray.rds")) #microarray file for 91 donors from ADNI. This has 10116 protein-coding genes, gene id as column names, donorID as row names

# get the intersecting genes of two expression files
int_genes <- intersect(subset(cqn_raw, GeneBiotype == "protein_coding")$GeneId, colnames(adni_mc))

# MCSA
mcsa_covars <- read_delim(paste0(indir,"/R01_Resilience_PAXgene_Covars.txt"), delim = "\t") %>% mutate(ptnum=as.character(ptnum)) #covariates file for MCSA, rows are donor columns are covariates
mcsa_covars <- mcsa_covars %>% 
  mutate(
    PAXgene_flowcell = factor(PAXgene_flowcell), 
    PAXgene_Batch = factor(PAXgene_Batch) # factorize flowcell and batch
  )
## generate residuals
mcsa_res_eqtl <- cqn_raw %>%
  filter(GeneId %in% int_genes) %>% 
  select(-Chromosome, -Start, -End, -Length, -GeneName, -GeneBiotype) %>%
  column_to_rownames("GeneId") %>% 
  t() %>% 
  as.data.frame() %>%
  rownames_to_column("SubjectID") %>%
  pivot_longer(!SubjectID, names_to = "Gene", values_to = "GenExpr") %>% 
  inner_join(mcsa_covars, by=c("SubjectID" = "ptnum")) %>% 
  group_by(Gene) %>%
  do(data.frame(., resid = residuals(lm(GenExpr ~ RIN + PAXgene_flowcell + PAXgene_Batch, data=.)))) %>% 
  dplyr::select(SubjectID, Gene, resid) %>% pivot_wider(names_from = Gene, values_from = resid)

adni_covars <- read_delim(paste0(residir, "/ADNI_s91_covars.txt"), delim = "\t") # a covariate file for ADNI, generated in 1_CreateResids.R
adni_covars <- adni_covars %>% mutate(PTGENDER = factor(PTGENDER))
## generate residuals
adni_res_eqtl <- adni_mc %>% 
  rownames_to_column("SubjectID") %>%
  pivot_longer(!SubjectID, names_to = "Gene", values_to = "GenExpr") %>% 
  inner_join(adni_covars) %>% 
  drop_na() %>% 
  group_by(Gene) %>%
  do(data.frame(., resid = residuals(lm(GenExpr ~ RIN + AffyPlate.meanREE + SITE.meanREE, data=.)))) %>% 
  dplyr::select(SubjectID, Gene, resid) %>% pivot_wider(names_from = Gene, values_from = resid)

# save the residuals for later use
save(mcsa_res_eqtl, adni_res_eqtl, file = paste0(outdir, "/1_5/Resilience_Residuals_eQTL.RData"))
```

## 2. eQTL covariates files

```{r}
# MCSA 
mcsa_covars_eqtl <- read_delim(paste0(indir, "/mcsa_eqtl_covar.txt"), delim = "\t") # a file with eqtl IID
mcsa_covars_eqtl %>% 
  mutate(ptnum = as.character(ptnum)) %>% 
  dplyr::select(FID:ptnum) %>% 
  inner_join(mcsa_covars %>% dplyr::select(ptnum, Sex, PAXgene_Age) %>% dplyr::rename(Age = PAXgene_Age)) %>% 
  write_delim(paste0(outdir, "/1_5/mcsa_covars_eqtl.txt"), delim = "\t") # include id, sex, age at blood collection as covariates
# ADNI
adni_covars %>% 
  dplyr::select(SubjectID, PTGENDER, AGE_EXAM) %>% 
  dplyr::rename(Sex = PTGENDER, Age = AGE_EXAM)  %>% 
  write_delim(paste0(outdir, "/1_5/adni_covars_eqtl.txt"), delim = "\t") # include id, sex, age at blood collection as covariates
```

# Examine eQTL results

eQTL was run using this script: rbt_1_5_run_cisQTL_NDUFs_021726.R
PLINK meta-analysis: BETA in both input and output are with respect to the A1 allele. I.e. if it's greater than 1, that implies the A1 allele increases risk relative to A2.
```{r}
ndufs_eqtl <- readRDS(paste0(outdir, "/1_5/ADNI_R01_Resilience_Blood_cisQTL_metaAnalysis_021726.rds"))
# Adjust the results by FDR
ndufs_eqtl <- ndufs_eqtl %>% mutate(FDR = p.adjust(`P(R)`, method = "fdr"), .after = "P(R)") # A1=effect allele
sig_eqtl <- ndufs_eqtl %>% filter(FDR < 0.05) #810
saveRDS(sig_eqtl, paste0(outdir, "/1_5/sig_eqtl.rds"))
# sort the two allels based on lexicographic order 
# we did this so we can have consistent naming of the SNPs across multiple results
a_low <- pmin(sig_eqtl$A1, sig_eqtl$A2)
a_high <- pmax(sig_eqtl$A1, sig_eqtl$A2)
sig_eqtl <- sig_eqtl %>% 
  mutate(SNP_rename = paste0(sig_eqtl$CHR, ":", sig_eqtl$BP, ":", a_low, ":", a_high), .after = "SNP")#rename SNP field to be consistent with bim files
rm(a_low, a_high)

sig_eqtl %>% 
  dplyr::select(SNP_rename, `P(R)`) %>% 
  rename(SNP = SNP_rename, P = `P(R)`) %>% 
  write_delim(paste0(outdir, "/1_5/sig_eqtl_forCLUMP.txt"), delim = "\t")
```

Next, we clumped the significant eQTL to identify significant lead eQTL. We used the LD reference panel from 1000 genome project prepared in 1_10_PRSv3.Rmd

```{bash clump}
ld_dir="./indir/LDRef_PLINK2" # the LD files with updated variant IDs are done in rbt_1_10_PRSv3.Rmd
outdir="./Rebuttal/outdir/1_5"

plink \
  --bfile "${ld_dir}/EUR.Ref.nors.lexicoID" \
  --clump "${outdir}/sig_eqtl_forCLUMP.txt" \
  --clump-r2 0.001 \
  --clump-kb 10000 \
  --clump-p1 1 \
  --clump-p2 1 \
  --out "${outdir}/sig_eqtl_clumpV3" # the r2 and kb parameters are from Dr.Winham's paper; also in this paper: https://pmc.ncbi.nlm.nih.gov/articles/PMC12462376/; and in TwoSampleMR: https://mrcieu.github.io/TwoSampleMR/reference/clump_data.html; # use this results
awk 'NR>1 {print $3}' "${outdir}/sig_eqtl_clumpV3.clumped" > "${outdir}/lead_sigeqtlV3.txt" #write the 3rd column, SNP, to a file for extracting them later
```

Then, we acquired genotype information for those lead eQTL from MCSA and ADNI,

```{bash recode-A}
wgs_dir="./Rebuttal/outdir/1_10/plink_outs" # the WGS files with updated variant IDs are done in rbt_1_10_PRSv3.Rmd

plink \
  --bfile "${wgs_dir}/ADNI_WGS_CSHRP_GBHNF_UpdatedIIDs_HG38_UpdatedVarID" \
  --recode Av \
  --allow-extra-chr \
  --extract "${outdir}/lead_sigeqtlV3.txt" \
  --out "${outdir}/lead_sigeqtl_geno_ADNI_V3"
  
plink \
  --bfile "${wgs_dir}/MCSA_WGS_CHR1_22_UpdatedVarID" \
  --recode Av \
  --allow-extra-chr \
  --extract "${outdir}/lead_sigeqtlV3.txt" \
  --out "${outdir}/lead_sigeqtl_geno_MCSA_V3"
```



## Examined the lead sig eQTL and the genotype file
.traw: COUNTED	Counted allele (defaults to A1)
```{r}
clump_result <- read.table(paste0(outdir, "/1_5/sig_eqtl_clumpV3.clumped"), header = T, stringsAsFactors = F, quote = "") #11 SNPs
# filter sig_eqtl dataframe to contain only the clumped independent lead SNPs
clumped_eqtl <- sig_eqtl %>% filter(SNP_rename %in% clump_result$SNP)
write.xlsx(clumped_eqtl, paste0(outdir, "/1_5/clumpled_eqtl.xlsx"))

# organize the traw files of MCSA and ADNI to long format for later use;
# also rename rsid that's NA manually
mcsa_traw <- read.table(paste0(outdir, "/1_5/lead_sigeqtl_geno_MCSA_V3.traw"), header = T, stringsAsFactors = F, quote = "") %>% 
  rename_with(.cols = 7:ncol(.), ~ gsub("^X0_", "", .)) %>% 
  #mutate(across(7:ncol(.), ~ factor(.))) %>% 
  pivot_longer(cols = 7:ncol(.), names_to = "IID", values_to = "Dose") %>% 
  left_join(clumped_eqtl %>% 
               dplyr::select(SNP_rename, rsid, gene_name), by=c("SNP" = "SNP_rename")) %>% 
  dplyr::select(CHR, SNP, rsid, gene_name, everything())
unique(mcsa_traw$SNP[which(is.na(mcsa_traw$rsid))]) # "11:78067306:A:G" -- these SNPs' rsid are NA
# 11:78067306:A:G = rs634768 according to dbSNP
mcsa_traw <- mcsa_traw %>% mutate(rsid = case_when(SNP == "11:78067306:A:G" ~ "rs634768", TRUE ~ rsid))
  
adni_traw <- read.table(paste0(outdir, "/1_5/lead_sigeqtl_geno_ADNI_V3.traw"), header = T, stringsAsFactors = F, quote = "") %>% 
  rename_with(.cols = 7:ncol(.), ~ gsub("^X0_", "", .)) %>% 
  #mutate(across(7:ncol(.), ~ factor(.))) %>% 
  pivot_longer(cols = 7:ncol(.), names_to = "IID", values_to = "Dose") %>% 
  left_join(clumped_eqtl %>% 
               dplyr::select(SNP_rename, rsid, gene_name), by=c("SNP" = "SNP_rename")) %>% 
  dplyr::select(CHR, SNP, rsid, gene_name, everything()) %>% 
  mutate(rsid = case_when(SNP == "11:78067306:A:G" ~ "rs634768", TRUE ~ rsid))

```


## Check against Bellenguez GWAS summary stats

Check if any of our significant eQTL is in significant loci highlighted in Bellenguez

```{r Bellenguez results}
st5 <- read.xlsx(paste0(indir, "/Bellenguez/harmonized/Bellenguez_ST5.xlsx")) # from Bellenguez et al supplementary table 5. these are loci reaching genome-wide significance
st5 <- st5 %>% mutate(CHR_POS = paste0(Chr, ":", Position), .after = "Position")

bellenguez_h <- data.table::fread(paste0(indir, "/Bellenguez/harmonized/35379992-GCST90027158-MONDO_0004975.h.tsv.gz")) # bellenguez harmonized summary stats from GWAS catalog

# check whether any of the significant cis-eqtl from our study is 500kb from any of the AD GWAS SNPs in Bellenguez
bell_sig_summ <- bellenguez_h %>% filter(hm_rsid %in% st5$Variant)

eqtl_adrisk = list()
for (i in 1:nrow(bell_sig_summ)){
  chr = bell_sig_summ$hm_chrom[i]
  pos = bell_sig_summ$hm_pos[i]
  rsid = bell_sig_summ$hm_rsid[i]
  
  upper = pos + 500000
  lower = pos - 500000
  
  sub_eqtl = sig_eqtl %>% filter(CHR == chr) %>% filter(BP <= upper & BP >= lower)
  
  eqtl_adrisk[[rsid]] <- sub_eqtl
}
eqtl_adrisk <- bind_rows(eqtl_adrisk, .id = "ADRisk_rsid")
write.xlsx(eqtl_adrisk, paste0(outdir, "/1_5/Sig_eQTL_ADRisk.xlsx")) # the results are organized in first rebuttal Rebuttal Table 7

rm(i, chr, pos, rsid, upper, lower, sub_eqtl)
```

# Colocalization
We found SNPs in NDUFB1 and NDUFA9 that are 500kb from AD GWAS SNPs in Bellenguez, so we performed colocalization.
All the colocalization results are organized in the first rebuttal Rebuttal Table 8
## NDUFB1, rs7401792
```{r}
chr <- 14
pos <- 92464917 # position of rs7401792

# acquire all SNPs in our eQTL study and in Bellenguez that are 500kb from rs740192
eqtl_ndufb1 <- ndufs_eqtl %>% filter(CHR == chr & BP >= pos - 500000 & BP <= pos + 500000) %>% filter(gene_name == "NDUFB1")
bell_ndufb1 <- bellenguez_h %>% filter(hm_chrom == chr & hm_pos >= pos - 500000 & hm_pos <= pos + 500000)
bell_ndufb1$SNP <- paste0(bell_ndufb1$hm_chrom, ":", bell_ndufb1$hm_pos)

#Get only SNPs present in both datasets
overlap <- intersect(eqtl_ndufb1$SNP, bell_ndufb1$SNP) #3182 SNPs
eqtl_ndufb1 <- eqtl_ndufb1 %>% filter(SNP %in% overlap)
bell_ndufb1 <- bell_ndufb1 %>% filter(SNP %in% overlap)

#Make sure the alleles are aligned in direction of effect
#Aligning the GWAS betas to the tested alleles of the eQTL
ndufb1_aligned <- bell_ndufb1 %>% 
  mutate(varbeta = standard_error^2, N = n_cas + n_con) %>% 
  dplyr::select(SNP, hm_rsid, hm_chrom, hm_pos, hm_effect_allele, hm_other_allele, hm_beta, varbeta, N) %>% 
  left_join(
    eqtl_ndufb1 %>% 
      mutate(N = 193) %>% #102 people in MCSA and 91 people in ADNI have genetic data for eQTL analysis
      # plink meta-analysis doesn't produce SE so we manually calculate it. These two formula produce the same results: 
          ## 1. abs(beta/ qnorm(p/tail)) (tail = 2) (https://www.biostars.org/p/431875/)
          ## 2. sqrt(((BETA)^2)/qchisq(P,1,lower.tail=F)) (https://groups.google.com/g/plink2-users/c/6g9GM3KcrlI)
      mutate(SE = abs(`BETA(R)` / qnorm(`P(R)`/2)), varbeta = SE^2, .after = "BETA(R)") %>%  
      dplyr::select(SNP, rsid, CHR, BP, A1, A2, `BETA(R)`, `P(R)`, varbeta, N),
    by = "SNP"
  ) %>% 
  mutate(beta_align = ifelse(hm_effect_allele == A1, hm_beta, -hm_beta), 
         check = ifelse(hm_effect_allele == A1, TRUE, FALSE),
         check2 = case_when(
           hm_effect_allele == A1 & hm_other_allele == A2 ~ TRUE, # this is what we want
           hm_effect_allele == A2 & hm_other_allele == A1 ~ TRUE, # this can be resolved by sorting lexicographically (below)
           TRUE ~ FALSE # check for both effect and other alleles
         )) # there are two alleles that check2 fail, remove them. 

a_low <- pmin(ndufb1_aligned$A1, ndufb1_aligned$A2)
a_high <- pmax(ndufb1_aligned$A1, ndufb1_aligned$A2)
ndufb1_aligned <- ndufb1_aligned %>% 
  mutate(SNP_rename = paste0(ndufb1_aligned$CHR, ":", ndufb1_aligned$BP, ":", a_low, ":", a_high), .after = "SNP")#rename SNP field to be consistent with LD bim files
rm(a_low, a_high)
write.table(ndufb1_aligned$SNP_rename, paste0(outdir, "/1_5/ndufb1_aligned_snps.txt"), sep = "\t", col.names = F, quote = F, row.names = F)
```


```{bash}
wgs_dir="./Rebuttal/outdir/1_10/plink_outs" # the WGS files with updated variant IDs are done in rbt_1_10_PRSv3.Rmd
outdir="./Rebuttal/outdir/1_5"

# calculate allele frequency for NDUFB1 SNPs
plink \
  --bfile "${wgs_dir}/ADNI_WGS_CSHRP_GBHNF_UpdatedIIDs_HG38_UpdatedVarID" \
  --freq \
  --allow-extra-chr \
  --extract "${outdir}/ndufb1_aligned_snps.txt" \
  --out "${outdir}/ndufb1_snps_freq_ADNI"
  
plink \
  --bfile "${wgs_dir}/MCSA_WGS_CHR1_22_UpdatedVarID" \
  --freq \
  --allow-extra-chr \
  --extract "${outdir}/ndufb1_aligned_snps.txt" \
  --out "${outdir}/ndufb1_snps_freq_MCSA"
```

```{r}
# prepare colocalization input from Bellenguez
## need to remove beta with NA or will receive error when doing check_dataset()
## need to remove beta = 0 or will receive this error: Error in if (min(p) > warn.minp) warning("minimum p value is: ", format.pval(min(p)),  : 
  # missing value where TRUE/FALSE needed
bell_ndufb1_forcoloc <- ndufb1_aligned %>% 
  filter(check2 != FALSE) %>% 
  drop_na(hm_beta) %>% 
  filter(`BETA(R)` != 0) %>%  
  dplyr::select(hm_beta, varbeta.x, SNP_rename, hm_pos, N.x) %>% 
  rename(beta = hm_beta, varbeta = varbeta.x, snp = SNP_rename, position = hm_pos, N = N.x) %>% 
  as.list()
bell_ndufb1_forcoloc$type = "cc"
check_dataset(bell_ndufb1_forcoloc)

# We need EAF because we don't know the sdY
## sdY should be the standard deviation of residuals but because I'm using eQTL results from meta-analysis, I can't just take either one of sd of residuals from MCSA or ADNI, therefore I estimated the MAF from MCSA and do a sample-size weighted average MAF.
eaf_ndufb1_mcsa <- read.table(paste0(outdir, "/1_5/ndufb1_snps_freq_MCSA.frq"), header = T, stringsAsFactors = F, quote = "")
eaf_ndufb1_adni <- read.table(paste0(outdir, "/1_5/ndufb1_snps_freq_ADNI.frq"), header = T, stringsAsFactors = F, quote = "")
maf_ndufb1_meta <- eaf_ndufb1_mcsa %>% left_join(eaf_ndufb1_adni, by=c("CHR", "SNP"), suffix = c("_MCSA", "_ADNI"))
maf_ndufb1_meta <- maf_ndufb1_meta %>% 
  mutate(check = case_when(
           A1_MCSA == A1_ADNI & A2_MCSA == A2_ADNI ~ TRUE,
           A1_MCSA == A2_ADNI & A2_MCSA == A1_ADNI ~ TRUE,
           TRUE ~ FALSE # check this for both effect and other alleles
         )) %>% #all TRUE
  mutate(MAF_ADNI_aligned = ifelse(A1_MCSA == A1_ADNI, MAF_ADNI, 1-MAF_ADNI)) %>% 
  # do a sample-size-weighted average MAF
  mutate(MAF_meta = (MAF_MCSA * NCHROBS_MCSA/2 + MAF_ADNI * NCHROBS_ADNI/2) / (NCHROBS_MCSA/2 + NCHROBS_ADNI/2),
         MAF_meta_final = pmin(MAF_meta, 1 - MAF_meta))
all.equal(maf_ndufb1_meta$MAF_meta, maf_ndufb1_meta$MAF_meta_final) #TRUE

# prepare colocalizatin input from eQTL
eqtl_ndufb1_forcoloc <- ndufb1_aligned %>% 
  left_join(maf_ndufb1_meta %>% dplyr::select(SNP, MAF_meta_final), by=c("SNP_rename" = "SNP")) %>% 
  filter(check2 != FALSE) %>% 
  drop_na(hm_beta) %>% 
  filter(`BETA(R)` != 0) %>% 
  dplyr::select(`BETA(R)`, varbeta.y, SNP_rename, BP, MAF_meta_final) %>% 
  rename(beta = `BETA(R)`, varbeta = varbeta.y, snp = SNP_rename, position = BP, MAF = MAF_meta_final) %>% 
  as.list()
eqtl_ndufb1_forcoloc$N = 193
eqtl_ndufb1_forcoloc$type = "quant"
check_dataset(eqtl_ndufb1_forcoloc)


# run coloc
ndufb1_coloc <- coloc.abf(dataset1=bell_ndufb1_forcoloc, dataset2=eqtl_ndufb1_forcoloc) # "PP abf for shared variant: 3.87%"
#PP.H0.abf PP.H1.abf PP.H2.abf PP.H3.abf PP.H4.abf 
 #2.61e-09  6.54e-01  1.23e-09  3.07e-01  3.87e-02 
```

## NDUFB1, rs12590654
```{r}
chr <- 14
pos <- 92472511 # position for rs12590654

# acquire all SNPs in our eQTL study and in Bellenguez that are 500kb from rs12590654
eqtl_ndufb1_pos2 <- ndufs_eqtl %>% filter(CHR == chr & BP >= pos - 500000 & BP <= pos + 500000) %>% filter(gene_name == "NDUFB1")
bell_ndufb1_pos2 <- bellenguez_h %>% filter(hm_chrom == chr & hm_pos >= pos - 500000 & hm_pos <= pos + 500000)
bell_ndufb1_pos2$SNP <- paste0(bell_ndufb1_pos2$hm_chrom, ":", bell_ndufb1_pos2$hm_pos)

#Get only SNPs present in both datasets
overlap <- intersect(eqtl_ndufb1_pos2$SNP, bell_ndufb1_pos2$SNP) #3188 SNPs
eqtl_ndufb1_pos2 <- eqtl_ndufb1_pos2 %>% filter(SNP %in% overlap)
bell_ndufb1_pos2 <- bell_ndufb1_pos2 %>% filter(SNP %in% overlap)

#Make sure the alleles are alligned in direction of effect
#Aligning the GWAS betas to the tested alleles of the eQTL
ndufb1_pos2_aligned <- bell_ndufb1_pos2 %>% 
  mutate(varbeta = standard_error^2, N = n_cas + n_con) %>% 
  dplyr::select(SNP, hm_rsid, hm_chrom, hm_pos, hm_effect_allele, hm_other_allele, hm_beta, varbeta, N) %>% 
  left_join(
    eqtl_ndufb1_pos2 %>% 
      mutate(N = 193) %>% #102 people in MCSA and 91 people in ADNI have genetic data for eQTL analysis
      # plink meta-analysis doesn't produce SE so we manually calculate it. These two formula produce the same results: 
          ## 1. abs(beta/ qnorm(p/tail)) (tail = 2) (https://www.biostars.org/p/431875/)
          ## 2. sqrt(((BETA)^2)/qchisq(P,1,lower.tail=F)) (https://groups.google.com/g/plink2-users/c/6g9GM3KcrlI)
      mutate(SE = abs(`BETA(R)` / qnorm(`P(R)`/2)), varbeta = SE^2, .after = "BETA(R)") %>%  
      dplyr::select(SNP, rsid, CHR, BP, A1, A2, `BETA(R)`, `P(R)`, varbeta, N),
    by = "SNP"
  ) %>% 
  mutate(beta_align = ifelse(hm_effect_allele == A1, hm_beta, -hm_beta), 
         check = ifelse(hm_effect_allele == A1, TRUE, FALSE),
         check2 = case_when(
           hm_effect_allele == A1 & hm_other_allele == A2 ~ TRUE, # this is what we want
           hm_effect_allele == A2 & hm_other_allele == A1 ~ TRUE, # this can be resolved by sorting lexicographically (below)
           TRUE ~ FALSE # check for both effect and other alleles
         )) # there are two alleles that check2 fail, remove them. 

a_low <- pmin(ndufb1_pos2_aligned$A1, ndufb1_pos2_aligned$A2)
a_high <- pmax(ndufb1_pos2_aligned$A1, ndufb1_pos2_aligned$A2)
ndufb1_pos2_aligned <- ndufb1_pos2_aligned %>% 
  mutate(SNP_rename = paste0(ndufb1_pos2_aligned$CHR, ":", ndufb1_pos2_aligned$BP, ":", a_low, ":", a_high), .after = "SNP")#rename SNP field to be consistent with LD bim files
rm(a_low, a_high)
write.table(ndufb1_pos2_aligned$SNP_rename, paste0(outdir, "/1_5/ndufb1_pos2_aligned_snps.txt"), sep = "\t", col.names = F, quote = F, row.names = F)
```


```{bash}
wgs_dir="./Rebuttal/outdir/1_10/plink_outs"
outdir="./Rebuttal/outdir/1_5"


# calculate allele frequency for NDUFB1 SNPs
plink \
  --bfile "${wgs_dir}/ADNI_WGS_CSHRP_GBHNF_UpdatedIIDs_HG38_UpdatedVarID" \
  --freq \
  --allow-extra-chr \
  --extract "${outdir}/ndufb1_pos2_aligned_snps.txt" \
  --out "${outdir}/ndufb1_pos2_snps_freq_ADNI"
  
plink \
  --bfile "${wgs_dir}/MCSA_WGS_CHR1_22_UpdatedVarID" \
  --freq \
  --allow-extra-chr \
  --extract "${outdir}/ndufb1_pos2_aligned_snps.txt" \
  --out "${outdir}/ndufb1_pos2_snps_freq_MCSA"
```

```{r}
# prepare colocalization input from Bellenguez
## need to remove beta with NA or will receive error when doing check_dataset()
## need to remove beta = 0 or will receive this error: Error in if (min(p) > warn.minp) warning("minimum p value is: ", format.pval(min(p)),  : 
  # missing value where TRUE/FALSE needed
bell_ndufb1_pos2_forcoloc <- ndufb1_pos2_aligned %>% 
  filter(check2 != FALSE) %>% 
  drop_na(hm_beta) %>% 
  filter(`BETA(R)` != 0) %>% 
  dplyr::select(hm_beta, varbeta.x, SNP_rename, hm_pos, N.x) %>% 
  rename(beta = hm_beta, varbeta = varbeta.x, snp = SNP_rename, position = hm_pos, N = N.x) %>% 
  as.list()
bell_ndufb1_pos2_forcoloc$type = "cc"
check_dataset(bell_ndufb1_pos2_forcoloc)

# We need EAF because we don't know the sdY
## sdY should be the standard deviation of residuals but because I'm using eQTL results from meta-analysis, I can't just take either one of sd of residuals from MCSA or ADNI, therefore I estimated the MAF from MCSA and do a sample-size weighted average MAF.
eaf_ndufb1_pos2_mcsa <- read.table(paste0(outdir, "/1_5/ndufb1_pos2_snps_freq_MCSA.frq"), header = T, stringsAsFactors = F, quote = "")
eaf_ndufb1_pos2_adni <- read.table(paste0(outdir, "/1_5/ndufb1_pos2_snps_freq_ADNI.frq"), header = T, stringsAsFactors = F, quote = "")
maf_ndufb1_pos2_meta <- eaf_ndufb1_pos2_mcsa %>% left_join(eaf_ndufb1_pos2_adni, by=c("CHR", "SNP"), suffix = c("_MCSA", "_ADNI"))
maf_ndufb1_pos2_meta <- maf_ndufb1_pos2_meta %>% 
  mutate(check = case_when(
           A1_MCSA == A1_ADNI & A2_MCSA == A2_ADNI ~ TRUE,
           A1_MCSA == A2_ADNI & A2_MCSA == A1_ADNI ~ TRUE,
           TRUE ~ FALSE # check this for both effect and other alleles
         )) %>% #all TRUE
  mutate(MAF_ADNI_aligned = ifelse(A1_MCSA == A1_ADNI, MAF_ADNI, 1-MAF_ADNI)) %>% 
  mutate(MAF_meta = (MAF_MCSA * NCHROBS_MCSA/2 + MAF_ADNI * NCHROBS_ADNI/2) / (NCHROBS_MCSA/2 + NCHROBS_ADNI/2),
         MAF_meta_final = pmin(MAF_meta, 1 - MAF_meta))
all.equal(maf_ndufb1_pos2_meta$MAF_meta, maf_ndufb1_pos2_meta$MAF_meta_final) #TRUE

# prepare coloc input from eQTL
eqtl_ndufb1_pos2_forcoloc <- ndufb1_pos2_aligned %>% 
  left_join(maf_ndufb1_pos2_meta %>% dplyr::select(SNP, MAF_meta_final), by=c("SNP_rename" = "SNP")) %>% 
  filter(check2 != FALSE) %>% 
  drop_na(hm_beta) %>% 
  filter(`BETA(R)` != 0) %>% 
  dplyr::select(`BETA(R)`, varbeta.y, SNP_rename, BP, MAF_meta_final) %>% 
  rename(beta = `BETA(R)`, varbeta = varbeta.y, snp = SNP_rename, position = BP, MAF = MAF_meta_final) %>% 
  as.list()
eqtl_ndufb1_pos2_forcoloc$N = 193
eqtl_ndufb1_pos2_forcoloc$type = "quant"
check_dataset(eqtl_ndufb1_pos2_forcoloc)

# run coloc
ndufb1_pos2_coloc <- coloc.abf(dataset1=bell_ndufb1_pos2_forcoloc, dataset2=eqtl_ndufb1_pos2_forcoloc) # "PP abf for shared variant: 3.88%"
#PP.H0.abf PP.H1.abf PP.H2.abf PP.H3.abf PP.H4.abf 
 #2.62e-09  6.55e-01  1.23e-09  3.07e-01  3.88e-02 
```

## NDUFA3, rs587709
```{r}
chr <- 19
pos <- 54267597 # position for rs587709

# acquire all SNPs in our eQTL study and in Bellenguez that are 500kb from rs740192
eqtl_ndufa3 <- ndufs_eqtl %>% filter(CHR == chr & BP >= pos - 500000 & BP <= pos + 500000) %>% filter(gene_name == "NDUFA3")
bell_ndufa3 <- bellenguez_h %>% filter(hm_chrom == chr & hm_pos >= pos - 500000 & hm_pos <= pos + 500000)
bell_ndufa3$SNP <- paste0(bell_ndufa3$hm_chrom, ":", bell_ndufa3$hm_pos)

#Get only SNPs present in both datasets
overlap <- intersect(eqtl_ndufa3$SNP, bell_ndufa3$SNP) #2616 SNPs
eqtl_ndufa3 <- eqtl_ndufa3 %>% filter(SNP %in% overlap)
bell_ndufa3 <- bell_ndufa3 %>% filter(SNP %in% overlap)

#Make sure the alleles are alligned in direction of effect
#Aligning the GWAS betas to the tested alleles of the eQTL
ndufa3_aligned <- bell_ndufa3 %>% 
  mutate(varbeta = standard_error^2, N = n_cas + n_con) %>% 
  dplyr::select(SNP, hm_rsid, hm_chrom, hm_pos, hm_effect_allele, hm_other_allele, hm_beta, varbeta, N) %>% 
  left_join(
    eqtl_ndufa3 %>% 
      mutate(N = 193) %>% #102 people in MCSA and 91 people in ADNI have genetic data for eQTL analysis
      # plink meta-analysis doesn't produce SE so we manually calculate it. These two formula produce the same results: 
          ## 1. abs(beta/ qnorm(p/tail)) (tail = 2) (https://www.biostars.org/p/431875/)
          ## 2. sqrt(((BETA)^2)/qchisq(P,1,lower.tail=F)) (https://groups.google.com/g/plink2-users/c/6g9GM3KcrlI)
      mutate(SE = abs(`BETA(R)` / qnorm(`P(R)`/2)), varbeta = SE^2, .after = "BETA(R)") %>%  
      dplyr::select(SNP, rsid, CHR, BP, A1, A2, `BETA(R)`, `P(R)`, varbeta, N),
    by = "SNP"
  ) %>% 
  mutate(beta_align = ifelse(hm_effect_allele == A1, hm_beta, -hm_beta), 
         check = ifelse(hm_effect_allele == A1, TRUE, FALSE),
         check2 = case_when(
           hm_effect_allele == A1 & hm_other_allele == A2 ~ TRUE, # this is what we want
           hm_effect_allele == A2 & hm_other_allele == A1 ~ TRUE, # this can be resolved by sorting lexicographically (below)
           TRUE ~ FALSE # check for both effect and other alleles
         )) # there are two alleles that check2 fail, remove them. 

a_low <- pmin(ndufa3_aligned$A1, ndufa3_aligned$A2)
a_high <- pmax(ndufa3_aligned$A1, ndufa3_aligned$A2)
ndufa3_aligned <- ndufa3_aligned %>% 
  mutate(SNP_rename = paste0(ndufa3_aligned$CHR, ":", ndufa3_aligned$BP, ":", a_low, ":", a_high), .after = "SNP")#rename SNP field to be consistent with LD bim files
rm(a_low, a_high)
write.table(ndufa3_aligned$SNP_rename, paste0(outdir, "/1_5/ndufa3_aligned_snps.txt"), sep = "\t", col.names = F, quote = F, row.names = F)
```


```{bash}
wgs_dir="./Rebuttal/outdir/1_10/plink_outs"
outdir="./Rebuttal/outdir/1_5"


# calculate allele frequency for NDUFA3 SNPs
plink \
  --bfile "${wgs_dir}/ADNI_WGS_CSHRP_GBHNF_UpdatedIIDs_HG38_UpdatedVarID" \
  --freq \
  --allow-extra-chr \
  --extract "${outdir}/ndufa3_aligned_snps.txt" \
  --out "${outdir}/ndufa3_snps_freq_ADNI"
  
plink \
  --bfile "${wgs_dir}/MCSA_WGS_CHR1_22_UpdatedVarID" \
  --freq \
  --allow-extra-chr \
  --extract "${outdir}/ndufa3_aligned_snps.txt" \
  --out "${outdir}/ndufa3_snps_freq_MCSA"
```

```{r}
# prepare colocalization input from Bellenguez
## need to remove beta with NA or will receive error when doing check_dataset()
## need to remove beta = 0 or will receive this error: Error in if (min(p) > warn.minp) warning("minimum p value is: ", format.pval(min(p)),  : 
  # missing value where TRUE/FALSE needed
bell_ndufa3_forcoloc <- ndufa3_aligned %>% 
  filter(check2 != FALSE) %>% 
  drop_na(hm_beta) %>% 
  filter(`BETA(R)` != 0) %>% 
  dplyr::select(hm_beta, varbeta.x, SNP_rename, hm_pos, N.x) %>% 
  rename(beta = hm_beta, varbeta = varbeta.x, snp = SNP_rename, position = hm_pos, N = N.x) %>% 
  as.list()
bell_ndufa3_forcoloc$type = "cc"
check_dataset(bell_ndufa3_forcoloc)

# We need EAF because we don't know the sdY
## sdY should be the standard deviation of residuals but because I'm using eQTL results from meta-analysis, I can't just take either one of sd of residuals from MCSA or ADNI, therefore I estimated the MAF from MCSA and do a sample-size weighted average MAF.
eaf_ndufa3_mcsa <- read.table(paste0(outdir, "/1_5/ndufa3_snps_freq_MCSA.frq"), header = T, stringsAsFactors = F, quote = "")
eaf_ndufa3_adni <- read.table(paste0(outdir, "/1_5/ndufa3_snps_freq_ADNI.frq"), header = T, stringsAsFactors = F, quote = "")
maf_ndufa3_meta <- eaf_ndufa3_mcsa %>% left_join(eaf_ndufa3_adni, by=c("CHR", "SNP"), suffix = c("_MCSA", "_ADNI"))
maf_ndufa3_meta <- maf_ndufa3_meta %>% 
  mutate(check = case_when(
           A1_MCSA == A1_ADNI & A2_MCSA == A2_ADNI ~ TRUE,
           A1_MCSA == A2_ADNI & A2_MCSA == A1_ADNI ~ TRUE,
           TRUE ~ FALSE # check this for both effect and other alleles
         )) %>% #all TRUE
  mutate(MAF_ADNI_aligned = ifelse(A1_MCSA == A1_ADNI, MAF_ADNI, 1-MAF_ADNI)) %>% 
  # do a sample-size-weighted average MAF
  mutate(MAF_meta = (MAF_MCSA * NCHROBS_MCSA/2 + MAF_ADNI * NCHROBS_ADNI/2) / (NCHROBS_MCSA/2 + NCHROBS_ADNI/2),
         MAF_meta_final = pmin(MAF_meta, 1 - MAF_meta))
all.equal(maf_ndufa3_meta$MAF_meta, maf_ndufa3_meta$MAF_meta_final) #TRUE


# prepare coloc input from eQTL
eqtl_ndufa3_forcoloc <- ndufa3_aligned %>% 
  left_join(maf_ndufa3_meta %>% dplyr::select(SNP, MAF_meta_final), by=c("SNP_rename" = "SNP")) %>% 
  filter(check2 != FALSE) %>% 
  drop_na(hm_beta) %>% 
  filter(`BETA(R)` != 0) %>% 
  dplyr::select(`BETA(R)`, varbeta.y, SNP_rename, BP, MAF_meta_final) %>% 
  rename(beta = `BETA(R)`, varbeta = varbeta.y, snp = SNP_rename, position = BP, MAF = MAF_meta_final) %>% 
  as.list()
eqtl_ndufa3_forcoloc$N = 193
eqtl_ndufa3_forcoloc$type = "quant"
check_dataset(eqtl_ndufa3_forcoloc)


# run coloc
ndufa3_coloc <- coloc.abf(dataset1=bell_ndufa3_forcoloc, dataset2=eqtl_ndufa3_forcoloc) # "PP abf for shared variant: 5.02%"
#PP.H0.abf PP.H1.abf PP.H2.abf PP.H3.abf PP.H4.abf 
 #0.000974  0.690000  0.000365  0.258000  0.050200 
```

# MR

## Check MR requirements

1)  IVs need to be independent --\> focus on 11 eQTL pairs after clumping
2)  F-stats \>=10 -- not a strict requirement, but IV with F-stats \< 10 will lead to [weak instrument bias](https://mr-dictionary.mrcieu.ac.uk/term/weak-instrument/).

### Get average F-stats 
Average F-stats: (sum of beta^2/SE^2)/number of variants. Follow this paper: https://journals.lww.com/hep/fulltext/2025/06000/mendelian_randomization_in_hepatology__a_review_of.25.aspx.

The F-stats are in Supplementary Table 16a
```{r}
# plink meta-analysis doesn't produce SE so we manually calculate it.
# these two formula produce the same results: 
 ## 1. abs(beta/ qnorm(p/tail)) (tail = 2) (https://www.biostars.org/p/431875/)
 ## 2. sqrt(((BETA)^2)/qchisq(P,1,lower.tail=F)) (https://groups.google.com/g/plink2-users/c/6g9GM3KcrlI)

clumped_eqtl <- clumped_eqtl %>% 
  mutate(SE = abs(`BETA(R)` / qnorm(`P(R)`/2)), Var = SE^2, Fstat = `BETA(R)`^2/Var, .after = "BETA(R)") %>% 
  # change the rsid of two SNPs that are originally NA. The rsid were obtained from dbSNP
  mutate(rsid = case_when(SNP_rename == "11:78067306:A:G" ~ "rs634768", TRUE ~ rsid)) # when examining SNPs individually, all have Fstat > 10

```

## Run regression between IV and outcome
```{r run_reg function}
# a function to run regression analysis with or without covariates
reg_fn <- function(df_ls, 
                   responses,
                   model_type,
                   include_covariates,
                   covariates){
  run_reg <- function(response){
    if (include_covariates == TRUE){
        map(names(df_ls), function(SNP){
          df <- df_ls[[SNP]] # df = dataframe for the called SNP
          snp <- SNP # snp = called SNP
          df <- df %>% drop_na(any_of(c(snp, covariates))) # remove any NAs in snp or covariates
          
          form <- reformulate(c(snp, covariates), response = response) # formula of the regression
          
          if (model_type == "linear"){
            m <- lm(form, data = df)
          }
          else if (model_type == "binomial"){
            m <- glm(form, data = df, family = "binomial")
          }
          
          tidy(m, conf.int = TRUE) %>% 
            filter(grepl(snp, term)) %>% 
            select(rsid = term, estimate, std.error, statistic, p.value, conf.low, conf.high) # final dataframe with relevant statistics
        }
      ) %>% 
        set_names(names(df_ls))
    }
  
    else {
      map(names(df_ls), function(SNP){
          df <- df_ls[[SNP]]
          snp <- SNP
          df <- df %>% drop_na(any_of(c(snp)))
          
          form <- reformulate(c(snp), response = response)
          
          if (model_type == "linear"){
            m <- lm(form, data = df)
          }
          else if (model_type == "binomial"){
            m <- glm(form, data = df, family = "binomial")
          }
          
          tidy(m, conf.int = TRUE) %>% 
            filter(grepl(snp, term)) %>% 
            select(rsid = term, estimate, std.error, statistic, p.value, conf.low, conf.high)
        }
      ) %>% 
        set_names(names(df_ls))
    }
  }
  
  # Run for all combinations
  results <- expand.grid(
      Outcomes = responses,
      stringsAsFactors = F
  ) %>%
  mutate(result = pmap(list(Outcomes),
                      ~ run_reg(..1))) %>%
  unnest(result)
  
  return(results)
}
```
Prepare dataframe and run regression
```{r IV --> outcome}
comTrait <- readRDS(paste0(indir, "/comTrait.rds")) # a file having cross-sectional endophenotypes for MCSA (from row 1-105) and ADNI (from row 106-196); row names are donorID

# read in population substructure covariates 
mcsa_pcs <- read_delim(paste0(indir, "/mcsa_eqtl_covar.txt"), delim = "\t") %>% # a file with WGS IID and population substructure PCs and also ID used in comTrait and other analyses 
  dplyr::rename(PC1 = EV1, PC2 = EV2, PC3 = EV3) %>% 
  dplyr::select(IID, ptnum, PC1:PC3) %>% 
  mutate(SubjectID = as.character(ptnum)) # ptnum = SubjectID
adni_pcs <- read.table(paste0(indir, "/WGS/ADNI_WGS_CSHR_GBHNFL.pca.evec"), skip = 1, header = F, stringsAsFactors = F)
colnames(adni_pcs)<- c("IID","PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7", "PC8", "PC9", "PC10", "Dx")

## prepare MCSA phenotype and covariates files 
mcsa_df <- comTrait[1:105,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Memory, LMDR, HippVol) %>% 
  left_join(mcsa_pcs) %>% 
  left_join(mcsa_traw %>% dplyr::select(IID, gene_name, SNP, rsid, Dose), by = c("IID")) %>% 
  drop_na(SNP) %>% 
  dplyr::select(IID, everything())


## prepare ADNI phenotype and covariates files
adni_df <- comTrait[106:196,] %>% 
  rownames_to_column("IID") %>% 
  dplyr::select(IID, Memory, LMDR, HippVol) %>% 
  left_join(adni_pcs %>% dplyr::select(IID:PC3) %>% mutate(IID = toupper(IID))) %>% 
  left_join(adni_traw %>% dplyr::select(IID, gene_name, SNP, rsid, Dose), by = c("IID")) %>% 
  drop_na(SNP) %>% 
  dplyr::select(IID, everything())


mcsa_df_ls <- mcsa_df %>% 
  named_group_split(rsid) %>% # this creates a list of 11 dataframes, one for each SNP
  map(~ dplyr::select(.x, -SNP)) %>% 
  map(~ pivot_wider(.x, names_from = "rsid", values_from = "Dose"))
mcsa_IV_outcome_PCs <- reg_fn(df_ls = mcsa_df_ls, responses = c("Memory", "LMDR", "HippVol"), model_type = "linear", include_covariates = TRUE, covariates = c("PC1", "PC2", "PC3")) %>% # we adjust for first 3 principal componenets of population substructure
  unnest(result) %>% 
  group_by(Outcomes) %>% 
  mutate(q.value = p.adjust(p.value, "fdr"), .after = "p.value")


adni_df_ls <- adni_df %>% 
  named_group_split(rsid) %>% # this creates a list of 11 dataframes, one for each SNP
  map(~ dplyr::select(.x, -SNP)) %>% 
  map(~ pivot_wider(.x, names_from = "rsid", values_from = "Dose"))
adni_IV_outcome_PCs <- reg_fn(df_ls = adni_df_ls, responses = c("Memory", "LMDR", "HippVol"), model_type = "linear", include_covariates = TRUE, covariates = c("PC1", "PC2", "PC3")) %>% # we adjust for first 3 principal componenets of population substructure
  unnest(result) %>% 
  group_by(Outcomes) %>% 
  mutate(q.value = p.adjust(p.value, "fdr"), .after = "p.value")

# Save those objects to an .RData file
save(list = ls(pattern = "IV_Outcome"), file = paste0(outdir, "/1_5/IV_Outcome.RData"))
```

```{r meta-analyze IV-->outcome}
# prepare input dataframe for meta-analysis
meta_df_PCs = inner_join(mcsa_IV_outcome_PCs, adni_IV_outcome_PCs, by = c("Outcomes","rsid"), suffix = c("_MCSA", "_ADNI")) %>%
  mutate(Out_rsid = paste0(Outcomes,"_",rsid)) %>% 
  column_to_rownames("Out_rsid")

meta_ana_fn <- function(meta_df){
  # this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
  eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
  eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
  err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
  err_idx2 = which(colnames(meta_df) == "std.error_ADNI")
  
  # run meta-analysis using the function meta_gen_fn defined above
  meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)
  
  # Get all fixed-effect results
  fix_res = meta_raw %>%
    pbsapply(function(x) x$fixed) %>%
    t() %>%
    as.data.frame() %>%
    rename_with(~paste0(.x, "_fixed")) %>%
    rownames_to_column(var = "rn") %>%
    mutate_all(.funs = ~unlist(.x)) %>% 
    mutate_all(.funs = ~unname(.x))
  
  # Get all random-effect results
  rand_res = meta_raw %>%
    pbsapply(function(x) x$random) %>%
    t() %>%
    data.frame() %>%
    rename_with(~paste0(.x, "_random")) %>%
    rownames_to_column(var = "rn") %>%
    mutate_all(.funs = ~unlist(.x)) %>%
    mutate_all(.funs = ~unname(.x))
  
  # Get heterogenity meassurments Q, I2, tau2
  het_res = meta_raw %>%
    map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
    bind_rows(.id = "rn")
  
  # Join the results to the original regression coeffcients
  meta_df = meta_df %>%
    rownames_to_column(var = "rn") %>% 
    inner_join(fix_res) %>%
    inner_join(rand_res) %>% 
    rename_with(~sub("^TE", "beta", .x), everything()) %>%
    rename_with(~sub("^seTE", "se", .x), everything()) %>%
    inner_join(het_res, by = "rn")
  
  # final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
  meta_final = meta_df %>%
    group_by(Outcomes) %>%
    mutate(q_random = p.adjust(p_random, method = "fdr"),
           q_fixed = p.adjust(p_fixed, method = "fdr"),
           model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
           beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
           se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
           p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
           q_meta = ifelse(model_meta =="Fixed", q_fixed, q_random),
           CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random),
           CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random)) %>%
    dplyr::select(-c(ends_with("_random"), ends_with("_fixed"))) %>% 
    dplyr::select(-rn)
}  

meta_IV_outs_PCs <- meta_ana_fn(meta_df_PCs)
```

# Run MR (TwoSampleMR)
```{r}
# prepare dataframe for TwoSampleMR input
exp_dat <- clumped_eqtl %>% 
  dplyr::select(rsid, `BETA(R)`, SE, A1, A2, gene_name) %>% 
  rename(
    SNP = rsid,
    exposure = gene_name,
    beta.exposure = `BETA(R)`,
    se.exposure = SE,
    effect_allele.exposure = A1,
    other_allele.exposure = A2
  ) %>% 
  mutate(id.exposure = exposure)
out_dat_PCs <- meta_IV_outs_PCs %>% 
  dplyr::select(rsid, Outcomes, beta_meta, se_meta) %>% 
  left_join(clumped_eqtl %>% dplyr::select(rsid, A1, A2)) %>% 
  rename(
    SNP = rsid,
    outcome = Outcomes,
    beta.outcome = beta_meta,
    se.outcome = se_meta,
    effect_allele.outcome = A1,
    other_allele.outcome = A2
  ) %>% 
  mutate(id.outcome = outcome, mr_keep = TRUE)

mr_in_PCs_split <- exp_dat %>% left_join(out_dat_PCs, by = "SNP")
# run MR
mr_PCs_split <- mr(mr_in_PCs_split)
mr_PCs_split <- mr_PCs_split %>% left_join(mr_in_PCs_split %>% dplyr::select(SNP, outcome, exposure)) 
mr_PCs_split <- mr_PCs_split %>% dplyr::select(-id.exposure, -id.outcome) %>% relocate(SNP, .before = "outcome")
write.xlsx(mr_PCs_split, paste0(outdir, "/1_5/TwoSampleMR_WaldRatio.xlsx")) # the organized table is in Supplementary Table 16b
```



