---
title: "rbt_1_4_cbx: LM10"
output: html_document
date: "2026-02-21"
---

Author: Wei (Adelyn) Tsai; [tsai.wei\@mayo.edu](mailto:tsai.wei@mayo.edu){.email}.

If you are using this script, please cite our study.

This script addresses comment 4 about cell type proportion from reviewer 1
Please note that based on the 2nd rebuttal, results related to Bl_M3 have been updated. Please see rebuttal_2nd/rbt_1_4.Rmd for more details.

```{r prep}
suppressPackageStartupMessages(library(tidyverse)) #v2.0.0
suppressPackageStartupMessages(library(patchwork)) #v1.3.2
suppressPackageStartupMessages(library(broom)) #v1.0.12
suppressPackageStartupMessages(library(openxlsx)) #v4.2.8
suppressPackageStartupMessages(library(Hmisc)) #v5.1
suppressPackageStartupMessages(library(pbapply)) #v1.7
suppressPackageStartupMessages(library(ggpubr)) #v0.6.0
suppressPackageStartupMessages(library(lmerTest)) #v3.1
suppressPackageStartupMessages(library(broom.mixed)) #v0.2.9

set.seed(123)

# list of input/output directories
indir <- "./indir"
blwgcna_dir <- "./outdir/bl_wgcna"
nduf_assoc_dir <- "./outdir/nduf_assoc_de"
residir <- "./outdir/resids"

# output directory
outdir <- "./Rebuttal/outdir/1_4/CBX_LM10"

source("./Codes/fns.R") #source a list of frequently used functions
```

# Analyze results after adjusting for fractions

## Prepare fraction dataframes for analyses
```{r}
mcsa_frac <- read_delim(paste0(indir, "/CBXHiRes_upload/CIBERSORTxGEP_R01_Fractions-Adjusted.txt")) # this is the CIBERSORTx fraction results of MCSA
adni_frac <- read_delim(paste0(indir, "/CBXHiRes_upload/CIBERSORTxGEP_ADNI_s91_Fractions.txt")) # this is the CIBERSORTx fraction results of ADNI
# Rows are subjects and each column is fraction for each cell type

# renames for simplicity
mcsa_frac <- mcsa_frac %>% 
  dplyr::select(-`P-value`, -Correlation, -RMSE) %>% 
  mutate(Mixture = as.character(Mixture)) %>%
  dplyr::rename(
    SubjectID = Mixture,
    B_naive = `B cells naive`,
    B_memory = `B cells memory`,
    Plasma = `Plasma cells`,
    CD8T = `T cells CD8`,
    CD4T_naive = `T cells CD4 naive`,
    CD4T_memory_resting = `T cells CD4 memory resting`,
    CD4T_memory_activated = `T cells CD4 memory activated`,
    Tfh = `T cells follicular helper`,
    Tregs = `T cells regulatory (Tregs)`,
    γδT = `T cells gamma delta`,
    NK_resting = `NK cells resting`,
    NK_activated = `NK cells activated`,
    M0 = `Macrophages M0`,
    M1 = `Macrophages M1`,
    M2 = `Macrophages M2`,
    DC_resting = `Dendritic cells resting`,
    DC_activated = `Dendritic cells activated`,
    Mast_resting = `Mast cells resting`,
    Mast_activated = `Mast cells activated`
    )
adni_frac <- adni_frac %>% 
  dplyr::select(-`P-value`, -Correlation, -RMSE) %>% 
  mutate(Mixture = gsub("^X", "", Mixture)) %>% 
  dplyr::rename(
    SubjectID = Mixture,
    B_naive = `B cells naive`,
    B_memory = `B cells memory`,
    Plasma = `Plasma cells`,
    CD8T = `T cells CD8`,
    CD4T_naive = `T cells CD4 naive`,
    CD4T_memory_resting = `T cells CD4 memory resting`,
    CD4T_memory_activated = `T cells CD4 memory activated`,
    Tfh = `T cells follicular helper`,
    Tregs = `T cells regulatory (Tregs)`,
    γδT = `T cells gamma delta`,
    NK_resting = `NK cells resting`,
    NK_activated = `NK cells activated`,
    M0 = `Macrophages M0`,
    M1 = `Macrophages M1`,
    M2 = `Macrophages M2`,
    DC_resting = `Dendritic cells resting`,
    DC_activated = `Dendritic cells activated`,
    Mast_resting = `Mast cells resting`,
    Mast_activated = `Mast cells activated`
  )

ct_grps <- list(
  "Bcells"  = c("B_memory", "B_naive"),
  "Plasma" = c("Plasma"),
  "CD4T" = c("CD4T_memory_activated", "CD4T_memory_resting", "CD4T_naive", "Tfh", "Tregs", "γδT"),
  "CD8T" = c("CD8T"),
  "NK" = c("NK_activated", "NK_resting"),
  "DC" = c("DC_activated", "DC_resting"),
  "Monocytes" = c("Monocytes", "M0", "M1", "M2"),
  "Mast" = c("Mast_activated", "Mast_resting"),
  "Eosinophils" = "Eosinophils",
  "Neutrophils" = "Neutrophils"
) 

# we added up the fractions based on grouping of LM10
mcsa_frac_grp <- mcsa_frac %>%
  rowwise() %>%
  mutate(
    Bcells = sum(c_across(any_of(ct_grps[["Bcells"]])), na.rm = TRUE),
    Plasma = sum(c_across(any_of(ct_grps[["Plasma"]])), na.rm = TRUE),
    CD4T = sum(c_across(any_of(ct_grps[["CD4T"]])), na.rm = TRUE),
    CD8T = sum(c_across(any_of(ct_grps[["CD8T"]])), na.rm = TRUE),
    NK = sum(c_across(any_of(ct_grps[["NK"]])), na.rm = TRUE),
    DC = sum(c_across(any_of(ct_grps[["DC"]])), na.rm = TRUE),
    Monocytes = sum(c_across(any_of(ct_grps[["Monocytes"]])),  na.rm = TRUE),
    Mast = sum(c_across(any_of(ct_grps[["Mast"]])), na.rm = TRUE),
    Eosinophils = sum(c_across(any_of(ct_grps[["Eosinophils"]])), na.rm = TRUE),
    Neutrophils = sum(c_across(any_of(ct_grps[["Neutrophils"]])), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::select(SubjectID, names(ct_grps)) %>% 
  mutate(Cohort = "MCSA", .before = "SubjectID")

adni_frac_grp <- adni_frac %>%
  rowwise() %>%
  mutate(
    Bcells = sum(c_across(any_of(ct_grps[["Bcells"]])), na.rm = TRUE),
    Plasma = sum(c_across(any_of(ct_grps[["Plasma"]])), na.rm = TRUE),
    CD4T = sum(c_across(any_of(ct_grps[["CD4T"]])), na.rm = TRUE),
    CD8T = sum(c_across(any_of(ct_grps[["CD8T"]])), na.rm = TRUE),
    NK = sum(c_across(any_of(ct_grps[["NK"]])), na.rm = TRUE),
    DC = sum(c_across(any_of(ct_grps[["DC"]])), na.rm = TRUE),
    Monocytes = sum(c_across(any_of(ct_grps[["Monocytes"]])),  na.rm = TRUE),
    Mast = sum(c_across(any_of(ct_grps[["Mast"]])), na.rm = TRUE),
    Eosinophils = sum(c_across(any_of(ct_grps[["Eosinophils"]])), na.rm = TRUE),
    Neutrophils = sum(c_across(any_of(ct_grps[["Neutrophils"]])), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::select(SubjectID, names(ct_grps))  %>% 
  mutate(Cohort = "ADNI", .before = "SubjectID")
```

Plot the fractions by cohort in a heatmap
```{r heatmap by cohort}
## heatmap 
hm_df <- rbind(mcsa_frac_grp %>% pivot_longer(cols = (names(ct_grps)), names_to = "CellType", values_to = "Fraction"),
               adni_frac_grp %>% pivot_longer(cols = (names(ct_grps)), names_to = "CellType", values_to = "Fraction")) %>% 
  group_by(CellType, Cohort) %>% 
  summarise(AveFrac = mean(Fraction)) %>%
  mutate(Labels = scales::percent(AveFrac, accuracy = 0.1),
         CellType = factor(CellType, levels = names(ct_grps)),
         Cohort = factor(Cohort, levels = c("MCSA", "ADNI")))

pdf(paste0(outdir, "/CBX_10CT_AveFrac_hm.pdf"), width = 3.5, height = 5)
hm_df %>% 
  ggplot(aes(x = Cohort, y = CellType, fill = AveFrac)) +
  geom_tile(color = "black") +
  geom_text(
    aes(label = Labels),
    color = "black", size = 11/.pt
  ) +
  labs(x = "", y = "") +
  scale_fill_gradient(high = "firebrick", low = "white", name = "Average\nFraction") +
  scale_x_discrete(expand = c(0,0)) + #remove space between ticks and plot
  scale_y_discrete(expand = c(0,0), limits = rev) +
  theme_classic(base_size = 12) +
  theme(axis.line = element_blank(),
        legend.key.size = unit(0.2, "in"),
        axis.text.x = element_text(angle = 30, hjust =1 ))
dev.off() # this is in extended data fig.13a
```


## Calculate collinearity b/w proportions 
Focus on: Bcells, CD4T, CD8T, NK, Monocytes, Mast, Neutrophils
```{r}
ct_ana <- c("Bcells", "CD4T", "CD8T", "NK", "Monocytes", "Mast", "Neutrophils")


## Perform CLR transformation
mcsa_frac0 <- mcsa_frac_grp %>% column_to_rownames("SubjectID")
mcsa_frac0[mcsa_frac0 <= 1e-6] <- 1e-6 # add offset for proportions estimated to be 0 
# similar to what was done in https://www.science.org/doi/10.1126/sciadv.adn7655

adni_frac0 <- adni_frac_grp %>% column_to_rownames("SubjectID")
adni_frac0[adni_frac0 <= 1e-6] <- 1e-6

mcsa_cp_clr <- mcsa_frac0 %>%
  mutate(
    clr = compositions::clr(across(all_of(ct_ana)))
  )
mcsa_cp_clr <- as_tibble(mcsa_cp_clr$clr)
mcsa_cp_clr_wID <- mcsa_cp_clr %>% mutate(SubjectID = mcsa_frac_grp$SubjectID, .before = "Bcells")

adni_cp_clr <- adni_frac0 %>%
  mutate(
    clr = compositions::clr(across(all_of(ct_ana)))
  )
adni_cp_clr <- as_tibble(adni_cp_clr$clr)
adni_cp_clr_wID <- adni_cp_clr %>% mutate(SubjectID = adni_frac_grp$SubjectID, .before = "Bcells")
save(mcsa_cp_clr_wID, adni_cp_clr_wID, file = paste0(outdir, "/cp_clr.RData"))

rm(adni_frac0, mcsa_frac0)
```

## Cross-sectional association adjusting fractions
```{r prepare for associations}
comTrait <- readRDS(paste0(indir,"/comTrait.rds")) # a file having cross-sectional endophenotypes for MCSA (from row 1-105) and ADNI (from row 106-196); row names are donorID
MEs1 <- read.csv(paste0(blwgcna_dir, "/MEs1_SP12.csv")) #from 2_blwgcna.R
nduf_bb <- read.csv(paste0(nduf_assoc_dir, "/nduf_bb.csv")) #generated in 4_nduf_assoc_de.R

cqn_raw <- read_delim(paste0(indir, "/PaX108_R01resilience_gene_CQN_neg3_postQC_s105.txt"), delim = "\t") # MCSA cqn file for 18046 genes. Each row is a gene and each column is a donor
# filter to only focus on NDUFs 
cqn_ndufs <- cqn_raw %>% filter(GeneId %in% nduf_bb$gene_id) %>% 
  dplyr::select(-Chromosome, -Start, -End, -Length, -GeneId, -GeneBiotype) %>%
  pivot_longer(-GeneName, names_to = "SubjectID", values_to = "Expr") %>%
  pivot_wider(names_from = "GeneName", values_from = "Expr")
mcsa_covars <- read_delim(paste0(indir,"/R01_Resilience_PAXgene_Covars.txt"), delim = "\t") %>% mutate(SubjectID=as.character(ptnum)) #covariates file for MCSA, rows are donor columns are covariates
# prepare dataframe for association analysis
cp4assoc_mcsa <- comTrait[1:105,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Age, Age_last, Sex, Educ, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV) %>% 
  left_join(mcsa_covars %>% dplyr::select(SubjectID, RIN, PAXgene_flowcell, PAXgene_Batch)) %>% # select technical variables of RNAsequencing for adjustment
  bind_cols(mcsa_cp_clr) %>% 
  left_join(MEs1 %>% dplyr::select(SubjectID, ME3)) %>% 
  left_join(cqn_ndufs) 
  
  
adni_mc <- readRDS(paste0(indir,"/ADNI_s91_microarray.rds")) #microarray file for 91 donors from ADNI. This has 10116 protein-coding genes, gene id as column names, donorID as row names
adni_covars <- read_delim(paste0(residir,"/ADNI_s91_covars.txt"), delim = "\t") #covariates file for 91 donors from ADNI. Rows are donors columns are covariates
mc_ndufs <- adni_mc %>% dplyr::select(nduf_bb$gene_id)
mc_ndufs <- mc_ndufs %>% rename_at(vars(1:(ncol(.))), ~ nduf_bb$gene_name[match(names(mc_ndufs)[1:(ncol(mc_ndufs))], nduf_bb$gene_id)]) 
# prepare dataframe for association analysis
cp4assoc_adni <- comTrait[106:196,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Age, Age_last, Sex, Educ, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV) %>% 
  mutate(Mag_HippICV = factor(Mag_HippICV)) %>% 
  left_join(adni_covars %>% dplyr::select(SubjectID, RIN, AffyPlate.meanREE, SITE.meanREE)) %>% # select technical variables of RNAsequencing for adjustment
  bind_cols(adni_cp_clr) %>% 
  left_join(MEs1 %>% dplyr::select(SubjectID, ME3)) %>% 
  left_join(mc_ndufs %>% rownames_to_column("SubjectID")) 
```

```{r associations functions}
# functions for performing associations for:
# 1) no adjustment (the original model);
# 2) adjust for all analyzed cell proportions;
# 3) adjust for one analyzed cell proportion at a time
assoc_fn <- function(df, 
                     xvars,
                     yvars,
                     adjust_types,
                     base_covs,
                     base_covs_for_HV,
                     cp_covs){
  run_assoc <- function(xvar, yvar, adjustment){
    message("Test association between ", xvar, " and ", yvar, " adjusting for ", adjustment)
    # run original association without cell type proportion adjustment
    if (adjustment == "None"){
      if (xvar %in% c("Memory", "LMDR")){
        m <- lm(paste0(yvar, " ~ ", xvar, " + ", paste0(base_covs, collapse = " + ")), data = df)
      }
      else if (xvar == "HippVol"){
        m <- lm(paste0(yvar, " ~ ", xvar, " + ", paste0(base_covs_for_HV, collapse = " + ")), data = df)
      }
    }
    # run association adjusting for all cell type proportions
    else if (adjustment == "All"){
      if (xvar %in% c("Memory", "LMDR")){
        m <- lm(paste0(yvar, " ~ ", xvar, " + ", paste0(c(base_covs, cp_covs), collapse = " + ")), data = df)
      }
      else if (xvar == "HippVol"){
        m <- lm(paste0(yvar, " ~ ", xvar, " + ", paste0(c(base_covs_for_HV, cp_covs), collapse = " + ")), data = df)
      }
    }
    # run association adjusting for one cell type proportion at a time
    else if (adjustment %in% c(cp_covs)){
      if (xvar %in% c("Memory", "LMDR")){
        m <- lm(paste0(yvar, " ~ ", xvar, " + ", paste0(c(base_covs, adjustment), collapse = " + ")), data = df)
      }
      else if (xvar == "HippVol"){
        m <- lm(paste0(yvar, " ~ ", xvar, " + ", paste0(c(base_covs_for_HV, adjustment), collapse = " + ")), data = df)
      }
    }
    
    res_df <- tidy(m, conf.int = T)
    
    tbl <- tibble(
      Model = list(m),
      SummDf = list(res_df)
    )
    return(tbl)
  }
  
  # Run for all combinations
  results <- expand.grid(
    XVar = xvars,
    YVar = yvars,
    ADJUST = adjust_types,
    stringsAsFactors = F
  ) %>%
  mutate(result = pmap(list(XVar, YVar, ADJUST),
                       ~ run_assoc(..1, ..2, ..3))) %>%
  unnest(result)
  
  return(results)
}
```

```{r run assoc}
# run associations for MCSA 
mcsa_assoc_me3 <- assoc_fn(df = cp4assoc_mcsa,
                       xvars = c("Memory", "LMDR", "HippVol"),
                       yvars = c("ME3"),
                       adjust_types = c("None", "All", ct_ana),
                       base_covs = c("Age", "Sex", "Educ"),
                       base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV"),
                       cp_covs = ct_ana
                       )
mcsa_assoc <- assoc_fn(df = cp4assoc_mcsa,
                       xvars = c("Memory", "LMDR", "HippVol"),
                       yvars = c(nduf_bb$gene_name),
                       adjust_types = c("None", "All", ct_ana),
                       base_covs = c("Age", "Sex", "Educ", "RIN", "PAXgene_flowcell", "PAXgene_Batch"),
                       base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "RIN", "PAXgene_flowcell", "PAXgene_Batch"),
                       cp_covs = ct_ana
                       )
mcsa_assoc_df <- rbind(mcsa_assoc_me3, mcsa_assoc) %>% 
  dplyr::select(XVar, YVar, ADJUST, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, XVar)) %>% 
  group_by(XVar, ADJUST) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for number of genes x adjustment tested (21*9)
  ungroup()

# run associations for ADNI
adni_assoc_me3 <- assoc_fn(df = cp4assoc_adni,
                       xvars = c("Memory", "LMDR", "HippVol"),
                       yvars = c("ME3"),
                       adjust_types = c("None", "All", ct_ana),
                       base_covs = c("Age", "Sex", "Educ"),
                       base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "Mag_HippICV"),
                       cp_covs = ct_ana
                       )
adni_assoc <- assoc_fn(df = cp4assoc_adni,
                       xvars = c("Memory", "LMDR", "HippVol"),
                       yvars = c(nduf_bb$gene_name),
                       adjust_types = c("None", "All", ct_ana),
                       base_covs = c("Age", "Sex", "Educ", "RIN", "AffyPlate.meanREE", "SITE.meanREE"),
                       base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "Mag_HippICV", "RIN", "AffyPlate.meanREE", 'SITE.meanREE'),
                       cp_covs = ct_ana
                       )
adni_assoc_df <- rbind(adni_assoc_me3, adni_assoc) %>% 
  dplyr::select(XVar, YVar, ADJUST, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, XVar)) %>% 
  group_by(XVar, ADJUST) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for numer of genes x adjustment tested (21*9)
  ungroup()

save(mcsa_assoc_me3, mcsa_assoc, adni_assoc_me3, adni_assoc, file = paste0(outdir, "/CBX_CPAdj_Assoc.RData"))
```

```{r meta-analysis}
# perform meta-analysis
meta_df = inner_join(mcsa_assoc_df, adni_assoc_df, by = c("XVar","YVar", "ADJUST", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(rn = paste0(XVar, "_", YVar, "_", ADJUST)) %>% 
  column_to_rownames("rn")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "rn")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "rn") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "rn")

# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_CP_final = meta_df %>%
  group_by(XVar, ADJUST) %>%
  mutate(
    q_local_random = p.adjust(p_random, method = "fdr"),
    q_local_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(
    q_global_random = p.adjust(p_random, method = "fdr"),
    q_global_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for number of genes x adjustment tested (21*9)
  ungroup() %>% 
  mutate(model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_local_meta = ifelse(model_meta =="Fixed", q_local_fixed, q_local_random),
         q_global_meta = ifelse(model_meta =="Fixed", q_global_fixed, q_global_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random)
         ) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"))) %>% 
  dplyr::select(-rn, -term)
write.xlsx(meta_CP_final, paste0(outdir, "/CBX_meta_CPAdj_final.xlsx"))

rm(meta_df, fix_res, rand_res, het_res, meta_raw, eff_idx1, eff_idx2, err_idx1, err_idx2)
```

```{r plot the results}
# plot the meta-analyzed results
# prepare dataframe for plotting
plot_df <- meta_CP_final %>%
  dplyr::select(XVar:estimate_MCSA, p.value_MCSA:q_global_MCSA, estimate_ADNI, p.value_ADNI:q_global_ADNI, beta_meta, p_meta:q_global_meta) %>% 
  dplyr::rename(
    estimate_meta = beta_meta,
    p.value_meta = p_meta,
    ) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(XVar=factor(XVar, levels=c("Memory", "LMDR", "HippVol")),
         ADJUST = factor(ADJUST, levels = c("None", "All", ct_ana)),
         Cohort = ifelse(Cohort=="meta", "Meta-analyzed", Cohort),
         Cohort = factor(Cohort, levels=c("MCSA", "ADNI","Meta-analyzed")),
         lbl = if_else(q_global<0.05, "✱✱",  ""), 
         lbl = if_else(q_local<0.05 & q_global > 0.05, "✱", lbl), 
         lbl = if_else(p.value<0.05 & q_local>0.05,"Δ", lbl)
         )

cairo_pdf(paste0(outdir, "/CBX_assoc_adjustbyCP.pdf"), width = 10, height = 6.5, onefile = T)
plot_df %>% 
  filter(Cohort == "Meta-analyzed") %>% 
  ggplot(aes(x=ADJUST, y=YVar)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=9/.pt) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  scale_y_discrete(limits = rev) +
  labs(
    x = "Adjustment",
    caption = "Δ:P<0.05; ✱:local q<0.05; ✱✱:global q<0.05") +
  facet_grid(.~XVar) +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic"),
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.title.y = element_blank(),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        text = element_text(family = "ArialMT"))
dev.off()
```

```{r plot the correlations b/w estimates}
plot_df2 <- plot_df %>% 
  filter(ADJUST == "None") %>% 
  dplyr::select(XVar:estimate) %>% 
  dplyr::rename(est_original = estimate) %>% 
  dplyr::select(-ADJUST) %>% 
  left_join(
    plot_df %>% 
      filter(ADJUST != "None") %>% 
      dplyr::select(XVar:estimate) %>% 
      dplyr::rename(est_new = estimate)
  )

pdf(paste0(outdir, "/CBX_CPAdjAssocEst_corr_.pdf"), width = 7.25, height = 3, onefile = T)
plot_df2 %>%
  filter(Cohort == "Meta-analyzed" & YVar != "ME3") %>% 
  ggplot(aes(x = est_original, y = est_new)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_grid(XVar ~ ADJUST, scales = "free") +
  # Add correlation coefficient (r) and p-value
  stat_cor(
    method = "pearson",
    label.x.npc = "left",   # position (can be "left", "right", etc.)
    label.y.npc = "top",
    label.sep = "\n",
    size = 7/.pt
  ) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle=25, vjust=1))
dev.off() # this is extended data fig.14b

```


## Longitudinal association
```{r prepare input files for longitudinal association}
# load in longitudinal files of mcsa and adni
  # for both files, rows are donors, and there are at least two or more rows per donor because there are data from different time point.
  # columns are cognitive endophenotypes/hippcampal volume at each time point. 
  # there is also a column of difference in years of each time point from the baseline. 
  # and there are columns of demographic information including sex, education yrs, age at baseline, intracranial volume or MRI magnetic field
mcsa_long <- read.csv(paste0(indir,"/mcsa_long.csv")) %>% mutate(SubjectID = as.character(SubjectID))
adni_long <- read.csv(paste0(indir, "/adni_long.csv"))

# check for each subject, how many missing across visits there are for different endophenotypes 
mcsa_na_check = mcsa_long %>% group_by(SubjectID) %>% summarise_all(list(~sum(!is.na(.))))
adni_na_check = adni_long %>% group_by(SubjectID) %>% summarise_all(list(~sum(!is.na(.))))


comRes <- readRDS(paste0(residir, "/comRes.rds")) # from 1_CreateResids.R

# Prepare association input files
cp4longassoc_mcsa <- mcsa_long %>% 
  dplyr::select(SubjectID, YrDiff, Sex, Educ, Age_bl, APOE4, LMDR, Memory, HippVol, ICV) %>% 
  left_join(
    mcsa_cp_clr %>% cbind(comTrait[1:105, ] %>% rownames_to_column("SubjectID") %>% dplyr::select(SubjectID))
  ) %>%
  left_join(comRes[1:105, ] %>% as.data.frame() %>% rownames_to_column("SubjectID") %>% dplyr::select(SubjectID, any_of(nduf_bb$gene_id))) %>% 
  rename_with(~ nduf_bb$gene_name[match(.x, nduf_bb$gene_id)], .cols = any_of(nduf_bb$gene_id))

cp4longassoc_adni <- adni_long %>% 
  dplyr::select(SubjectID, YrDiff, Sex, Educ, Age_bl, APOE4, LMDR, Memory, HippVol, ICV, Mag) %>% 
  mutate(Mag = factor(Mag)) %>% 
  left_join(
    adni_cp_clr %>% cbind(comTrait[106:196, ] %>% rownames_to_column("SubjectID") %>% dplyr::select(SubjectID))
  ) %>%
  left_join(comRes[106:196, ] %>% as.data.frame() %>% rownames_to_column("SubjectID") %>% dplyr::select(SubjectID, any_of(nduf_bb$gene_id))) %>% 
  rename_with(~ nduf_bb$gene_name[match(.x, nduf_bb$gene_id)], .cols = any_of(nduf_bb$gene_id))

```

```{r associations functions}
# functions for performing associations for:
# 1) no adjustment (the original model);
# 2) adjust for all analyzed cell proportions;
# 3) adjust for one cell proportion at a time
long_assoc_fn <- function(df,
                          na_check_df,
                          xvars,
                          yvars,
                          adjust_types,
                          base_covs,
                          base_covs_for_HV,
                          cp_covs
                          ){
  run_assoc <- function(xvar, yvar, adjustment){
    message("Test association between ", xvar, " and ", yvar, " adjusting for ", adjustment)
    # no cell type prportion adjustment
    if (adjustment == "None"){
      if (yvar %in% c("Memory", "LMDR")){
        inputdf <- df %>% filter(! SubjectID %in% na_check_df$SubjectID[which(na_check_df[[yvar]]<=1)])
        m <- lmer(paste0(yvar, " ~ ", xvar, "*YrDiff", " + ", paste0(base_covs, collapse = " + "), " + (1 + YrDiff|SubjectID)"), data = inputdf)
      }
      else if (yvar == "HippVol"){
        inputdf <- df %>% filter(! SubjectID %in% na_check_df$SubjectID[which(na_check_df[[yvar]]<=1)])
        m <- lmer(paste0(yvar, " ~ ", xvar, "*YrDiff", " + ", paste0(base_covs_for_HV, collapse = " + "), " + (1 + YrDiff|SubjectID)"), data = inputdf)
      }
    }
    # adjust for all cell type proportions
    else if (adjustment == "All"){
      if (yvar %in% c("Memory", "LMDR")){
        inputdf <- df %>% filter(! SubjectID %in% na_check_df$SubjectID[which(na_check_df[[yvar]]<=1)])
        m <- lmer(paste0(yvar, " ~ ", xvar, "*YrDiff", " + ", paste0(c(base_covs, cp_covs), collapse = " + "), " + (1 + YrDiff|SubjectID)"), data = inputdf)
      }
      else if (yvar == "HippVol"){
        inputdf <- df %>% filter(! SubjectID %in% na_check_df$SubjectID[which(na_check_df[[yvar]]<=1)])
        m <- lmer(paste0(yvar, " ~ ", xvar, "*YrDiff", " + ", paste0(c(base_covs_for_HV, cp_covs), collapse = " + "), " + (1 + YrDiff|SubjectID)"), data = inputdf)
      }
    }
    # adjust for one cell type proportion at a time
    else if (adjustment %in% c(cp_covs)){
      if (yvar %in% c("Memory", "LMDR")){
        inputdf <- df %>% filter(! SubjectID %in% na_check_df$SubjectID[which(na_check_df[[yvar]]<=1)])
        m <- lmer(paste0(yvar, " ~ ", xvar, "*YrDiff", " + ", paste0(c(base_covs, adjustment), collapse = " + "), " + (1 + YrDiff|SubjectID)"), data = inputdf)
      }
      else if (yvar == "HippVol"){
        inputdf <- df %>% filter(! SubjectID %in% na_check_df$SubjectID[which(na_check_df[[yvar]]<=1)])
        m <- lmer(paste0(yvar, " ~ ", xvar, "*YrDiff", " + ", paste0(c(base_covs_for_HV, adjustment), collapse = " + "), " + (1 + YrDiff|SubjectID)"), data = inputdf)
      }
    }
    
    res_df <- tidy(m, conf.int = T)
    
    tbl <- tibble(
      Model = list(m),
      SummDf = list(res_df)
    )
    return(tbl)
  }
  
  # Run for all combinations
  results <- expand.grid(
    XVar = xvars,
    YVar = yvars,
    ADJUST = adjust_types,
    stringsAsFactors = F
  ) %>%
  mutate(result = pmap(list(XVar, YVar, ADJUST),
                       ~ run_assoc(..1, ..2, ..3))) %>%
  unnest(result)
  
  return(results)
}
```

```{r run longitudinal assoc}
mcsa_longassoc <- long_assoc_fn(df = cp4longassoc_mcsa,
                                na_check_df = mcsa_na_check,
                                yvars = c("Memory", "LMDR", "HippVol"),
                                xvars = c(nduf_bb$gene_name),
                                adjust_types = c("None", "All", ct_ana),
                                base_covs = c("Age_bl", "Sex", "Educ"),
                                base_covs_for_HV = c("Age_bl", "Sex", "Educ", "ICV"),
                                cp_covs = ct_ana
                                )
mcsa_longassoc_df <- mcsa_longassoc %>% 
  dplyr::select(XVar, YVar, ADJUST, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, ":")) %>% 
  group_by(YVar, ADJUST) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes/ME tested (n=20)
  ungroup() %>% 
  group_by(YVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for numer of genes x adjustment tested (20*9)
  ungroup() %>% 
  dplyr::select(-effect, -group)

adni_longassoc <- long_assoc_fn(df = cp4longassoc_adni,
                                na_check_df = adni_na_check,
                                yvars = c("Memory", "LMDR", "HippVol"),
                                xvars = c(nduf_bb$gene_name),
                                adjust_types = c("None", "All", ct_ana),
                                base_covs = c("Age_bl", "Sex", "Educ"),
                                base_covs_for_HV = c("Age_bl", "Sex", "Educ", "ICV", "Mag"),
                                cp_covs = ct_ana
                                )
adni_longassoc_df <- adni_longassoc %>% 
  dplyr::select(XVar, YVar, ADJUST, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, ":")) %>% 
  group_by(YVar, ADJUST) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes/ME tested (n=20)
  ungroup() %>% 
  group_by(YVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for numer of genes x adjustment tested (20*9)
  ungroup()

save(mcsa_longassoc, adni_longassoc, file = paste0(outdir, "/CBX_CPAdj_LongAssoc.RData"))
```

```{r longitudinal meta-analysis}
meta_df = inner_join(mcsa_longassoc_df, adni_longassoc_df, by = c("XVar","YVar", "ADJUST", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(rn = paste0(XVar, "_", YVar, "_", ADJUST)) %>% 
  column_to_rownames("rn")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "rn")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "rn") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "rn")

# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_CP_long_final = meta_df %>%
  group_by(YVar, ADJUST) %>%
  mutate(
    q_local_random = p.adjust(p_random, method = "fdr"),
    q_local_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(YVar) %>% 
  mutate(
    q_global_random = p.adjust(p_random, method = "fdr"),
    q_global_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for numer of genes x adjustment tested (21*9)
  ungroup() %>% 
  mutate(model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_local_meta = ifelse(model_meta =="Fixed", q_local_fixed, q_local_random),
         q_global_meta = ifelse(model_meta =="Fixed", q_global_fixed, q_global_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random)
         ) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"))) %>% 
  dplyr::select(-rn, -term)
write.xlsx(meta_CP_long_final, paste0(outdir, "/CBX_meta_CPAdj_long_final.xlsx"))

rm(meta_df, fix_res, rand_res, het_res, meta_raw, eff_idx1, eff_idx2, err_idx1, err_idx2)
```

```{r longitudinal: plot the results}
plot_df <- meta_CP_long_final %>%
  dplyr::select(XVar:estimate_MCSA, p.value_MCSA:q_global_MCSA, estimate_ADNI, p.value_ADNI:q_global_ADNI, beta_meta, p_meta:q_global_meta) %>% 
  dplyr::rename(
    estimate_meta = beta_meta,
    p.value_meta = p_meta,
    ) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(YVar=factor(YVar, levels=c("Memory", "LMDR", "HippVol")),
         ADJUST = factor(ADJUST, levels = c("None", "All", ct_ana)),
         Cohort = ifelse(Cohort=="meta", "Meta-analyzed", Cohort),
         Cohort = factor(Cohort, levels=c("MCSA", "ADNI","Meta-analyzed")),
         lbl = if_else(q_global<0.05, "✱✱",  ""), 
         lbl = if_else(q_local<0.05 & q_global > 0.05, "✱", lbl), 
         lbl = if_else(p.value<0.05 & q_local>0.05,"Δ", lbl)
         )


cairo_pdf(paste0(outdir, "/CBX_longassoc_adjustbyCP.pdf"), width = 10, height = 6.5, onefile = T)
plot_df %>% 
  filter(Cohort == "Meta-analyzed") %>% 
  ggplot(aes(x=ADJUST, y=XVar)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=9.5/.pt) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  scale_y_discrete(limits = rev) +
  labs(
    x = "Adjustment",
    caption = "Δ:P<0.05; ✱:local q<0.05; ✱✱:global q<0.05") +
  facet_grid(.~YVar) +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic"),
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.title.y = element_blank(),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        text = element_text(family = "ArialMT"))
dev.off() # this is extended data fig.16a
```

```{r longitudinal: plot the correlations b/w estimates}
plot_df2 <- plot_df %>% 
  filter(ADJUST == "None") %>% 
  dplyr::select(XVar:estimate) %>% 
  dplyr::rename(est_original = estimate) %>% 
  dplyr::select(-ADJUST) %>% 
  left_join(
    plot_df %>% 
      filter(ADJUST != "None") %>% 
      dplyr::select(XVar:estimate) %>% 
      dplyr::rename(est_new = estimate)
  )

pdf(paste0(outdir, "/CBX_CPAdjlong_est_corr.pdf"), width = 7.25, height = 3, onefile = T)
plot_df2 %>%
  filter(Cohort == "Meta-analyzed") %>% 
  ggplot(aes(x = est_original, y = est_new)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_grid(YVar ~ ADJUST, scales = "free") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle=25, vjust=1))
dev.off() # this is extended data fig.16b
```


# Cell-specific gene expression analysis
## Prepare dataframes
- For high resolution gene expression (GE) files, need to remove genes that were either “NA” or “1”, or had the same deconvoluted expression values across all samples, which indicated insufficient evidence of estimation. (https://www.nature.com/articles/s41598-023-44183-7#Sec11).
- The resulting GE files from CIBERSORTx are in non-log space so we performed log2 transformation.
```{r}
mcsa_gep_filepath <- list.files(path = paste0(indir,"/CBXHiRes_upload"), pattern="^CIBERSORTxHiRes_R01_", full.names = T) # the file format is each column is subject and each row is a gene
mcsa_gep_filepath <- mcsa_gep_filepath[-c(2,3,8)] #remove path to GE files of cell types not analyzed

adni_gep_filepath <- list.files(path = paste0(indir,"/CBXHiRes_upload"), pattern="^CIBERSORTxHiRes_ADNI_s91_", full.names = T)
adni_gep_filepath <- adni_gep_filepath[-c(2,3,8)] #remove path to GE files of cell types not analyzed

# readFile: to read all the .txt files in the specified path
readFile <- function(f) {df <- read_delim(file = f, delim = "\t")}
# read .txt files and prepare for association analysis
mcsa_gep_ls <- lapply(mcsa_gep_filepath, readFile)
names(mcsa_gep_ls) <- c("Bcells", "Mast", "Monocytes", "Neutrophils", "NK", "CD4T", "CD8T")
mcsa_gep_clean <- mcsa_gep_ls %>% 
  map(~ filter(.x, GeneSymbol %in% nduf_bb$gene_name)) %>% 
  map(~ rowwise(.x)) %>% 
  map(~filter(.x, n_distinct(c_across(-1), na.rm = TRUE) > 1)) %>%
  map(~mutate(.x, across(2:ncol(.x), ~ log2(.x)))) %>% 
  keep(~ nrow(.x) > 0) %>% 
  map(~ pivot_longer(.x, cols = 2:ncol(.x), names_to = "SubjectID", values_to = "GE")) %>% 
  map(~ pivot_wider(.x, names_from = "GeneSymbol", values_from = "GE")) %>% 
  map(~ left_join(.x,
                    comTrait[1:105,] %>% 
                       rownames_to_column("SubjectID") %>% 
                       dplyr::select(SubjectID, Age, Sex, Educ, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV))) %>% 
  map(~ left_join(.x, 
                    mcsa_covars %>% dplyr::select(SubjectID, RIN, PAXgene_flowcell, PAXgene_Batch)))
# read .txt files and prepare for association analysis
adni_gep_ls <- lapply(adni_gep_filepath, readFile)
names(adni_gep_ls) <- c("Bcells", "Mast", "Monocytes", "Neutrophils", "NK", "CD4T", "CD8T")
adni_gep_clean <- adni_gep_ls %>% 
  map(~ rename_with(.x, ~ gsub("^X", "", .x), .cols = -1)) %>% 
  map(~ filter(.x, GeneSymbol %in% nduf_bb$gene_name)) %>% 
  map(~ rowwise(.x)) %>% 
  map(~filter(.x, n_distinct(c_across(-1), na.rm = TRUE) > 1)) %>%
  map(~mutate(.x, across(2:ncol(.x), ~ log2(.x)))) %>% 
  keep(~ nrow(.x) > 0) %>% 
  map(~ pivot_longer(.x, cols = 2:ncol(.x), names_to = "SubjectID", values_to = "GE")) %>% 
  map(~ pivot_wider(.x, names_from = "GeneSymbol", values_from = "GE")) %>% 
  map(~ left_join(.x,
                   comTrait[106:196,] %>% 
                       rownames_to_column("SubjectID") %>% 
                       dplyr::select(SubjectID, Age, Sex, Educ, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV) %>% 
                       mutate(Mag_HippICV = factor(Mag_HippICV)))) %>% 
  map(~ left_join(.x, 
                     adni_covars %>% dplyr::select(SubjectID, RIN, AffyPlate.meanREE, SITE.meanREE)))
```

## Run cell-specific associations
```{r HiResGEP: assoc}
mcsa_gep_results <- list()
for (ct in names(mcsa_gep_clean)){
  data = mcsa_gep_clean[[ct]]
  ndufs_in_mx = intersect(names(data), nduf_bb$gene_name)
  
  results <- assoc_fn(df = data, 
                      xvars = c("Memory", "LMDR", "HippVol"),
                      yvars = ndufs_in_mx,
                      adjust_types = "None",
                      base_covs = c("Age", "Sex", "Educ", "RIN", "PAXgene_flowcell", "PAXgene_Batch"),
                      base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "RIN", "PAXgene_flowcell", "PAXgene_Batch"),
                      cp_covs = NULL)
  results <- results %>% dplyr::select(-ADJUST)
  
  mcsa_gep_results[[ct]] <- results
}
mcsa_gep_results <- bind_rows(mcsa_gep_results, .id = "CellType")
rm(ct, data, ndufs_in_mx, results)

mcsa_gep_df <- mcsa_gep_results %>% 
  dplyr::select(CellType, XVar, YVar, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, XVar)) %>% 
  group_by(XVar, CellType) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes tested in each cell type
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for number of genes x cell types tested
  ungroup()

adni_gep_results <- list()
for (ct in names(adni_gep_clean)){
  data = adni_gep_clean[[ct]]
  ndufs_in_mx = intersect(names(data), nduf_bb$gene_name)
  
  results <- assoc_fn(df = data, 
                      xvars = c("Memory", "LMDR", "HippVol"),
                      yvars = ndufs_in_mx,
                      adjust_types = "None",
                      base_covs = c("Age", "Sex", "Educ", "RIN", "AffyPlate.meanREE", "SITE.meanREE"),
                      base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "Mag_HippICV", "RIN", "AffyPlate.meanREE", 'SITE.meanREE'),
                      cp_covs = NULL)
  results <- results %>% dplyr::select(-ADJUST)
  
  adni_gep_results[[ct]] <- results
}
adni_gep_results <- bind_rows(adni_gep_results, .id = "CellType")
rm(ct, data, ndufs_in_mx, results)
adni_gep_df <- adni_gep_results %>% 
  dplyr::select(CellType, XVar, YVar, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, XVar)) %>% 
  group_by(XVar, CellType) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes tested in each cell type
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for number of genes x cell types tested
  ungroup()

save(mcsa_gep_results, adni_gep_results, file = paste0(outdir, "/HiResGEP_Results.RData"))
```

```{r HiResGEP: meta-analysis}
meta_df = inner_join(mcsa_gep_df, adni_gep_df, by = c("CellType", "XVar","YVar", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(rn = paste0(CellType, "_", XVar, "_", YVar, "_")) %>% 
  column_to_rownames("rn")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "rn")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "rn") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "rn")

# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_gep_final = meta_df %>%
  group_by(XVar, CellType) %>%
  mutate(
    q_local_random = p.adjust(p_random, method = "fdr"),
    q_local_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(
    q_global_random = p.adjust(p_random, method = "fdr"),
    q_global_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for numer of genes x adjustment tested (21*7)
  ungroup() %>% 
  mutate(model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_local_meta = ifelse(model_meta =="Fixed", q_local_fixed, q_local_random),
         q_global_meta = ifelse(model_meta =="Fixed", q_global_fixed, q_global_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random)
         ) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"))) %>% 
  dplyr::select(-rn, -term)
write.xlsx(meta_gep_final, paste0(outdir, "/CBX_meta_gep_final.xlsx")) # this is in supplementary table 13b

rm(meta_df, fix_res, rand_res, het_res, meta_raw, eff_idx1, eff_idx2, err_idx1, err_idx2)
```

```{r HiRes: plot the meta-analyzed GEP results}
plot_df <- meta_gep_final %>%
  dplyr::select(CellType:estimate_MCSA, p.value_MCSA:q_global_MCSA, estimate_ADNI, p.value_ADNI:q_global_ADNI, beta_meta, p_meta:q_global_meta) %>% 
  dplyr::rename(
    estimate_meta = beta_meta,
    p.value_meta = p_meta,
    ) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(XVar=factor(XVar, levels=c("Memory", "LMDR", "HippVol")),
         Cohort = ifelse(Cohort=="meta", "Meta-analyzed", Cohort),
         Cohort = factor(Cohort, levels=c("MCSA", "ADNI","Meta-analyzed")),
         lbl = if_else(q_global<0.05, "✱✱",  ""), 
         lbl = if_else(q_local<0.05 & q_global > 0.05, "✱", lbl), 
         lbl = if_else(p.value<0.05 & q_local>0.05,"Δ", lbl)
         )


cairo_pdf(paste0(outdir, "/CBX_GEPAssoc.pdf"), width = 6, height = 4.5)
plot_df %>% 
  ggplot(aes(x=Cohort, y=YVar)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=9.5/.pt) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  scale_y_discrete(limits = rev) +
  labs(
    x = "Adjustment",
    caption = "Δ:P<0.05; ✱:local q<0.05; ✱✱:global q<0.05") +
  facet_grid(CellType~XVar, scales = "free", space = "free") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic"),
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.title.y = element_blank(),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        text = element_text(family = "ArialMT"))
dev.off() # this is in Extended Data fig.14c
```
