---
title: "rbt_2_3"
output: html_document
date: "2026-02-16"
---

Author: Wei (Adelyn) Tsai; [tsai.wei\@mayo.edu](mailto:tsai.wei@mayo.edu){.email} If you are using this script, please cite our study This script addresses comment 3 about APOE4 sensitivity analysis Reviewer 2. 

We performed APOE4 sensitivity analysis for:
- cross-sectional antemortem cohorts for Bl_M3 and 20 NDUFs
- longitudinal antemortem cohorts for 20 NDUFs

```{r setup, include=FALSE}
suppressPackageStartupMessages(library(tidyverse)) #v2.0.0
suppressPackageStartupMessages(library(patchwork)) #v1.3.2
suppressPackageStartupMessages(library(broom)) #v1.0.12
suppressPackageStartupMessages(library(openxlsx)) #v4.2.8
suppressPackageStartupMessages(library(Hmisc)) #v5.1
suppressPackageStartupMessages(library(pbapply)) #v1.7
suppressPackageStartupMessages(library(ggpubr)) #v0.6.0
suppressPackageStartupMessages(library(lmerTest)) #v3.1
suppressPackageStartupMessages(library(broom.mixed)) #v0.2.9
suppressPackageStartupMessages(library(ggh4x)) #v0.3.1

set.seed(123)

# list of input/output directories
indir <- "./indir"
blwgcna_dir <- "./outdir/bl_wgcna"
brwgcna_dir <- "./outdir/br_wgcna"
nduf_assoc_dir <- "./outdir/nduf_assoc_de"
residir <- "./outdir/resids"
long_dir <- "./outdir/longitudinal"

# output directory
outdir <- "./Rebuttal/outdir/2_3"

source("./Codes/fns.R") #source a list of frequently used functions
```


# Prepare input dataframes
## Cross-sectional antemortem
```{r}
comTrait <- readRDS(paste0(indir, "/comTrait.rds")) # a file having cross-sectional endophenotypes for MCSA (from row 1-105) and ADNI (from row 106-196); row names are donorID
MEs1 <- read.csv(paste0(blwgcna_dir, "/MEs1_SP12.csv")) #from 2_blwgcna.R
nduf_bb <- read.csv(paste0(nduf_assoc_dir, "/nduf_bb.csv")) #generated in 4_nduf_assoc_de.R

cqn_raw <- read_delim(paste0(indir, "/PaX108_R01resilience_gene_CQN_neg3_postQC_s105.txt"), delim = "\t") # MCSA cqn file for 18046 genes. Each row is a gene and each column is a donor
# transform the CQN file to be long format
cqn_ndufs <- cqn_raw %>% filter(GeneId %in% nduf_bb$gene_id) %>% 
  dplyr::select(-Chromosome, -Start, -End, -Length, -GeneId, -GeneBiotype) %>%
  pivot_longer(-GeneName, names_to = "SubjectID", values_to = "Expr") %>%
  pivot_wider(names_from = "GeneName", values_from = "Expr")
mcsa_covars <- read_delim(paste0(indir,"/R01_Resilience_PAXgene_Covars.txt"), delim = "\t") %>% mutate(SubjectID=as.character(ptnum)) #covariates file for MCSA, rows are donor columns are covariates
# mcsa input file for cross-sectional analysis
mcsa_df_cs <- comTrait[1:105,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Age, Age_last, Sex, Educ, APOE4, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV) %>% 
  left_join(mcsa_covars %>% dplyr::select(SubjectID, RIN, PAXgene_flowcell, PAXgene_Batch)) %>% # select technical variables of RNAsequencing for adjustment
  left_join(MEs1 %>% dplyr::select(SubjectID, ME3)) %>% 
  left_join(cqn_ndufs) 
  
  
adni_mc <- readRDS(paste0(indir,"/ADNI_s91_microarray.rds")) #microarray file for 91 donors from ADNI. This has 10116 protein-coding genes, gene id as column names, donorID as row names
adni_covars <- read_delim(paste0(residir,"/ADNI_s91_covars.txt"), delim = "\t") #covariates file for 91 donors from ADNI. Rows are donors columns are covariates
# select only NDUFs in the microarray file and change the column name from gene ID to gene name
mc_ndufs <- adni_mc %>% dplyr::select(nduf_bb$gene_id)
mc_ndufs <- mc_ndufs %>% rename_at(vars(1:(ncol(.))), ~ nduf_bb$gene_name[match(names(mc_ndufs)[1:(ncol(mc_ndufs))], nduf_bb$gene_id)]) 
# adni input file for cross-sectional analysis
adni_df_cs <- comTrait[106:196,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Age, Age_last, Sex, Educ, APOE4, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV) %>% 
  mutate(Mag_HippICV = factor(Mag_HippICV)) %>% 
  left_join(adni_covars %>% dplyr::select(SubjectID, RIN, AffyPlate.meanREE, SITE.meanREE)) %>% # select technical variables of RNAsequencing for adjustment
  left_join(MEs1 %>% dplyr::select(SubjectID, ME3)) %>% 
  left_join(mc_ndufs %>% rownames_to_column("SubjectID")) 
```



## Longitudinal antemortem 
```{r}
comRes <- readRDS(paste0(residir, "/comRes.rds")) # from 1_CreateResids.R, 
# load in longitudinal files of mcsa and adni
  # for both files, rows are donors, and there are at least two or more rows per donor because there are data from different time point.
  # columns are cognitive endophenotypes/hippcampal volume at each time point. 
  # there is also a column of difference in years of each time point from the baseline. 
  # and there are columns of demographic information including sex, education yrs, age at baseline, intracranial volume or MRI magnetic field
mcsa_long <- read.csv(paste0(indir,"/mcsa_long.csv")) %>% mutate(SubjectID = as.character(SubjectID))
adni_long <- read.csv(paste0(indir, "/adni_long.csv"))

# check for each subject, how many missing across visits there are for different endophenotypes 
mcsa_na_check = mcsa_long %>% group_by(SubjectID) %>% summarise_all(list(~sum(!is.na(.))))
adni_na_check = adni_long %>% group_by(SubjectID) %>% summarise_all(list(~sum(!is.na(.))))

# input file for longitudinal association
mcsa_df <- mcsa_long %>% 
  dplyr::select(SubjectID, YrDiff, Sex, Educ, Age_bl, APOE4, LMDR, Memory, HippVol, ICV) %>% 
  left_join(comRes[1:105, ] %>% as.data.frame() %>% rownames_to_column("SubjectID") %>% dplyr::select(SubjectID, any_of(nduf_bb$gene_id)))
# rename the columns from NDUF gene ID to gene name
mcsa_df <- mcsa_df %>% rename_at(vars(11:(ncol(.))), ~ nduf_bb$gene_name[match(names(mcsa_df)[11:(ncol(mcsa_df))], nduf_bb$gene_id)])

# input file for longitudinal association
adni_df <- adni_long %>% 
  dplyr::select(SubjectID, YrDiff, Sex, Educ, Age_bl, APOE4, LMDR, Memory, HippVol, ICV, Mag) %>% 
  left_join(comRes[106:196, ] %>% as.data.frame() %>% rownames_to_column("SubjectID") %>% dplyr::select(SubjectID, any_of(nduf_bb$gene_id)))
# rename the columns from NDUF gene ID to gene name
adni_df <- adni_df %>% rename_at(vars(12:(ncol(.))), ~ nduf_bb$gene_name[match(names(adni_df)[12:(ncol(adni_df))], nduf_bb$gene_id)])

```


# Run associations adjusting for APOE4 dose
## Cross-sectional antemortem
```{r mcsa - cross-sectional}
mcsa_models_apoe4adj <- list()
mcsa_coefs_apoe4adj <- list()
for (i in c("Memory", "LMDR", "HippVol")){
  if (i != "HippVol"){
    for (j in c("ME3", nduf_bb$gene_name)){
      if (j == "ME3"){
        model <- lm(paste0(j, "~", i, "+ Age + Sex + Educ + APOE4"), data = mcsa_df_cs)
      }
      else{
        model <- lm(paste0(j, "~", i, "+ Age + Sex + Educ + APOE4 + RIN + PAXgene_flowcell + PAXgene_Batch"), data = mcsa_df_cs)
      }
      df <- tidy(model, conf.int = T)
  
      mcsa_models_apoe4adj[[paste0("CrossSectional_", i, "_", j)]] <- model
      mcsa_coefs_apoe4adj[[paste0("CrossSectional_", i, "_", j)]] <- df
    }
  }
  else{
    for (j in c("ME3", nduf_bb$gene_name)){
      if (j == "ME3"){
        model <- lm(paste0(j, "~", i, "+ Age + Sex + Educ + APOE4 + Hipp_ICV"), data = mcsa_df_cs)
      }
      else{
        model <- lm(paste0(j, "~", i, "+ Age + Sex + Educ + APOE4 + Hipp_ICV + RIN + PAXgene_flowcell + PAXgene_Batch"), data = mcsa_df_cs)
      }
      df <- tidy(model, conf.int = T)
  
      mcsa_models_apoe4adj[[paste0("CrossSectional_", i, "_", j)]] <- model
      mcsa_coefs_apoe4adj[[paste0("CrossSectional_", i, "_", j)]] <- df
    }
  }
}
```

```{r adni - cross-sectional}
adni_models_apoe4adj <- list()
adni_coefs_apoe4adj <- list()
for (i in c("Memory", "LMDR", "HippVol")){
  if (i != "HippVol"){
    for (j in c("ME3", nduf_bb$gene_name)){
      if (j == "ME3"){
        model <- lm(paste0(j, "~", i, "+ Age + Sex + Educ + APOE4"), data = adni_df_cs)
      }
      else{
        model <- lm(paste0(j, "~", i, "+ Age + Sex + Educ + APOE4 + RIN + AffyPlate.meanREE + SITE.meanREE"), data = adni_df_cs)
      }
      df <- tidy(model, conf.int = T)
  
      adni_models_apoe4adj[[paste0("CrossSectional_", i, "_", j)]] <- model
      adni_coefs_apoe4adj[[paste0("CrossSectional_", i, "_", j)]] <- df
    }
  }
  else{
    for (j in c("ME3", nduf_bb$gene_name)){
      if (j == "ME3"){
        model <- lm(paste0(j, "~", i, "+ Age + Sex + Educ + APOE4 + Hipp_ICV + Mag_HippICV"), data = adni_df_cs)
      }
      else{
        model <- lm(paste0(j, "~", i, "+ Age + Sex + Educ + APOE4 + Hipp_ICV + Mag_HippICV + RIN + AffyPlate.meanREE + SITE.meanREE"), data = adni_df_cs)
      }
      df <- tidy(model, conf.int = T)
  
      adni_models_apoe4adj[[paste0("CrossSectional_", i, "_", j)]] <- model
      adni_coefs_apoe4adj[[paste0("CrossSectional_", i, "_", j)]] <- df
    }
  }
}
```


## Longitudinal antemortem
```{r mcsa longitudinal}
for (i in c("Memory", "LMDR", "HippVol")){

  if (i != "HippVol"){
    inputdf <- mcsa_df %>% 
      filter(! SubjectID %in% mcsa_na_check$SubjectID[which(mcsa_na_check[[i]]<=1)]) # only retain subjects with more than 1 longitudinal measures
    for (j in c(nduf_bb$gene_name)){
      model <- lmer(paste0(i, "~", j , "*YrDiff + Age_bl + Sex + Educ + APOE4 + (1 + YrDiff|SubjectID)"), data=inputdf)
      
      df <- tidy(model, conf.int = T)
      
      mcsa_models_apoe4adj[[paste0("Longitudinal_", i,"_",j)]] <- model
      mcsa_coefs_apoe4adj[[paste0("Longitudinal_", i,"_",j)]] <- df
    }
    
  } else {
    inputdf <- mcsa_df %>% 
      filter(! SubjectID %in% mcsa_na_check$SubjectID[which(mcsa_na_check[[i]]<=1)])
    for (j in c(nduf_bb$gene_name)){
      model <- lmer(paste0(i, "~", j, "*YrDiff + Age_bl + Sex + Educ + ICV + APOE4 + (1 + YrDiff|SubjectID)"), data=inputdf)
      
      df <- tidy(model, conf.int = T)
      
      mcsa_models_apoe4adj[[paste0("Longitudinal_", i,"_",j)]] <- model
      mcsa_coefs_apoe4adj[[paste0("Longitudinal_", i,"_",j)]] <- df
    }
  }
}
```

```{r adni longitudinal}
for (i in c("Memory", "LMDR", "HippVol")){

  if (i != "HippVol"){
    inputdf <- adni_df %>% 
      filter(! SubjectID %in% adni_na_check$SubjectID[which(adni_na_check[[i]]<=1)]) # only retain subjects with more than 1 longitudinal measures
    for (j in c(nduf_bb$gene_name)){
      model <- lmer(paste0(i, "~", j , "*YrDiff + Age_bl + Sex + Educ + APOE4 + (1 + YrDiff|SubjectID)"), data=inputdf)
      
      df <- tidy(model, conf.int = T)
      
      adni_models_apoe4adj[[paste0("Longitudinal_", i,"_",j)]] <- model
      adni_coefs_apoe4adj[[paste0("Longitudinal_", i,"_",j)]] <- df
    }
    
  } else {
    inputdf <- adni_df %>% 
      filter(! SubjectID %in% adni_na_check$SubjectID[which(adni_na_check[[i]]<=1)])
    for (j in c(nduf_bb$gene_name)){
      model <- lmer(paste0(i, "~", j, "*YrDiff + Age_bl + Sex + Educ + ICV + Mag + APOE4 + (1 + YrDiff|SubjectID)"), data=inputdf)
      
      df <- tidy(model, conf.int = T)
      
      adni_models_apoe4adj[[paste0("Longitudinal_", i,"_",j)]] <- model
      adni_coefs_apoe4adj[[paste0("Longitudinal_", i,"_",j)]] <- df
    }
  }
}

```


# Organize results (antemortem)
```{r}
# cross-sectional
mcsa_coefs_cs <- mcsa_coefs_apoe4adj[which(startsWith(names(mcsa_coefs_apoe4adj), "CrossSectional_"))] %>% 
  bind_rows(.id="x") %>% 
  separate(col = "x", into = c("ModelType","Phenotype", "NDUF"), sep = "_") %>%
  filter(Phenotype == term) %>% 
  group_by(Phenotype) %>% 
  mutate(q.value = p.adjust(p.value, "fdr"))
adni_coefs_cs <- adni_coefs_apoe4adj[which(startsWith(names(adni_coefs_apoe4adj), "CrossSectional_"))]  %>% 
  bind_rows(.id="x") %>% 
  separate(col = "x", into = c("ModelType","Phenotype", "NDUF"), sep = "_") %>%
  filter(Phenotype == term) %>% 
  group_by(Phenotype) %>% 
  mutate(q.value = p.adjust(p.value, "fdr"))

# longitudinal
mcsa_coefs_long <- mcsa_coefs_apoe4adj[which(startsWith(names(mcsa_coefs_apoe4adj), "Longitudinal_"))]  %>% 
  bind_rows(.id="x") %>% 
  separate(col = "x", into = c("ModelType", "Phenotype", "NDUF"), sep = "_") %>%
  filter(str_detect(term, ":")) %>%
  group_by(Phenotype) %>% 
  mutate(q.value = p.adjust(p.value, "fdr")) %>% 
  dplyr::select(-effect, -group)
adni_coefs_long <- adni_coefs_apoe4adj[which(startsWith(names(adni_coefs_apoe4adj), "Longitudinal_"))]  %>% 
  bind_rows(.id="x") %>% 
  separate(col = "x", into = c("ModelType", "Phenotype", "NDUF"), sep = "_") %>%
  filter(str_detect(term, ":")) %>%
  group_by(Phenotype) %>% 
  mutate(q.value = p.adjust(p.value, "fdr")) %>% 
  dplyr::select(-effect, -group)
```

# Meta-analysis (antemortem)
## Cross-sectional
```{r}
# prepare files
options(stringsAsFactors = FALSE)
pboptions(type = "timer")
meta_df = inner_join(mcsa_coefs_cs, adni_coefs_cs, by = c("ModelType", "Phenotype", "NDUF", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(Pheno_NDUF = paste0(Phenotype, "_", NDUF)) %>% 
  column_to_rownames("Pheno_NDUF")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "Pheno_NDUF") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "Pheno_NDUF") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "Pheno_NDUF")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "Pheno_NDUF") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "Pheno_NDUF")


# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_cs_final = meta_df %>%
  group_by(Phenotype) %>%
  mutate(q_random = p.adjust(p_random, method = "fdr"),
         q_fixed = p.adjust(p_fixed, method = "fdr"),
         model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_meta = ifelse(model_meta =="Fixed", q_fixed, q_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random)) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"))) %>% 
  dplyr::select(-Pheno_NDUF, -term)


rm(fix_res, rand_res, het_res, meta_df, meta_raw, eff_idx1, eff_idx2, err_idx1, err_idx2)

```

## Longitudinal
```{r}
# prepare files
options(stringsAsFactors = FALSE)
pboptions(type = "timer")
meta_df = inner_join(mcsa_coefs_long, adni_coefs_long, by = c("ModelType", "Phenotype", "NDUF", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(Pheno_NDUF = paste0(Phenotype, "_", NDUF)) %>% 
  column_to_rownames("Pheno_NDUF")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "Pheno_NDUF") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "Pheno_NDUF") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "Pheno_NDUF")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "Pheno_NDUF") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "Pheno_NDUF")


# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_long_final = meta_df %>%
  group_by(Phenotype) %>%
  mutate(q_random = p.adjust(p_random, method = "fdr"),
         q_fixed = p.adjust(p_fixed, method = "fdr"),
         model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_meta = ifelse(model_meta =="Fixed", q_fixed, q_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random)) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"))) %>% 
  dplyr::select(-Pheno_NDUF, -term)


rm(fix_res, rand_res, het_res, meta_df, meta_raw, eff_idx1, eff_idx2, err_idx1, err_idx2)

write.xlsx(list("CrossSectional" = meta_cs_final, "Longitudinal" = meta_long_final), paste0(outdir, "/meta_assoc_antemortem_e4.xlsx"))
```

# Make plots
## Dot plots
```{r dot plots}
# prepare files for making the plots
plot_df_cs <- meta_cs_final %>%
  dplyr::select(Phenotype:estimate_MCSA, p.value_MCSA, q.value_MCSA, estimate_ADNI, p.value_ADNI, q.value_ADNI, beta_meta, p_meta, q_meta) %>% 
  dplyr::rename(
    estimate_meta = beta_meta,
    p.value_meta = p_meta,
    q.value_meta = q_meta
    ) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(Phenotype=factor(Phenotype, levels=c("Memory", "LMDR", "HippVol")),
         Cohort = ifelse(Cohort=="meta", "Meta-analyzed", Cohort),
         Cohort = factor(Cohort, levels=c("Meta-analyzed", "ADNI", "MCSA")),
         NDUF = if_else(NDUF == "ME3", "Bl_M3", NDUF),
         lbl = if_else(q.value<0.05, "✱", ""), 
         lbl = if_else(p.value<0.05 & q.value > 0.05,"Δ", lbl)
         )

plot_df_long <- meta_long_final %>%
  dplyr::select(Phenotype:estimate_MCSA, p.value_MCSA, q.value_MCSA, estimate_ADNI, p.value_ADNI, q.value_ADNI, beta_meta, p_meta, q_meta) %>% 
  dplyr::rename(
    estimate_meta = beta_meta,
    p.value_meta = p_meta,
    q.value_meta = q_meta
    ) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(Phenotype=factor(Phenotype, levels=c("Memory", "LMDR", "HippVol")),
         Cohort = ifelse(Cohort=="meta", "Meta-analyzed", Cohort),
         Cohort = factor(Cohort, levels=c("Meta-analyzed", "ADNI", "MCSA")),
         lbl = if_else(q.value<0.05, "✱", ""), 
         lbl = if_else(p.value<0.05 & q.value > 0.05,"Δ", lbl)
         )

cairo_pdf(paste0(outdir, "/apoe4adj_cs_long_assoc_dotplots.pdf"), width = 6, height = 3.5, onefile = T)
plot_df_cs %>% 
  ggplot(aes(x=NDUF, y=Cohort)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=8.5/.pt) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  scale_y_discrete(limits = rev) +
  labs(
    title = "Cross-sectional analysis adjusted for APOE4 dose",
    x = "Bl_M3/NDUFs",
    y = "",
    caption = "Δ:P<0.05\n✱:q<0.05") +
  facet_grid(Phenotype ~. ) +
  theme_minimal(base_size = 9.5) +
  theme(axis.text.x = element_text(face = "italic", angle = 30, hjust = 1),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5),
        text = element_text(family = "ArialMT")) # this is extended data fig.8a
plot_df_long %>% 
  ggplot(aes(x=NDUF, y=Cohort)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=8.5/.pt) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  scale_y_discrete(limits = rev) +
  labs(
    title = "Longitudinal analysis adjusted for APOE4 dose",
    x = "NDUFs",
    y = "",
    caption = "Δ:P<0.05\n✱:q<0.05") +
  facet_grid(Phenotype ~. ) +
  theme_minimal(base_size = 9.5) +
  theme(axis.text.x = element_text(face = "italic", angle = 30, hjust = 1),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5),
        text = element_text(family = "ArialMT")) # this is extended data fig.8c
dev.off()
```

## Correlation plots
Plot correlation between estimates before and after adjusting for APOE4
### antemoretem
```{r organize dataframes for making plot}
# acquire estimates before adjusting for APOE4
metr <- read.csv(paste0(blwgcna_dir, "/module_trait.csv")) # this is created from 2_bl_wgcna.R
nduf_assoc <- read.csv(paste0(nduf_assoc_dir, "/Bl_Assoc.csv")) # this is created from 4_nduf_assoc_de.R
long_assoc <- read.xlsx(paste0(long_dir, "/longitudinal_new.xlsx"), sheet = "Meta") # this is from 5_longitudinal_new.R

# cross-sectional original results
og_cs <- metr %>% 
  dplyr::select(Module, Pheno, estimate_MCSA, estimate_ADNI, beta_meta) %>% 
  filter(Module == "ME3" & Pheno %in% c("Memory", "LMDR", "HippVol")) %>% 
  rename(NDUF = Module, Phenotype = Pheno) %>% 
  rbind(
    nduf_assoc %>% 
      dplyr::select(gene_name, Phenotype, estimate_MCSA, estimate_ADNI, beta_meta) %>% 
      rename(NDUF = gene_name)
  ) %>% 
  dplyr::rename(estimate_meta = beta_meta) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(Cohort = ifelse(Cohort == "meta", "Meta-analyzed", Cohort),
         NDUF = ifelse(NDUF == "ME3", "Bl_M3", NDUF))
# longitudinal original results
og_long <- long_assoc %>% 
  dplyr::select(Phenotype, NDUF, Estimate_MCSA, Estimate_ADNI, beta_meta) %>% 
  dplyr::rename(estimate_MCSA = Estimate_MCSA, estimate_ADNI = Estimate_ADNI, estimate_meta = beta_meta) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(Cohort = ifelse(Cohort == "meta", "Meta-analyzed", Cohort))

plot_df_cs2 <- plot_df_cs %>% 
  dplyr::select(Phenotype:estimate) %>% 
  dplyr::rename(est_APOE4Adjusted = estimate) %>% 
  left_join(
    og_cs %>% dplyr::rename(est_NoAPOE4Adjusted = estimate)
  ) %>% 
  mutate(Cohort = factor(Cohort, levels = c("MCSA", "ADNI", "Meta-analyzed")),
         Phenotype = factor(Phenotype, levels = c("Memory", "LMDR", "HippVol")))
plot_df_long2 <- plot_df_long %>% 
  dplyr::select(Phenotype:estimate) %>% 
  dplyr::rename(est_APOE4Adjusted = estimate) %>% 
  left_join(
    og_long %>% dplyr::rename(est_NoAPOE4Adjusted = estimate)
  ) %>% 
  mutate(Cohort = factor(Cohort, levels = c("MCSA", "ADNI", "Meta-analyzed")),
         Phenotype = factor(Phenotype, levels = c("Memory", "LMDR", "HippVol")))

```


```{r correlation plots}
pdf(paste0(outdir, "/apoe4adj_cs_long_est_corr.pdf"), width = 3, height = 3, onefile = T)
plot_df_cs2 %>%
  filter(NDUF != "Bl_M3") %>%
  ggplot(aes(x = est_NoAPOE4Adjusted, y = est_APOE4Adjusted)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_grid(Cohort ~ Phenotype, scales = "free") +
  labs(title = "Cross-sectional Estimate Correlation") +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 20, hjust = .75)) # this is extended data fig.8b
plot_df_long2 %>%
  filter(NDUF != "Bl_M3") %>%
  ggplot(aes(x = est_NoAPOE4Adjusted, y = est_APOE4Adjusted)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_grid(Cohort ~ Phenotype, scales = "free") +
  labs(title = "Longitudinal Estimate Correlation") +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 20, hjust = .75)) # extended data fig.8d
dev.off()
```
