---
title: "rbt_2_1_3_BF_resid"
output: html_document
date: "2026-01-30"
---

Author: Wei (Adelyn) Tsai; [tsai.wei\@mayo.edu](mailto:tsai.wei@mayo.edu){.email} If you are using this script, please cite our study This script addresses comment 1 about number of resilience per diagnosis from Reviewer 2. This script also addresses comment B and F from Reviewer 3 regarding prediction on resilience status

Please not that prediction results are updated based on comment from 2nd rebuttal. Please refer to rebuttal_2nd/rbt_1_6.Rmd for updated results.

# Libraries and directories

```{r}
suppressPackageStartupMessages(library(tidyverse)) #v2.0.0
suppressPackageStartupMessages(library(tidymodels)) #v1.4.1
suppressPackageStartupMessages(library(openxlsx)) #v4.2.8
suppressPackageStartupMessages(library(ggcorrplot)) #v0.1.4.1
suppressPackageStartupMessages(library(patchwork)) #v1.3.2
suppressPackageStartupMessages(library(ggfortify)) #v0.4.19
suppressPackageStartupMessages(library(broom)) #v1.0.12
suppressPackageStartupMessages(library(pbapply)) #v1.7


set.seed(123)

# list of input/output directories
indir <- "./indir"
residir <- "./outdir/resids"
blwgcna_dir <- "./outdir/bl_wgcna"
nduf_assoc_dir <- "./outdir/nduf_assoc_de"

# output directory
outdir <- "./Rebuttal/outdir"

source("./Codes/fns.R") #source a list of frequently used functions
```

# Calculate CR using residuals

```{r}
# load needed files
comTrait <- readRDS(paste0(indir,"/comTrait.rds")) # a file having cross-sectional endophenotypes for MCSA (from row 1-105) and ADNI (from row 106-196); row names are donorID

## we extracted Abeta centiloids for calculating cognitive resilience based on Abeta levels
## we used Abeta centiloid two years from visit date for blood collection

mcsa_pet = readxl::read_excel(paste0(indir, "/MCSA_Updated_PET_data.xlsx")) # a file with PET imaging data from MCSA; relevant PiB columns including:
# pibdate: date of PiB
# pibmanufacturer: manufacturer of PiB tracer
# pib_suvr
# pib_centiloid
# abnormal_pib = cutoff is 25 Centiloid (1.52 SUVR) for positivity
adni_pet <- read_csv(paste0(indir, "/ADSP_PHC_PET_Amyloid_Simple_04Nov2025.csv")) # file with PET imaging data from ADNI; effort of ADSP harmonization; relevant AV45 columns include:
# PHC_SCANDATE: PET scan date
# PHC_TRACER
# PHC_QC_IMAGE: QC of image quality
# PHC_QC_TIMING: QC of imaging acquisition time
# PHC_CL_FAIL: projected centiloid values for images that failed QC_TIMING
# PHC_CENTILOIDS
# PHC_AMYLOID_STATUS: cutoff is 25 centiloid for positivity


mcsa_ab <- mcsa_pet %>% 
  mutate(SubjectID = as.character(ptnum)) %>% 
  dplyr::select(SubjectID, pibdate, pibmanufacturer, pib_suvr, pib_centiloid, abnormal_pib) %>% 
  inner_join(comTrait %>% rownames_to_column("SubjectID") %>% dplyr::select(SubjectID, PaxDt)) %>% #PaxDt = date of blood collection visit
  mutate(daydiff = abs(difftime(PaxDt, pibdate))) %>% 
  group_by(SubjectID) %>% 
  arrange(daydiff) %>% 
  slice_head() #98 people

adni_ab <- adni_pet %>% 
  dplyr::select(PTID, PHC_SCANDATE, PHC_TRACER, PHC_QC_IMAGE, PHC_QC_TIMING, PHC_CENTILOIDS, PHC_CL_FAIL, PHC_AMYLOID_STATUS) %>% 
  dplyr::rename(SubjectID = PTID) %>% 
  inner_join(comTrait %>% rownames_to_column("SubjectID") %>% dplyr::select(SubjectID, PaxDt)) %>% 
  mutate(daydiff = abs(difftime(PaxDt, PHC_SCANDATE, units = "days"))) %>% 
  group_by(SubjectID) %>% 
  arrange(daydiff) %>% 
  slice_head() # 80 people; all images passed QC

# combine data of Abeta with covariates (age, sex, education, APOE4) and cognition (memory/LMDR)
mcsa_df <- comTrait[1:105,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Sex, Educ, Age, APOE4, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV, PaxDx, LastDx, PaxDx_binary, LastDx_binary) %>%
  mutate(SubjectID = paste0("s_", SubjectID)) %>% 
  left_join(mcsa_ab %>% mutate(SubjectID = paste0("s_", SubjectID)))
rownames(mcsa_df) <- mcsa_df$SubjectID

adni_df <- comTrait[106:196,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Sex, Educ, Age, APOE4, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV, PaxDx, LastDx, PaxDx_binary, LastDx_binary) %>%
  left_join(adni_ab)
rownames(adni_df) <- adni_df$SubjectID

save(mcsa_df, adni_df, file = paste0(outdir, "/2_1_3_BF/MCSA_ADNI_s196_DemoDf.RData"))


# establish CR calculation functions
update_geom_defaults("text", list(size = 5/.pt)) # update the default text size
cr_fn <- function(df, x, y, covars){
  print(paste0(x, "; ", y))
  fit = lm(paste0(y, " ~ ", paste0(c(x, covars), collapse = " + ")), data = df)
  Statdf <- augment(fit) %>% dplyr::rename(SubjectID = .rownames)
  CRdf <- Statdf %>% 
    dplyr::select(SubjectID, .resid, Age, Sex, Educ, APOE4) %>% 
    dplyr::rename(CR = .resid) %>% # CR = residuals of the models
    mutate(Resilience = if_else(CR > 0, "Resilient", "Non-resilient"),
           Resilience = factor(Resilience, levels = c("Non-resilient", "Resilient"))) # Resilience = resilience status
  # diagnose_p is to check whether the models meet the assumption of linear regression
  diagnose_p <- autoplot(fit) + labs(subtitle = paste0(y, " ~ ", paste0(c(x, covars), collapse = " + "))) + theme(text = element_text(size = 6.5))
  # scatter plot plotting the correlation between dependent and independent variables
  scatter_p <- df %>% 
    ggplot(aes_string(x = x, y = y)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(subtitle = paste0(y, " ~ ", paste0(c(x, covars), collapse = " + "))) +
    theme_classic(base_size = 6.5)
  
  result <- tibble(
    X = x,
    Y = y,
    Model = list(fit),
    Statdf = list(Statdf),
    CRdf = list(CRdf),
    Dx_P = list(diagnose_p),
    Scatter_P = list(scatter_p)
  )
  return(result)
}

mcsa_cr_2yrs <- rbind(
  cr_fn(df = subset(mcsa_df, daydiff<730), x = "pib_centiloid", y = "Memory", covars = c("Age", "Sex", "Educ", "APOE4", "daydiff")), # Y = Memory, regressing on centiloids and risk factors and demographics
  cr_fn(df = subset(mcsa_df, daydiff<730), x = "pib_centiloid", y = "LMDR", covars = c("Age", "Sex", "Educ", "APOE4", "daydiff")), # Y = LMDR, regressing on centiloids and risk factors and demographics
  cr_fn(df = mcsa_df, x = "Age", y = "Memory", covars = c("Sex", "Educ", "APOE4")), # Y = Memory, regressing on risk factors and demographics
  cr_fn(df = mcsa_df, x = "Age", y = "LMDR", covars = c("Sex", "Educ", "APOE4")) # Y = LMDR, regressing on risk factors and demographics
)
adni_cr_2yrs <- rbind(
  cr_fn(df = subset(adni_df, daydiff<730), x = "PHC_CENTILOIDS", y = "Memory", covars = c("Age", "Sex", "Educ", "APOE4", "daydiff")), # Y = Memory, regressing on centiloids and risk factors and demographics
  cr_fn(df = subset(adni_df, daydiff<730), x = "PHC_CENTILOIDS", y = "LMDR", covars = c("Age", "Sex", "Educ", "APOE4", "daydiff")), # Y = LMDR, regressing on centiloids and risk factors and demographics
  cr_fn(df = adni_df, x = "Age", y = "Memory", covars = c("Sex", "Educ", "APOE4")), # Y = Memory, regressing on risk factors and demographics
  cr_fn(df = adni_df, x = "Age", y = "LMDR", covars = c("Sex", "Educ", "APOE4")) # Y = LMDR, regressing on risk factors and demographics
)
#save
save(mcsa_cr_2yrs, adni_cr_2yrs, file = paste0(outdir, "/2_1_3_BF/CRlast_2yrs.RData"))


# also save a dataframe for organized from the calculated residuals
mcsa_CRdf_2yrs <- bind_rows(
  mcsa_cr_2yrs %>% dplyr::select(X, Y, CRdf) %>% 
    unnest(CRdf)
) %>% 
  dplyr::select(X, Y, SubjectID, CR, Resilience)
adni_CRdf_2yrs <- bind_rows(
  adni_cr_2yrs %>% dplyr::select(X, Y, CRdf) %>% 
    unnest(CRdf)
) %>% 
  dplyr::select(X, Y, SubjectID, CR, Resilience)

# organize the cognitive resilience status dataframe
mcsa_CRdf_2yrs <- mcsa_CRdf_2yrs %>% 
  filter(X == "Age" & Y == "Memory") %>% 
  dplyr::rename(CR_X.Age_Y.Mem = CR, Res_X.Age_Y.Mem = Resilience) %>% 
  dplyr::select(-X, -Y) %>% #has to start with scores calculating without centiloid first because there're more people in this model, so no people will be lost through left_join later
  left_join(
    mcsa_CRdf_2yrs %>% 
      filter(X == "Age" & Y == "LMDR") %>% 
      dplyr::rename(CR_X.Age_Y.LMDR = CR, Res_X.Age_Y.LMDR = Resilience) %>% 
      dplyr::select(-X, -Y)
  ) %>% 
  left_join(
    mcsa_CRdf_2yrs %>% 
      filter(X == "pib_centiloid" & Y == "Memory") %>% 
      dplyr::rename(CR_X.CL_Y.Mem = CR, Res_X.CL_Y.Mem = Resilience) %>% 
      dplyr::select(-X, -Y)
  ) %>% 
  left_join(
    mcsa_CRdf_2yrs %>% 
      filter(X == "pib_centiloid" & Y == "LMDR") %>% 
      dplyr::rename(CR_X.CL_Y.LMDR = CR, Res_X.CL_Y.LMDR = Resilience) %>% 
      dplyr::select(-X, -Y)
  ) %>% 
  left_join(
    mcsa_df %>% 
      dplyr::select(SubjectID, PaxDx, LastDx, PaxDx_binary, LastDx_binary, Memory, LMDR, HippVol:Mag_HippICV, pib_centiloid, Sex:Age, APOE4, daydiff)
  ) %>% 
  dplyr::select(SubjectID, PaxDx, LastDx, PaxDx_binary, LastDx_binary, Memory, LMDR, HippVol:Mag_HippICV, pib_centiloid, Sex:Age, APOE4, daydiff, everything())


adni_CRdf_2yrs <- adni_CRdf_2yrs %>% 
  filter(X == "Age" & Y == "Memory") %>% 
  dplyr::rename(CR_X.Age_Y.Mem = CR, Res_X.Age_Y.Mem = Resilience) %>% 
  dplyr::select(-X, -Y) %>% 
  left_join(
    adni_CRdf_2yrs %>% 
      filter(X == "Age" & Y == "LMDR") %>% 
      dplyr::rename(CR_X.Age_Y.LMDR = CR, Res_X.Age_Y.LMDR = Resilience) %>% 
      dplyr::select(-X, -Y)
  ) %>% 
  left_join(
    adni_CRdf_2yrs %>% 
      filter(X == "PHC_CENTILOIDS" & Y == "Memory") %>% 
      dplyr::rename(CR_X.CL_Y.Mem = CR, Res_X.CL_Y.Mem = Resilience) %>% 
      dplyr::select(-X, -Y)
  ) %>% 
  left_join(
    adni_CRdf_2yrs %>% 
      filter(X == "PHC_CENTILOIDS" & Y == "LMDR") %>% 
      dplyr::rename(CR_X.CL_Y.LMDR = CR, Res_X.CL_Y.LMDR = Resilience) %>% 
      dplyr::select(-X, -Y)
  ) %>% 
  left_join(
    adni_df %>% 
      dplyr::select(SubjectID, PaxDx, LastDx, PaxDx_binary, LastDx_binary, Memory, LMDR, HippVol:Mag_HippICV, PHC_CENTILOIDS, Sex:Age, APOE4, daydiff)
  ) %>% 
  dplyr::select(SubjectID, PaxDx, LastDx, PaxDx_binary, LastDx_binary, Memory, LMDR, HippVol:Mag_HippICV, PHC_CENTILOIDS, Sex:Age, APOE4, daydiff, everything())

# count the number of predicted resilience based on clinical diagnosis
mcsa_CRdf_2yrs %>% 
  dplyr::select(starts_with("Res_")) %>%
  mutate(Cohort = "MCSA") %>% 
  rbind(adni_CRdf_2yrs %>% 
          dplyr::select(starts_with("Res_")) %>% 
          mutate(Cohort = "ADNI")) %>% 
  pivot_longer(
    cols = starts_with("Res_"),
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(name = factor(name, levels = c("Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR", "Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR")),
         Cohort = factor(Cohort, levels = c("MCSA"), "ADNI")) %>% 
  group_by(name, Cohort) %>%
  summarise(
    'Resilient'  = sum(value == "Resilient", na.rm = TRUE),
    'Non-resilient' = sum(value == "Non-resilient", na.rm = TRUE),
    'NA'    = sum(is.na(value))
  ) %>%
  kableExtra::kbl(caption = "Resilient status based on residuals", format = "html", table.attr = "style='width:50%;'") %>%
  kableExtra::kable_styling() # this information is in supplementary table 14a


# save model summary
mcsa_cr_modelsum_2yrs = mcsa_cr_2yrs %>% mutate(model_sum = map(Model, tidy, conf.int = T)) %>% select(X, Y, model_sum) %>% unnest(model_sum)
adni_cr_modelsum_2yrs = adni_cr_2yrs %>% mutate(model_sum = map(Model, tidy, conf.int = T)) %>% select(X, Y, model_sum) %>% unnest(model_sum)

# save as excel
write.xlsx(list("MCSA" = mcsa_CRdf_2yrs, "ADNI" = adni_CRdf_2yrs), paste0(outdir, "/2_1_3_BF/CRdf_2yrs.xlsx"))
write.xlsx(list("MCSA" = mcsa_cr_modelsum_2yrs, "ADNI" = adni_cr_modelsum_2yrs), paste0(outdir, "/2_1_3_BF/ModelSumm_2yrs.xlsx"))

# save the resilience status in comTrait
comTrait %>% 
  rownames_to_column("SubjectID") %>% 
  left_join(
    bind_rows(
      mcsa_CRdf_2yrs %>% dplyr::select(SubjectID, Res_X.CL_Y.Mem, Res_X.CL_Y.LMDR, Res_X.Age_Y.Mem, Res_X.Age_Y.LMDR) %>% mutate(SubjectID = str_remove(SubjectID, "s_")),
      adni_CRdf_2yrs %>% dplyr::select(SubjectID, Res_X.CL_Y.Mem, Res_X.CL_Y.LMDR, Res_X.Age_Y.Mem, Res_X.Age_Y.LMDR)
    ), by = "SubjectID"
  ) %>% 
  column_to_rownames("SubjectID") %>% 
  saveRDS(paste0(indir, "/comTrait.rds"))
```

| Column name      | Residual method to define          |
|------------------|------------------------------------|
| Res_X.Age_Y.Mem  | No centiloids, memory as outcome   |
| Res_X.Age_Y.LMDR | No centiloids, LMDR as outcome     |
| Res_X.CL_Y.Mem   | With centiloids, memory as outcome |
| Res_X.CL_Y.LMDR  | With centiloids, LMDR as outcome   |

# predictive modeling functions

Adopted from: <https://lsinks.github.io/posts/2023-04-10-tidymodels/tidymodels_tutorial.html#defining-the-model-with-parsnip>

```{r}

modeling <- function(
    data,
    outcome_col,
    predictor_col,
    n_splits,
    train_prop,
    mode = c("separate", "all") # test all or separate genes at a time
){
  mode <- match.arg(mode)
  
  # remove any NA and select only needed columns
  data <- data %>% 
    drop_na(outcome_col, predictor_col) %>% 
    dplyr::select(all_of(c(outcome_col, predictor_col))) 
  
  # set up dataframe for metrics and coefficients ----
  metrics <- data.frame()
  coefs   <- data.frame()
  preds   <- data.frame()
  
  # Define ----
  ## 1) models ----
  ### standard logistic regression
  logreg_spec <- 
    logistic_reg() %>% 
    set_engine("glm") %>% 
    set_mode("classification")

  ### ridge regularization
  ridge_spec <- 
    logistic_reg(penalty = tune(),
                 mixture = 0) %>%
    set_engine("glmnet") %>% 
    set_mode("classification")

  ### elastic net regularization
  en_spec <- 
    logistic_reg(penalty = tune(),
                 mixture = tune()) %>%
    set_engine("glmnet") %>% 
    set_mode("classification")

  ## 2) metrics to evaluate ----
  custom_metric_set = metric_set(roc_auc, sensitivity, specificity)
  
  ## 3) formula ----
  formula = as.formula(paste0(outcome_col, " ~ ", paste0(predictor_col, collapse = " + ")))
  
  if (mode == "separate"){
    # Loop over splits (separate) ----
    for (s in seq_len(n_splits)){
      set.seed(s)
      ## Split the data into a test and training set
      data_split <- initial_split(data, prop = train_prop, strata = outcome_col)     
      
      ## Create data frames for the two sets:
      train_data <- training(data_split)
      test_data  <- testing(data_split)
      
      ## Create recipe
      recipe <- 
        recipe(formula, data = train_data) %>% 
        step_dummy(all_nominal_predictors())
      
      ## create workflow
      wf <- 
        workflow() %>% 
        add_recipe(recipe) %>% 
        add_model(logreg_spec)
  
      ## fitting
      last_fit_results <- last_fit(wf, data_split, metrics = custom_metric_set)
  
      ## collect coefficients
      coefs <- rbind(
        coefs,
        last_fit_results %>% extract_fit_parsnip() %>% tidy() %>% mutate(split = s)
      )
  
      ## collect metrics
      metrics <- rbind(
        metrics,
        last_fit_results %>% collect_metrics() %>% mutate(split = s)
      )
      
      ## collect predictions
      preds <- rbind(
        preds,
        last_fit_results %>% collect_predictions() %>% mutate(split = s)
      )
    }
    # Create a summary metrics dataframe over all the splits
    summary <- metrics %>% 
      summarise(
        mean = mean(.estimate, na.rm = T), 
        sd = sd(.estimate, na.rm = T), 
        .by = c(.metric)
      )
  }
  
  else if (mode == "all"){
    # Loop over splits (all) ----
    for (s in seq_len(n_splits)){
      print(paste0("At split: ", s))
      set.seed(s)
      ## Split the data into a test and training set
      data_split <- initial_split(data, prop = train_prop, strata = outcome_col)     
      
      ## Create data frames for the two sets:
      train_data <- training(data_split)
      test_data  <- testing(data_split)
      
      ## Create recipe
      recipe <- 
        recipe(formula, data = train_data) %>% 
        step_dummy(all_nominal_predictors())
      
      ## create workflow
      wf_tune <- 
        workflow_set(
          list(data = recipe),
          list(logreg = logreg_spec, ridge = ridge_spec, en = en_spec)
        )

      ## tuning
      set.seed(1234)
      folds <- vfold_cv(train_data, v = 10, strata = outcome_col)
      tune_results <- 
        workflow_map(
          wf_tune,
          "tune_grid",
          resamples = folds,
          grid = 10,
          metrics = custom_metric_set,
          verbose = F
        )
      
      ## get best parameters (for ridge and en) and fitting 
      
      ### logreg
      validation_logreg <- tune_results %>% extract_workflow("data_logreg") %>% last_fit(split =  data_split, metrics = custom_metric_set)
      
      ### ridge
      best_ridge <- tune_results %>% extract_workflow_set_result("data_ridge") %>% select_best(metric = "roc_auc")
      validation_ridge <- tune_results %>% extract_workflow("data_ridge") %>% finalize_workflow(best_ridge) %>% last_fit(split =  data_split, metrics = custom_metric_set)
      
      ### EN
      best_en <- tune_results %>% extract_workflow_set_result("data_en") %>% select_best(metric = "roc_auc")
      validation_en <- tune_results %>% extract_workflow("data_en") %>% finalize_workflow(best_en) %>% last_fit(split =  data_split, metrics = custom_metric_set)

      ## collect coefficients
      coefs <- rbind(
        coefs,
        validation_logreg %>% extract_fit_parsnip() %>% tidy() %>% dplyr::select(term, estimate) %>% mutate(penalty = NA, split = s, model_type = "logreg"),
        validation_ridge %>% extract_fit_parsnip() %>% tidy() %>% mutate(split = s, model_type = "ridge"),
        validation_en %>% extract_fit_parsnip() %>% tidy() %>% mutate(split = s, model_type = "en")
      )
  
      ## collect metrics
      metrics <- rbind(
        metrics,
        validation_logreg %>% collect_metrics() %>% mutate(split = s, model_type = "logreg"),
        validation_ridge %>% collect_metrics() %>% mutate(split = s, model_type = "ridge"),
        validation_en %>% collect_metrics() %>% mutate(split = s, model_type = "en")
      )
      
      ## collect predictions
      preds <- rbind(
        preds,
        validation_logreg %>% collect_predictions() %>% mutate(split = s, model_type = "logreg"),
        validation_ridge %>% collect_predictions() %>% mutate(split = s, model_type = "ridge"),
        validation_en %>% collect_predictions() %>% mutate(split = s, model_type = "en")
      )
    }
    # Create a summary metrics dataframe over all the splits
    summary <- metrics %>% 
      summarise(
        mean = mean(.estimate, na.rm = T), 
        sd = sd(.estimate, na.rm = T), 
        .by = c(.metric, model_type)
      )
  }
  
  
  
  
  # Final storage 
  list(
    metrics_all = metrics,
    summary     = summary,
    coef_all    = coefs
  )
}

run_modeling <- function(
  data,
  outcome_cols, 
  gene_vars,
  demo_vars,
  mode = c("separate", "all"),
  n_splits = 100,
  train_prop = 0.7
){
  mode <- match.arg(mode)
  
  # store all results by outcome
  results <- list()
  
  for (outcome in outcome_cols) {
    print(paste0("OUTCOME: ", outcome))
    
    results[[outcome]] <- list()
    
    ## ----------------------------------------------------------
    ## MODE 1: EACH GENE SEPARATELY
    ## ----------------------------------------------------------
    if (mode == "separate") {
      
      separate_list <- list()
      
      for (g in gene_vars) {
        pred_cols <- c(demo_vars, g)
        
        print(paste0("Running gene: ", g))
        
        separate_list[[g]] <-
          modeling(
            data          = data,
            outcome_col   = outcome,
            predictor_col = pred_cols,
            n_splits      = n_splits,
            train_prop    = train_prop,
            mode          = mode
          )
      }
      
      results[[outcome]] <- separate_list
    }
    
    ## ----------------------------------------------------------
    ## MODE 2: ALL GENES AT ONCE
    ## ----------------------------------------------------------
    if (mode == "all") {
      pred_cols <- c(demo_vars, gene_vars)
      
      print("Running: all genes together")
      
      results[[outcome]] <-
        modeling(
          data          = data,
          outcome_col   = outcome,
          predictor_col = pred_cols,
          n_splits      = n_splits,
          train_prop    = train_prop,
          mode          = mode
        )
    }
  }
  
  return(results)
}


```

# Prepare input files

The input files are used for: 1) predictive modelling 2) regression with diagnosis as a covariate 3) stratified analysis

```{r}
comRes <- readRDS(paste0(residir, "/comRes.rds")) # from 1_CreateResids.R
nduf_bb <- read.csv(paste0(nduf_assoc_dir, "/nduf_bb.csv")) #generated in 4_nduf_assoc_de.R
MEs1 <- read.csv(paste0(blwgcna_dir, "/MEs1_SP12.csv")) #from 2_blwgcna.R

# select the residuals of NDUFs
nduf_res <- comRes[,nduf_bb$gene_id]
nduf_res <- data.frame(nduf_res)
# rename geneID to gene name of NDUFs
nduf_res  <- nduf_res  %>% rename_at(vars(1:(ncol(.))), ~ nduf_bb$gene_name[match(names(nduf_res)[1:(ncol(nduf_res))], nduf_bb$gene_id)])

mcsa_covars <- read_delim(paste0(indir,"/R01_Resilience_PAXgene_Covars.txt"), delim = "\t") %>% mutate(SubjectID=as.character(ptnum)) #covariates file for MCSA, rows are donor columns are covariates
adni_covars <- read_delim(paste0(residir,"/ADNI_s91_covars.txt"), delim = "\t") #covariates file for 91 donors from ADNI. Rows are donors columns are covariates

# prepare input dataframes
mcsa_input <- mcsa_CRdf_2yrs %>% 
  mutate(SubjectID = str_remove(SubjectID, "s_")) %>% 
  mutate(
    across(starts_with("Res_X."),
           ~ as.character(.x))
  ) %>% 
  left_join(mcsa_covars %>% dplyr::select(SubjectID, RIN, PAXgene_flowcell, PAXgene_Batch)) %>%  # select technical variables of RNAsequencing for adjustment
  left_join(nduf_res %>% rownames_to_column("SubjectID")) %>% 
  left_join(MEs1 %>% dplyr::select(SubjectID, ME3)) %>% 
  mutate(APOE4 = factor(APOE4)) %>% 
  mutate(across(
    starts_with("Res_X."),
    ~ factor(.x, levels = c("Resilient", "Non-resilient"))
  )) %>% 
  mutate(across(
    ends_with("_binary"),
    ~ factor(.x, levels = c("Control", "Case"))
  ))
adni_input <- adni_CRdf_2yrs %>%
  mutate(
    across(starts_with("Res_X."),
           ~ as.character(.x))
  ) %>%
  left_join(adni_covars %>% dplyr::select(SubjectID, RIN, AffyPlate.meanREE, SITE.meanREE)) %>% # select technical variables of RNAsequencing for adjustment
  left_join(nduf_res %>% rownames_to_column("SubjectID")) %>% 
  left_join(MEs1 %>% dplyr::select(SubjectID, ME3)) %>% 
  mutate(APOE4 = factor(APOE4)) %>% 
  mutate(across(
    starts_with("Res_X."),
    ~ factor(.x, levels = c("Resilient", "Non-resilient"))
  )) %>% 
  mutate(across(
    ends_with("_binary"),
    ~ factor(.x, levels = c("Control", "Case"))
  ))
save(mcsa_input, adni_input, file = paste0(outdir, "/2_1_3_BF/inputdf.RData"))
```

# Make gene-gene corr plot
```{r}
mcsa_nduf_corr <- cor(mcsa_input[,nduf_bb$gene_name])
mcsa_nduf_corr_p <- cor_pmat(mcsa_input[,nduf_bb$gene_name])
adni_nduf_corr <- cor(adni_input[,nduf_bb$gene_name])
adni_nduf_corr_p <- cor_pmat(adni_input[,nduf_bb$gene_name])

pdf(paste0(outdir, "/2_1_3_BF/ggcorrplot.pdf"), width = 7, height = 4)
ggcorrplot(mcsa_nduf_corr, 
           type = "lower",
           colors = c("dodgerblue", "white", "firebrick"),
           p.mat = mcsa_nduf_corr_p,
           pch.cex = 0.8
) +
  labs(title = "MCSA NDUFs Correlation", x = "", y= "") +
  theme_classic(base_size = 6.5) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, face = "italic"),
        axis.text.y = element_text(face = "italic"),
        legend.key.size = unit(0.1, "in")) +
ggcorrplot(adni_nduf_corr, 
           type = "lower",
           colors = c("dodgerblue", "white", "firebrick"),
           p.mat = adni_nduf_corr_p,
           pch.cex = 0.8
) +
  labs(title = "ADNI NDUFs Correlation", x = "", y= "") +
  theme_classic(base_size = 6.5) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, face = "italic"),
        axis.text.y = element_text(face = "italic"),
        legend.key.size = unit(0.1, "in"))
dev.off() # this is extended data fig.12b
```


# Run the predictive modeling

Running these takes some time

```{r mcsa -resids run modelling}
mcsa_sep <- run_modeling(
  data = mcsa_input,
  outcome_cols = c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
  demo_vars = NULL,
  gene_vars = c("ME3", nduf_bb$gene_name),
  mode = "separate"
)

mcsa_all <- run_modeling(
  data = mcsa_input,
  outcome_cols = c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
  demo_vars = NULL,
  gene_vars = nduf_bb$gene_name,
  mode = "all"
)

# organize the results into dataframe
mcsa_sep_sum <- map_df(
  names(mcsa_sep),
  function(outcome) {
    map_df(
      names(mcsa_sep[[outcome]]),
      function(x) {
        mcsa_sep[[outcome]][[x]]$summary %>%
          mutate(Outcome = outcome, XVar = x)
      }
    )
  }
)
mcsa_all_sum <- map_df(
  names(mcsa_all),
  function(outcome) {
    mcsa_all[[outcome]]$summary %>%
      mutate(Outcome = outcome, Gene = "all_genes")
  }
)
```

```{r adni - resids run modelling}
adni_sep <- run_modeling(
  data = adni_input,
  outcome_cols = c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
  demo_vars = NULL,
  gene_vars = c("ME3", nduf_bb$gene_name),
  mode = "separate"
)

adni_all <- run_modeling(
  data = adni_input,
  outcome_cols = c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
  demo_vars = NULL,
  gene_vars = nduf_bb$gene_name,
  mode = "all"
)

# organize the results into dataframe
adni_sep_sum <- map_df(
  names(adni_sep),
  function(outcome) {
    map_df(
      names(adni_sep[[outcome]]),
      function(x) {
        adni_sep[[outcome]][[x]]$summary %>%
          mutate(Outcome = outcome, XVar = x)
      }
    )
  }
)
adni_all_sum <- map_df(
  names(adni_all),
  function(outcome) {
    adni_all[[outcome]]$summary %>%
      mutate(Outcome = outcome, Gene = "all_genes")
  }
)
```


## save and summarize results

```{r save results}
save(mcsa_sep, mcsa_all, adni_sep, adni_all,
     file = paste0(outdir, "/2_1_3_BF/resids2yrs_TM.RData"))
# calculate standard error and low and high 95% CI for metrics across 100 resampling splits
mcsa_all_sum <- mcsa_all_sum %>% 
  mutate(Cohort = "MCSA", .before = ".metric") %>% 
  mutate(
    se = sd / sqrt(100),
    L95 = mean - 1.96*se,
    H95 = mean + 1.96*se,
    .after = "sd"
  )
mcsa_sep_sum <- mcsa_sep_sum %>% 
  mutate(Cohort = "MCSA", .before = ".metric") %>% 
  mutate(
    se = sd / sqrt(100),
    L95 = mean - 1.96*se,
    H95 = mean + 1.96*se,
    .after = "sd"
  )


adni_all_sum <- adni_all_sum %>% 
  mutate(Cohort = "ADNI", .before = ".metric") %>% 
  mutate(
    se = sd / sqrt(100),
    L95 = mean - 1.96*se,
    H95 = mean + 1.96*se,
    .after = "sd"
  )
adni_sep_sum <- adni_sep_sum %>% 
  mutate(Cohort = "ADNI", .before = ".metric") %>% 
  mutate(
    se = sd / sqrt(100),
    L95 = mean - 1.96*se,
    H95 = mean + 1.96*se,
    .after = "sd"
  )

write.xlsx(list(
  "All" = rbind(mcsa_all_sum, adni_all_sum),
  "Sep" = rbind(mcsa_sep_sum, adni_sep_sum),
  "Sep_AUC_Check" = rbind(mcsa_sep_sum, adni_sep_sum) %>% filter(.metric == "roc_auc") %>%  group_by(XVar, Outcome) %>% mutate(BothAbove0.6 = ifelse(sum(mean>=0.6)>=2, TRUE, FALSE))
), paste0(outdir, "/2_1_3_BF/resids2yrs_summ_TM.xlsx"))
```

# Perform meta-analysis

Follow the guides from <https://pmc.ncbi.nlm.nih.gov/articles/PMC7925865>; <https://www.bmj.com/content/356/bmj.i6460>

```{r}
# function for meta-analysis 
auc_meta <- function(auc_data){ 
  require(metafor) #v4.8
  
  # Logit transform + delta method for variance
  auc_data$yi <- log(auc_data$auc / (1 - auc_data$auc))
  auc_data$vi <- auc_data$se^2 / (auc_data$auc * (1 - auc_data$auc))^2 #this formula is from Debray 2017 appendix 6
  
  # run random-effect meta-analysis
  res <- rma(yi = yi, vi = vi, data = auc_data, method = "REML")
  
  # back-transform logit pooled AUC and 95% CI
  backtrns <- data.frame(
    AUC = plogis(res$b),
    L95 = plogis(res$ci.lb),
    H95 = plogis(res$ci.ub)
  )
  # heterogeneity
  heterogeneity <- data.frame(
    Q    = res$QE,       # Cochran's Q statistic
    Qp   = res$QEp,      # p-value for Q
    I2   = res$I2,       # % variance due to heterogeneity
    tau2 = res$tau2,     # between-study variance (on logit scale)
    se.tau2  = res$se.tau2       # SD of true effects (sqrt of tau2)
  )
  # forest plot
  metafor::forest(res, transf = plogis, slab = auc_data$study, xlab = "AUC")
  forest_p <- recordPlot()
  
  results <- list(
    Meta_analysis = res,
    Backtrns_df = backtrns,
    Heterogeneity = heterogeneity,
    ForestP = forest_p
  )
  return(results)
}


```

```{r perform meta-analysis}
# meta-analysis for results testing one NDUF at a time
## organize the input dataframe for meta-analysis, this splits into a list of dataframe grouped by outcome variable ("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR") and gene 
both_auc_sep_sum <- mcsa_sep_sum %>% 
  filter(.metric == "roc_auc") %>% 
  mutate(study = "MCSA") %>% 
  rbind(adni_sep_sum %>% 
          filter(.metric == "roc_auc") %>% 
          mutate(study = "ADNI")
  ) %>% 
  mutate(n_splits = 100) %>% 
  rename(auc = mean) %>% 
  named_group_split(XVar, Outcome)

## run the meta-analysis and transform the back-transformed AUC and 95% CI and heterogeneity stats into dataframes
meta_auc_sep <- map(both_auc_sep_sum, ~ auc_meta(.x))
meta_auc_sep_combined <- meta_auc_sep %>%
  map("Backtrns_df") %>%
  bind_rows(.id = "XVar_Outcome") %>% 
  remove_rownames() %>% 
  separate(XVar_Outcome, into = c("XVar", "Outcome"), sep = " / ")
meta_het_sep_combined <- meta_auc_sep %>%
  map("Heterogeneity") %>%
  bind_rows(.id = "XVar_Outcome") %>% 
  remove_rownames() %>% 
  separate(XVar_Outcome, into = c("XVar", "Outcome"), sep = " / ")

#replayPlot(meta_auc_sep[["ME3 / Res_X.Age_Y.LMDR"]]$ForestP)

# meta-analysis for results testing all NDUFs at once
both_auc_all_sum <- mcsa_all_sum %>% 
  filter(.metric == "roc_auc") %>% 
  mutate(study = "MCSA") %>% 
  rbind(adni_all_sum %>% 
          filter(.metric == "roc_auc") %>% 
          mutate(study = "ADNI")
  ) %>% 
  mutate(n_splits = 100) %>% 
  rename(auc = mean) %>% 
  named_group_split(model_type, Outcome)

meta_auc_all <- map(both_auc_all_sum, ~ auc_meta(.x))
meta_auc_all_combined <- meta_auc_all %>%
  map("Backtrns_df") %>%
  bind_rows(.id = "ModelType_Outcome") %>% 
  remove_rownames() %>% 
  separate(ModelType_Outcome, into = c("ModelType", "Outcome"), sep = " / ")
meta_het_all_combined <- meta_auc_all %>%
  map("Heterogeneity") %>%
  bind_rows(.id = "ModelType_Outcome") %>% 
  remove_rownames() %>% 
  separate(ModelType_Outcome, into = c("ModelType", "Outcome"), sep = " / ")

```

```{r save meta-analysis results}
# save in excel sheet
wb <- loadWorkbook(paste0(outdir, "/2_1_3_BF/resids2yrs_summ_TM.xlsx"))
addWorksheet(wb, "meta_auc_sep")
addWorksheet(wb, "meta_het_sep")
addWorksheet(wb, "meta_auc_all")
addWorksheet(wb, "meta_het_all")
writeData(wb, sheet = "meta_auc_sep", x = meta_auc_sep_combined)
writeData(wb, sheet = "meta_het_sep", x = meta_het_sep_combined)
writeData(wb, sheet = "meta_auc_all", x = meta_auc_all_combined)
writeData(wb, sheet = "meta_het_all", x = meta_het_all_combined)
saveWorkbook(wb, paste0(outdir, "/2_1_3_BF/resids2yrs_summ_TM.xlsx"), overwrite = T) 
```

| Outcome              | Meta AUC  |
|----------------------|-----------|
| Res_X.Age_Consistent | 0.6647678 |
| Res_X.Age_Y.LMDR     | 0.6695683 |
| Res_X.Age_Y.Mem      | 0.5639113 |
| Res_X.CL_Consistent  | 0.6026014 |
| Res_X.CL_Y.LMDR      | 0.6225901 |
| Res_X.CL_Y.Mem       | 0.4944292 |

# Check number of people with memory, LMDR and HV per dx/resilience group
This is for reviewer 2 comment 1
```{r check number}
mcsa_input %>%
  mutate(across(c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"), ~ ifelse(.x == "Resilient", "Control", "Case"))) %>% 
  pivot_longer(cols = c("Memory", "LMDR", "HippVol"),
               names_to = "Phenotype",
               values_to = "PhenoValue") %>%
  pivot_longer(cols = c("LastDx_binary", "Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
               names_to = "Stratification",
               values_to = "Status") %>% 
  mutate(
    Phenotype = factor(Phenotype, levels = c("Memory", "LMDR", "HippVol")),
    Stratification = factor(Stratification, levels = c("LastDx_binary", "Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"))
  ) %>%
  filter(!is.na(PhenoValue)) %>%
  filter(Status == "Control") %>% 
  group_by(Phenotype, Stratification) %>%
  summarise(N = n(), .groups = "drop")

adni_input %>%
  mutate(across(c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"), ~ ifelse(.x == "Resilient", "Control", "Case"))) %>% 
  pivot_longer(cols = c("Memory", "LMDR", "HippVol"),
               names_to = "Phenotype",
               values_to = "PhenoValue") %>%
  pivot_longer(cols = c("LastDx_binary", "Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
               names_to = "Stratification",
               values_to = "Status") %>% 
  mutate(
    Phenotype = factor(Phenotype, levels = c("Memory", "LMDR", "HippVol")),
    Stratification = factor(Stratification, levels = c("LastDx_binary", "Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"))
  ) %>% 
  filter(!is.na(PhenoValue)) %>%
  filter(Status == "Control") %>% 
  group_by(Phenotype, Stratification) %>%
  summarise(N = n(), .groups = "drop")
```

# Plot

## Plot AUC results
These are in the first rebuttal Rebuttal Fig.28
```{r plot resids}
sep_df4p <- read.xlsx(paste0(outdir, "/2_1_3_BF/resids2yrs_summ_TM.xlsx"), sheet = "Sep")
all_df4p <- read.xlsx(paste0(outdir, "/2_1_3_BF/resids2yrs_summ_TM.xlsx"), sheet = "All")
meta_sep_df4p <- read.xlsx(paste0(outdir, "/2_1_3_BF/resids2yrs_summ_TM.xlsx"), sheet = "meta_auc_sep")
meta_all_df4p <- read.xlsx(paste0(outdir, "/2_1_3_BF/resids2yrs_summ_TM.xlsx"), sheet = "meta_auc_all")
# organize the dataframe for plotting
sep_forestp_df <- sep_df4p %>%
  filter(.metric == "roc_auc") %>% 
  dplyr::select(Cohort, XVar, Outcome, mean, L95, H95) %>% 
  rename(AUC = mean) %>% 
  rbind(meta_sep_df4p %>% mutate(Cohort = "Meta-analyzed")) %>% 
  mutate(Outcome = case_when(
           Outcome == "Res_X.Age_Y.Mem" ~ "Memory\n(no centiloids)",
           Outcome == "Res_X.Age_Y.LMDR" ~ "LMDR\n(no centiloids)",
           Outcome == "Res_X.CL_Y.Mem" ~ "Memory\n(with centiloids)",
           Outcome == "Res_X.CL_Y.LMDR" ~ "LMDR\n(with centiloids)",
           TRUE ~ NA_character_
         ),
         Outcome = factor(Outcome, levels = c("Memory\n(no centiloids)", "LMDR\n(no centiloids)", "Memory\n(with centiloids)", "LMDR\n(with centiloids)")),
         Cohort = factor(Cohort, levels = c("ADNI", "MCSA", "Meta-analyzed")))

pdf(paste0(outdir, "/2_1_3_BF/resids_AUC_forestp_TM.pdf"), height = 7, width = 7)
sep_forestp_df %>% 
  filter(!is.na(Outcome)) %>% 
  ggplot(aes(x=AUC, y=XVar, col=Cohort)) + 
    geom_errorbar(aes(xmax = H95, xmin = L95, color = Cohort), linewidth = 0.5, width = 0.5, position=position_dodge(width = 0.5)) + 
    geom_point(aes(color = Cohort), size = 2, position=position_dodge(width = 0.5)) +
    labs(x = "AUC") +
    facet_grid(. ~ Outcome, scales = "free") + 
    scale_y_discrete(limits = rev) + 
    scale_color_manual(values = c("ADNI" = "green4", "MCSA" = "blue4", "Meta-analyzed" = "cyan2")) +
    geom_vline(xintercept = 0.5, color = "red", linewidth = 0.4, linetype = 2) +
    geom_vline(xintercept = 0.6, color = "orange3", linewidth = 0.4, linetype = 2) +
    theme_bw() +
    theme(
      axis.text.y = element_text(face= "italic"),
      axis.title.y = element_blank(),
      plot.title = element_text(hjust = 0.5)) 
dev.off()

all_forestp_df <- all_df4p %>% 
  filter(.metric == "roc_auc") %>% 
  dplyr::select(Cohort, model_type, Outcome, mean, L95, H95) %>% 
  rename(ModelType = model_type, AUC = mean) %>% 
  rbind(meta_all_df4p %>% mutate(Cohort = "Meta-analyzed")) %>% 
  mutate(Outcome = case_when(
           Outcome == "Res_X.Age_Y.Mem" ~ "Memory\n(no centiloids)",
           Outcome == "Res_X.Age_Y.LMDR" ~ "LMDR\n(no centiloids)",
           Outcome == "Res_X.CL_Y.Mem" ~ "Memory\n(with centiloids)",
           Outcome == "Res_X.CL_Y.LMDR" ~ "LMDR\n(with centiloids)",
           TRUE ~ NA_character_
         ),
         Outcome = factor(Outcome, levels = c("Memory\n(no centiloids)", "LMDR\n(no centiloids)", "Memory\n(with centiloids)", "LMDR\n(with centiloids)")),
         Cohort = factor(Cohort, levels = c("ADNI", "MCSA", "Meta-analyzed")),
         ModelType = case_when(
           ModelType == "logreg" ~ "No penalization",
           ModelType == "ridge" ~ "Ridge",
           ModelType == "en" ~ "EN"
         ),
         ModelType = factor(ModelType, levels = c("EN", "Ridge","No penalization")))


pdf(paste0(outdir, "/2_1_3_BF/resids_allgenes_AUC_forestp_TM.pdf"), height = 3, width = 7)
all_forestp_df %>% 
  filter(!is.na(Outcome)) %>% 
  ggplot(aes(x=AUC, y=ModelType, col=Cohort)) + 
  geom_errorbar(aes(xmax = H95, xmin = L95, color = Cohort), linewidth = 0.5, width = 0.5, position=position_dodge(width = 0.5)) + 
  geom_point(aes(color = Cohort), size = 2, position=position_dodge(width = 0.5)) +
  labs(x = "AUC") +
  scale_color_manual(values = c("ADNI" = "green4", "MCSA" = "blue4", "Meta-analyzed" = "cyan2")) +
  geom_vline(xintercept = 0.5, color = "red", linewidth = 0.4, linetype = 2) +
  geom_vline(xintercept = 0.6, color = "orange3", linewidth = 0.4, linetype = 2) +
  facet_grid(. ~ Outcome, scales = "free") +
  theme_bw() +
  theme(
      axis.title.y = element_blank(),
      plot.title = element_text(hjust = 0.5))
dev.off()

# plot prioritized ones having meta AUC>=0.6 in both models with or without using centiloids for LMDR
pdf(paste0(outdir, "/2_1_3_BF/resids_AUC_forestp_TM_prioritized.pdf"), height = 3, width = 4.5)
sep_forestp_df %>% 
  filter(grepl("LMDR", Outcome)) %>% 
  filter(XVar %in% c("NDUFA9", "NDUFV1", "NDUFAF2", "NDUFS8", "NDUFB8")) %>% 
  ggplot(aes(x=AUC, y=XVar, col=Cohort)) + 
    geom_errorbar(aes(xmax = H95, xmin = L95, color = Cohort), linewidth = 0.5, width = 0.5, position=position_dodge(width = 0.5)) + 
    geom_point(aes(color = Cohort), size = 2, position=position_dodge(width = 0.5)) +
    labs(x = "AUC") +
    facet_grid(. ~ Outcome, scales = "free") + 
    scale_y_discrete(limits = rev) + 
    scale_x_continuous(labels = scales::label_number(accuracy = 0.01)) +
    scale_color_manual(values = c("ADNI" = "green4", "MCSA" = "blue4", "Meta-analyzed" = "cyan2")) +
    geom_vline(xintercept = 0.5, color = "red", linewidth = 0.4, linetype = 2) +
    geom_vline(xintercept = 0.6, color = "orange3", linewidth = 0.4, linetype = 2) +
    theme_bw() +
    theme(
      axis.text.y = element_text(face= "italic"),
      axis.title.y = element_blank(),
      axis.text.x = element_text(angle = 25, hjust = 1),
      plot.title = element_text(hjust = 0.5),
      legend.margin = margin(0, 0, 0, 0, "in"), 
      legend.text = element_text(size = 8), 
      legend.title = element_text(size = 8.5))
dev.off()
```

