---
title: "rbt_2_1"
output: html_document
date: "2026-03-04"
---

Author: Wei (Adelyn) Tsai; [tsai.wei\@mayo.edu](mailto:tsai.wei@mayo.edu){.email} If you are using this script, please cite our study This script addresses comment 1 about analysis correcting for diagnosis from Reviewer 2. 

# Libraries and directories

```{r}
suppressPackageStartupMessages(library(tidyverse)) #v2.0.0
suppressPackageStartupMessages(library(openxlsx)) #v4.2.8
suppressPackageStartupMessages(library(patchwork)) #v1.3.2
suppressPackageStartupMessages(library(broom)) #v1.0.12
suppressPackageStartupMessages(library(pbapply)) #v1.7
suppressPackageStartupMessages(library(ggpubr)) #v0.6.3


set.seed(123)

# list of input/output directories
indir <- "./indir"
residir <- "./outdir/resids"
blwgcna_dir <- "./outdir/bl_wgcna"
nduf_assoc_dir <- "./outdir/nduf_assoc_de"

# output directory
outdir <- "./Rebuttal/outdir"

source("./Codes/fns.R") #source a list of frequently used functions

load(paste0(outdir, "/2_1_3_BF/inputdf.RData")) # this is prepared from rbt_2_1_3_BF_TM.Rmd
nduf_bb <- read.csv(paste0(nduf_assoc_dir, "/nduf_bb.csv")) #generated in 4_nduf_assoc_de.R
```

# Define Functions
Prepare a linear function that allows me to conveniently run regression adjusting for diagnosis status or stratified by diagnosis status

```{r function}
linear_fn_new <- function(data_df,
                     XVars,
                     YVars, 
                     adj_stratify_covs, 
                     all_status,
                     base_covs,
                     base_covs_for_HV,
                     stratified = F){
  run_assoc <- function(xvar, yvar, adj_stratify_covar, status){
    
    df_use <- data_df
    # if stratified is True, filter the data to keep only certain status
    if (stratified){df_use <- df_use %>% filter(.data[[adj_stratify_covar]] == status)}
    # define the covariates
    if (xvar %in% c("Memory", "LMDR")){
      covars <- if (stratified) {base_covs} else {c(base_covs, adj_stratify_covar)}
      } 
    else if (xvar == "HippVol"){
      covars <- if (stratified) {base_covs_for_HV} else {c(base_covs_for_HV, adj_stratify_covar)}
      } 
    # run regression
    m <- lm(paste0(yvar, " ~ ", xvar, " + ", paste0(covars, collapse = " + ")), data = df_use)
    res_df <- broom::tidy(m, conf.int = T)
    
    # save results
    tbl <- tibble(
      Model = list(m),
      SummDf = list(res_df)
    )
    return(tbl)
  }
  
  # Run for all combinations
  results <- if (!stratified) {
    expand.grid(
      X = XVars,
      Y = YVars,
      AdjStratify_Covar = adj_stratify_covs,
      stringsAsFactors = FALSE
      ) %>% 
      mutate(result = pmap(list(X, Y, AdjStratify_Covar),
                         ~ run_assoc(..1, ..2, ..3))) %>%
      unnest(result)
  } 
  else {
    expand.grid(
      X = XVars,
      Y = YVars,
      AdjStratify_Covar = adj_stratify_covs,
      Stratify_Status = all_status,
      stringsAsFactors = FALSE
      ) %>% 
      filter(
        (AdjStratify_Covar == "LastDx_binary" & Stratify_Status %in% c("Control", "Case"))
        ) %>% 
    mutate(result = pmap(list(X, Y, AdjStratify_Covar, Stratify_Status),
                         ~ run_assoc(..1, ..2, ..3, ..4))) %>%
    unnest(result)
  }
  
  return(results)
}
```

Run regression adjusting for last diagnosis
```{r regression - adjust for dx}
mcsa_m3_dxadj <- linear_fn_new(data_df = mcsa_input, 
                        XVars = c("Memory", "LMDR", "HippVol"),
                        YVars = "ME3",
                        adj_stratify_covs = c("LastDx_binary"),
                        all_status = NULL,
                        base_covs = c("Age", "Sex", "Educ"),
                        base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV"),
                        stratified = F)
mcsa_nduf_dxadj <- linear_fn_new(data_df = mcsa_input, 
                        XVars = c("Memory", "LMDR", "HippVol"),
                        YVars = nduf_bb$gene_name,
                        adj_stratify_covs = c("LastDx_binary"),
                        all_status = NULL,
                        base_covs = c("Age", "Sex", "Educ", "RIN", "PAXgene_flowcell", "PAXgene_Batch"),
                        base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "RIN", "PAXgene_flowcell", "PAXgene_Batch"),
                        stratified = F)
mcsa_dxadj <- rbind(mcsa_m3_dxadj, mcsa_nduf_dxadj)
rm(mcsa_m3_dxadj, mcsa_nduf_dxadj)

adni_m3_dxadj <- linear_fn_new(data_df = adni_input, 
                        XVars = c("Memory", "LMDR", "HippVol"),
                        YVars = "ME3",
                        adj_stratify_covs = c("LastDx_binary"),
                        all_status = NULL,
                        base_covs = c("Age", "Sex", "Educ"),
                        base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "Mag_HippICV"),
                        stratified = F)
adni_nduf_dxadj <- linear_fn_new(data_df = adni_input, 
                        XVars = c("Memory", "LMDR", "HippVol"),
                        YVars = nduf_bb$gene_name,
                        adj_stratify_covs = c("LastDx_binary"),
                        all_status = NULL,
                        base_covs = c("Age", "Sex", "Educ", "RIN", "AffyPlate.meanREE", "SITE.meanREE"),
                        base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV",  "Mag_HippICV", "RIN", "AffyPlate.meanREE", "SITE.meanREE"),
                        stratified = F)
adni_dxadj <- rbind(adni_m3_dxadj, adni_nduf_dxadj)
rm(adni_m3_dxadj, adni_nduf_dxadj)
# organize the dataframe
dxadj_summ <- mcsa_dxadj %>% 
  mutate(Cohort = "MCSA") %>% 
  dplyr::select(Cohort, X, Y, AdjStratify_Covar, SummDf) %>% 
  unnest(SummDf) %>% 
  rbind(
    adni_dxadj %>% 
  mutate(Cohort = "ADNI") %>% 
  dplyr::select(Cohort, X, Y, AdjStratify_Covar, SummDf) %>% 
  unnest(SummDf)
  ) %>% 
  filter(X == term) %>% 
  group_by(Cohort, X) %>% 
  mutate(q.value = p.adjust(p.value, method = "fdr"), .after = "p.value") 
```
Run regression stratifying for last diagnosis
```{r regression - stratified for dx}
mcsa_m3_stratified <-  linear_fn_new(data_df = mcsa_input, 
                           XVars = c("Memory", "LMDR", "HippVol"),
                           YVars = "ME3",
                           adj_stratify_covs = c("LastDx_binary"),
                           all_status = c("Case", "Control"),
                           base_covs = c("Age", "Sex", "Educ"),
                           base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV"),
                           stratified = T)
mcsa_nduf_stratified <-  linear_fn_new(data_df = mcsa_input, 
                           XVars = c("Memory", "LMDR", "HippVol"),
                           YVars = nduf_bb$gene_name,
                           adj_stratify_covs = c("LastDx_binary"),
                           all_status = c("Case", "Control"),
                           base_covs = c("Age", "Sex", "Educ", "RIN", "PAXgene_flowcell", "PAXgene_Batch"),
                           base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "RIN", "PAXgene_flowcell", "PAXgene_Batch"),
                           stratified = T)
mcsa_stratified = rbind(mcsa_m3_stratified, mcsa_nduf_stratified)
rm(mcsa_m3_stratified, mcsa_nduf_stratified)

adni_m3_stratified <-  linear_fn_new(data_df = adni_input, 
                           XVars = c("Memory", "LMDR", "HippVol"),
                           YVars = "ME3",
                           adj_stratify_covs = c("LastDx_binary"),
                           all_status = c("Case", "Control"),
                           base_covs = c("Age", "Sex", "Educ"),
                           base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV", "Mag_HippICV"),
                           stratified = T)
adni_nduf_stratified <-  linear_fn_new(data_df = adni_input, 
                           XVars = c("Memory", "LMDR", "HippVol"),
                           YVars = nduf_bb$gene_name,
                           adj_stratify_covs = c("LastDx_binary"),
                           all_status = c("Case", "Control"),
                           base_covs = c("Age", "Sex", "Educ", "RIN", "AffyPlate.meanREE", "SITE.meanREE"),
                           base_covs_for_HV = c("Age", "Sex", "Educ", "Hipp_ICV",  "Mag_HippICV", "RIN", "AffyPlate.meanREE", "SITE.meanREE"),
                           stratified = T)
adni_stratified = rbind(adni_m3_stratified, adni_nduf_stratified)
rm(adni_m3_stratified, adni_nduf_stratified)

stratified_summ <- mcsa_stratified %>% 
  mutate(Cohort = "MCSA") %>% 
  dplyr::select(Cohort, X, Y, AdjStratify_Covar, Stratify_Status, SummDf) %>% 
  unnest(SummDf) %>% 
  rbind(
    adni_stratified %>% 
      mutate(Cohort = "ADNI") %>% 
      dplyr::select(Cohort, X, Y, AdjStratify_Covar, Stratify_Status, SummDf) %>% 
      unnest(SummDf)
  ) %>% 
  filter(X == term) %>% 
  group_by(Cohort, X) %>% 
  mutate(q.value = p.adjust(p.value, method = "fdr"), .after = "p.value") 
```




```{r meta-analysis for dx adjusted regression}
# prepare meta-analysis input dataframe
mcsa_dxadj_df <- dxadj_summ %>% filter(Cohort == "MCSA") %>% dplyr::select(-Cohort)
adni_dxadj_df <- dxadj_summ %>% filter(Cohort == "ADNI") %>% dplyr::select(-Cohort)

meta_df = inner_join(mcsa_dxadj_df, adni_dxadj_df, by = c("X","Y", "AdjStratify_Covar", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(rn = paste0(X, "_", Y, "_", AdjStratify_Covar)) %>% 
  column_to_rownames("rn")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "rn")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "rn") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "rn")

# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_dxadj = meta_df %>%
  group_by(X) %>%
  mutate(q_random = p.adjust(p_random, method = "fdr"),
         q_fixed = p.adjust(p_fixed, method = "fdr"),
         model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_meta = ifelse(model_meta =="Fixed", q_fixed, q_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random)
         ) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"), starts_with("Cohort_"))) %>% 
  dplyr::select(-rn, -term)

rm(meta_df, meta_raw, fix_res, rand_res, het_res, eff_idx1, eff_idx2, err_idx1, err_idx2)
```

```{r meta-analysis for dx stratified regression}
# prepare meta-analysis input dataframe
mcsa_stratified_df <- stratified_summ %>% filter(Cohort == "MCSA") %>% dplyr::select(-Cohort)
adni_stratified_df <- stratified_summ %>% filter(Cohort == "ADNI") %>% dplyr::select(-Cohort)

meta_df = inner_join(mcsa_stratified_df, adni_stratified_df, by = c("X","Y", "AdjStratify_Covar", "Stratify_Status", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(rn = paste0(X, "_", Y, "_", AdjStratify_Covar, "_", Stratify_Status)) %>% 
  column_to_rownames("rn")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "rn")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "rn") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "rn")

# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_stratified = meta_df %>%
  group_by(X, Stratify_Status) %>%
  mutate(q_random = p.adjust(p_random, method = "fdr"),
         q_fixed = p.adjust(p_fixed, method = "fdr"),
         model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_meta = ifelse(model_meta =="Fixed", q_fixed, q_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random)
         ) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"), starts_with("Cohort_"))) %>% 
  dplyr::select(-rn, -term)

rm(meta_df, meta_raw, fix_res, rand_res, het_res, eff_idx1, eff_idx2, err_idx1, err_idx2)
```


```{r save results - dx only}
save(mcsa_dxadj, mcsa_stratified, adni_dxadj, adni_stratified, file = paste0(outdir, "/2_1/DXAdj_Stratified_Regression.RData"))

write.xlsx(list(
  "DXAdj" = dxadj_summ,
  "Stratified" = stratified_summ,
  "DXAdj_Meta" = meta_dxadj,
  "Stratified_Meta" = meta_stratified
),
paste0(outdir, "/2_1/DXAdj_Stratified_Regression_summ.xlsx"))
```

# Plot regression results


```{r plot results}
# organize dataframe to plot the results adjusting for dx
plot_df <- meta_dxadj %>%
  dplyr::select(X:estimate_MCSA, p.value_MCSA:q.value_MCSA, estimate_ADNI, p.value_ADNI:q.value_ADNI, beta_meta, p_meta:q_meta) %>% 
  dplyr::rename(
    estimate_meta = beta_meta,
    p.value_meta = p_meta,
    q.value_meta = q_meta
    ) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(X=factor(X, levels=c("Memory", "LMDR", "HippVol")),
         Y = ifelse(Y == "ME3", "Bl_M3", Y),
         Cohort = ifelse(Cohort=="meta", "Meta-analyzed", Cohort),
         Cohort = factor(Cohort, levels=c("MCSA", "ADNI","Meta-analyzed")),
         lbl = if_else(q.value<0.05, "✱", ""), 
         lbl = if_else(p.value<0.05 & q.value>0.05,"Δ", lbl)
         )

# organize dataframe to plot the results stratified by dx
plot_df2 <- meta_stratified %>%
  dplyr::select(X:estimate_MCSA, p.value_MCSA:q.value_MCSA, estimate_ADNI, p.value_ADNI:q.value_ADNI, beta_meta, p_meta:q_meta) %>% 
  dplyr::rename(
    estimate_meta = beta_meta,
    p.value_meta = p_meta,
    q.value_meta = q_meta
    ) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  filter(Stratify_Status == "Control") %>% 
  mutate(X=factor(X, levels=c("Memory", "LMDR", "HippVol")),
         Y = ifelse(Y == "ME3", "Bl_M3", Y),
         Cohort = ifelse(Cohort=="meta", "Meta-analyzed", Cohort),
         Cohort = factor(Cohort, levels=c("MCSA", "ADNI","Meta-analyzed")),
         lbl = if_else(q.value<0.05, "✱", ""), 
         lbl = if_else(p.value<0.05 & q.value>0.05,"Δ", lbl)
         )

# organize a dataframe to plot the correlation between estimates 
## acquire estimates before adjusting for DX
metr <- read.csv(paste0(blwgcna_dir, "/module_trait.csv")) # this is created from 2_bl_wgcna.R
nduf_assoc <- read.csv(paste0(nduf_assoc_dir, "/Bl_Assoc.csv")) # this is created from 4_nduf_assoc_de.R

og_cs <- metr %>% 
  dplyr::select(Module, Pheno, estimate_MCSA, estimate_ADNI, beta_meta) %>% 
  filter(Module == "ME3" & Pheno %in% c("Memory", "LMDR", "HippVol")) %>% 
  rename(NDUF = Module, Phenotype = Pheno) %>% 
  rbind(
    nduf_assoc %>% 
      dplyr::select(gene_name, Phenotype, estimate_MCSA, estimate_ADNI, beta_meta) %>% 
      rename(NDUF = gene_name)
  ) %>% 
  dplyr::rename(estimate_meta = beta_meta) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(Cohort = ifelse(Cohort == "meta", "Meta-analyzed", Cohort),
         NDUF = ifelse(NDUF == "ME3", "Bl_M3", NDUF))
plot_df3 <- plot_df %>% 
  dplyr::select(X, Y, Cohort, estimate) %>% 
  dplyr::rename(est_DXAdjusted = estimate) %>% 
  left_join(
    og_cs %>% dplyr::rename(est_NoDXAdjusted = estimate),
    by= c("X" = "Phenotype", "Y" = "NDUF", "Cohort")
  ) %>% 
  mutate(Cohort = factor(Cohort, levels = c("MCSA", "ADNI", "Meta-analyzed")),
         X=factor(X, levels=c("Memory", "LMDR", "HippVol")))

cairo_pdf(paste0(outdir, "/2_1/DXAdj_Stratified_Regression.pdf"), width = 19.5, height = 5.5)
plot_df %>% 
  ggplot(aes(x=Y, y=Cohort)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Last clinical diagnosis included as a covariate",
    x = "",
    caption = "Δ:P<0.05; ✱: q<0.05") +
  facet_grid(X~ ., space = "free", scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "italic", angle = 30, hjust = 1),
        axis.title.y = element_blank(),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        text = element_text(family = "ArialMT"),
        plot.title = element_text(hjust = .5)) + # this is extended data fig.7b
plot_df3 %>%
  filter(Y != "Bl_M3") %>%
  ggplot(aes(x = est_NoDXAdjusted, y = est_DXAdjusted)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_grid(Cohort ~ X, scales = "free") +
  # Add correlation coefficient (r) and p-value
  stat_cor(
    method = "pearson",
    label.x.npc = "left",   # position (can be "left", "right", etc.)
    label.y.npc = "top",
    label.sep = "\n"
  ) +
  theme_bw() + # this is extended data fig.7c
plot_df2 %>% 
  filter(Stratify_Status == "Control") %>% 
  ggplot(aes(x=Y, y=Cohort)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Stratified by DX",
    subtitle = "Only show results with CN",
    x = "",
    caption = "Δ:P<0.05; ✱:q<0.05") +
  facet_grid(X ~. , space = "free", scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "italic", angle = 30, hjust = 1),
        axis.title.y = element_blank(),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        text = element_text(family = "ArialMT"),
        plot.title = element_text(hjust = .5),
        plot.subtitle = element_text(hjust = .5)) # this is extended data fig.7d
dev.off()
```



