---
title: "rbt_1.6"
output: html_document
date: "2026-08-03"
---

Author: Wei (Adelyn) Tsai; [tsai.wei\@mayo.edu](mailto:tsai.wei@mayo.edu){.email} If you are using this script, please cite our study. This script addresses comment 6 about prediction analysis opitimization from reviwer 1.

Please refer rebuttal/rbt_2_1_3_BF_TM.Rmd for definition of resilience/non-resilience

# Libraries and directories

```{r}
library(tidyverse) #v2.0.0
library(tidymodels) #v1.4.1
library(openxlsx) #v4.2.8
library(patchwork) #v1.3.2

tidymodels_prefer()  # resolve function conflicts in favour of tidymodels

set.seed(123)

# list of input/output directories
indir <- "./indir"
rbt_v1_dir <- "./Rebuttal/outdir" # directory to the results from the first rebuttal
blwgcna_dir <- "./outdir/bl_wgcna"
nduf_assoc_dir <- "./outdir/nduf_assoc_de"

# output directory
outdir <- "./Rebuttal_2nd/outdir/1_6"

source("./Codes/fns.R") #source a list of frequently used functions
```

# File prep
```{r}
comTrait <- readRDS(paste0(indir,"/comTrait.rds")) # resilience status were determined in rebuttal/rbt_2_1_3_BF_TM.Rmd

comRes <- readRDS(paste0(rbt_v1_dir, "/1_11/comRes_Combat.rds")) # use the combat adjusted residuals since we are training in one dataset and test in another dataset; from rebuttal/rbt_1_11.R
nduf_bb <- read.csv(paste0(nduf_assoc_dir, "/nduf_bb.csv")) #generated in 4_nduf_assoc_de.R

# select the residuals of NDUFs
nduf_res <- comRes[,nduf_bb$gene_id]
nduf_res <- data.frame(nduf_res)
# rename geneID to gene name of NDUFs
nduf_res  <- nduf_res  %>% rename_at(vars(1:(ncol(.))), ~ nduf_bb$gene_name[match(names(nduf_res)[1:(ncol(nduf_res))], nduf_bb$gene_id)])

# prepare input dataframes
mcsa_input <- comTrait[1:105,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Sex, Educ, Age, APOE4, starts_with("Res_")) %>% 
  left_join(nduf_res %>% rownames_to_column("SubjectID")) %>% 
  mutate(APOE4 = factor(APOE4)) %>% 
  mutate(across(
    starts_with("Res_X."),
    ~ factor(.x, levels = c("Resilient", "Non-resilient"))
  ))

adni_input <- comTrait[106:196,] %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Sex, Educ, Age, APOE4, starts_with("Res_")) %>% 
  left_join(nduf_res %>% rownames_to_column("SubjectID")) %>% 
  mutate(APOE4 = factor(APOE4)) %>% 
  mutate(across(
    starts_with("Res_X."),
    ~ factor(.x, levels = c("Resilient", "Non-resilient"))
  ))


```

# predictive modeling functions

Adopted from: <https://lsinks.github.io/posts/2023-04-10-tidymodels/tidymodels_tutorial.html#defining-the-model-with-parsnip>

```{r classification}

modeling <- function(
    train_data,
    test_data,
    outcome_col,
    covar_col,
    gene_col,
    mode = c("separate", "all") # test all or separate genes at a time
){
  mode <- match.arg(mode)
  
  # remove any NA and select only needed columns
  train_data <- train_data %>% 
    drop_na(outcome_col, covar_col, gene_col) %>% 
    dplyr::select(all_of(c(outcome_col, covar_col, gene_col))) 
  
  test_data <- test_data %>% 
    drop_na(outcome_col, covar_col, gene_col) %>% 
    dplyr::select(all_of(c(outcome_col, covar_col, gene_col))) 
  
  # create data_split 
  data_split <- make_splits(
    x = train_data,   
    assessment = test_data
  )
  
  # set up dataframe for metrics and coefficients ----
  metrics <- data.frame()
  coefs   <- data.frame()
  preds   <- data.frame()
  auc_comp<- data.frame()
  
  # Define ----
  ## 1) models ----
  ### standard logistic regression
  logreg_spec <- 
    logistic_reg() %>% 
    set_engine("glm") %>% 
    set_mode("classification")
  
  ### ridge regularization
  ridge_spec <- 
    logistic_reg(penalty = tune(),
                 mixture = 0) %>%
    set_engine("glmnet") %>% 
    set_mode("classification")
  
  ### elastic net regularization
  en_spec <- 
    logistic_reg(penalty = tune(),
                 mixture = tune()) %>%
    set_engine("glmnet") %>% 
    set_mode("classification")
  
  ## 2) metrics to evaluate ----
  custom_metric_set = metric_set(roc_auc, sensitivity, specificity, brier_class)
  
  ## 3) formula ----
  formula_cov = as.formula(paste0(outcome_col, " ~ Sex + Educ + Age + APOE4"))
  formula_full = as.formula(paste0(outcome_col, " ~ ", paste0(c(covar_col, gene_col), collapse = " + ")))
  formula_gene = as.formula(paste0(outcome_col, " ~ ", paste0(gene_col, collapse = " + ")))
  
  ## 4) Create recipe ----
  recipe_cov <- recipe(formula_cov, data = train_data) %>% 
    step_dummy(all_nominal_predictors()) %>% 
    step_nzv(all_predictors()) %>% 
    step_normalize(all_numeric_predictors())
  recipe_full <- recipe(formula_full, data = train_data) %>% 
    step_dummy(all_nominal_predictors()) %>% 
    step_nzv(all_predictors()) %>% 
    step_normalize(all_numeric_predictors())
  recipe_gene <- recipe(formula_gene, data = train_data) %>% 
    step_dummy(all_nominal_predictors()) %>% 
    step_nzv(all_predictors()) %>% 
    step_normalize(all_numeric_predictors())
  
  ## Helper functions ----
  ### define metrics calculation function
  cal_metrics <- function(data, prob_col, class_col){
    
    metric_data <- tibble(
      truth = data$truth,
      pred_class = data[[class_col]],
      probability = data[[prob_col]]
    )
    
    # yardstick metrics
    yardstick_metrics <- custom_metric_set(
      metric_data,
      truth = truth,
      estimate = pred_class,
      probability
    ) %>%
      dplyr::select(.metric, .estimate)
    
    # calibration
    y_num   <- as.numeric(metric_data$truth == levels(metric_data$truth)[1])
    cal <- rms::val.prob(metric_data$probability, y_num, pl = F)
    calibration <- 
      tibble(".metric" = c("calibration_intercept", "calibration_slope"),
             ".estimate" = c(cal[["Intercept"]], cal[["Slope"]])
      )
    
    bind_rows(
      yardstick_metrics,
      calibration
    )
  }
  
  
  ### Perform AUC/ROC test
  roc_sig_test <- function(dat, seed = 1){
    roc_cov <- pROC::roc(dat$truth, dat$.pred_prob_cov, levels = rev(levels(dat$truth)), direction = "<")
    roc_full <- pROC::roc(dat$truth, dat$.pred_prob_full, levels = rev(levels(dat$truth)), direction = "<")
    
    set.seed(1)
    venkatram <- pROC::roc.test(roc_cov, roc_full,
                                method = "venkatraman",   # permutation test
                                paired = TRUE,            # correct here — same subjects, two prediction sets
                                boot.n = 2000              # number of permutations
    )
    delong <- pROC::roc.test(roc_cov, roc_full,
                             method = "delong",   # permutation test
                             paired = TRUE            # correct here — same subjects, two prediction sets
    )
    auc_comp <- tibble("Venkatram" = venkatram$p.value, "DeLong" = delong$p.value)
  }
  
  ## fit the model (mode = separate) ----
  if (mode == "separate"){
    
    ## create workflow
    wf_cov <- workflow() %>% 
      add_recipe(recipe_cov) %>% 
      add_model(logreg_spec)
    wf_full <- workflow() %>% 
      add_recipe(recipe_full) %>% 
      add_model(logreg_spec)
    wf_gene <- workflow() %>% 
      add_recipe(recipe_gene) %>% 
      add_model(logreg_spec)
    
    ## fitting
    last_fit_cov <- last_fit(wf_cov, data_split, metrics = custom_metric_set)
    last_fit_full <- last_fit(wf_full, data_split, metrics = custom_metric_set)
    last_fit_gene <- last_fit(wf_gene, data_split, metrics = custom_metric_set)
    
    ## collect coefficients (establish from train_data)
    coefs <- rbind(
      coefs,
      last_fit_cov %>% extract_fit_parsnip() %>% tidy() %>% mutate(covariate_only = TRUE),
      last_fit_full %>% extract_fit_parsnip() %>% tidy() %>% mutate(covariate_only = FALSE),
      last_fit_gene %>% extract_fit_parsnip() %>% tidy() %>% mutate(covariate_only = FALSE)
    )
    
    ## collect predictions (prediction on test_data)
    pred_cur_cov <- last_fit_cov %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_cov = .pred_Resilient, .pred_class_cov = .pred_class, truth = {{outcome_col}})
    pred_cur_full <- last_fit_full %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_full = .pred_Resilient, .pred_class_full = .pred_class, truth_full = {{outcome_col}})
    pred_cur_gene <- last_fit_gene %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_gene = .pred_Resilient, .pred_class_gene = .pred_class, truth_gene = {{outcome_col}})
    
    pred_cur <- pred_cur_cov %>% left_join(pred_cur_full, by = ".row") %>% left_join(pred_cur_gene, by = ".row") %>% mutate(model_type = "logreg")
    
    # Check that truth is identical
    if (!all(
      pred_cur$truth == pred_cur$truth_full & pred_cur$truth_full == pred_cur$truth_gene
    )) {
      stop(
        "The three last_fit objects do not appear to contain
       predictions for the same observations in the same order."
      )
    }
    
    preds <- rbind(
      preds,
      pred_cur
    ) %>% 
      dplyr::select(-truth_full, -truth_gene) %>% 
      relocate(.row, .before = "truth")
    
    ## collect metrics (evaluated on test_data)
    ### get the point estimates
    point_cov <- cal_metrics(pred_cur, prob_col = ".pred_prob_cov", class_col = ".pred_class_cov") %>% rename(.estimate_cov = .estimate)
    point_full <- cal_metrics(pred_cur, prob_col = ".pred_prob_full", class_col = ".pred_class_full") %>% rename(.estimate_full = .estimate)
    point_gene <- cal_metrics(pred_cur, prob_col = ".pred_prob_gene", class_col = ".pred_class_gene") %>% rename(.estimate_gene = .estimate)
    point <- point_cov %>% left_join(point_full, by = ".metric") %>% left_join(point_gene, by = ".metric") %>% mutate(model_type = "logreg")
    
    metrics <- rbind(
      metrics,
      point
    ) %>% 
      mutate(delta_metric = .estimate_full - .estimate_cov, .after = ".estimate_full")
    
    ## perform ROC AUC test
    auc_comp_cur = roc_sig_test(dat = pred_cur) %>% mutate(model_type = "logreg")
    
    auc_comp <- rbind(
      auc_comp,
      auc_comp_cur
    )
    
  }
  
  ## fit the model (mode = all) ----
  else if (mode == "all"){
    ## create workflow
    wf_tune_cov <- 
      workflow_set(
        list(data = recipe_cov),
        list(logreg = logreg_spec, ridge = ridge_spec, en = en_spec)
      )
    wf_tune_full <- 
      workflow_set(
        list(data = recipe_full),
        list(logreg = logreg_spec, ridge = ridge_spec, en = en_spec)
      )
    wf_tune_gene <- 
      workflow_set(
        list(data = recipe_gene),
        list(logreg = logreg_spec, ridge = ridge_spec, en = en_spec)
      )
    
    ## tuning
    set.seed(1234)
    folds <- vfold_cv(train_data, v = 10, strata = outcome_col)
    tune_results_cov <- 
      workflow_map(
        wf_tune_cov,
        "tune_grid",
        resamples = folds,
        grid = 10,
        metrics = custom_metric_set,
        verbose = F
      )
    tune_results_full <- 
      workflow_map(
        wf_tune_full,
        "tune_grid",
        resamples = folds,
        grid = 10,
        metrics = custom_metric_set,
        verbose = F
      )
    tune_results_gene <- 
      workflow_map(
        wf_tune_gene,
        "tune_grid",
        resamples = folds,
        grid = 10,
        metrics = custom_metric_set,
        verbose = F
      )
    
    ## get best parameters (for ridge and en) and fitting 
    
    ### logreg
    validation_logreg_cov <- tune_results_cov %>% extract_workflow("data_logreg") %>% last_fit(split =  data_split, metrics = custom_metric_set)
    validation_logreg_full <- tune_results_full %>% extract_workflow("data_logreg") %>% last_fit(split =  data_split, metrics = custom_metric_set)
    validation_logreg_gene <- tune_results_gene %>% extract_workflow("data_logreg") %>% last_fit(split =  data_split, metrics = custom_metric_set)
    
    ### ridge
    best_ridge_cov <- tune_results_cov %>% extract_workflow_set_result("data_ridge") %>% select_best(metric = "roc_auc")
    validation_ridge_cov <- tune_results_cov %>% extract_workflow("data_ridge") %>% finalize_workflow(best_ridge_cov) %>% last_fit(split =  data_split, metrics = custom_metric_set)
    
    best_ridge_full <- tune_results_full %>% extract_workflow_set_result("data_ridge") %>% select_best(metric = "roc_auc")
    validation_ridge_full <- tune_results_full %>% extract_workflow("data_ridge") %>% finalize_workflow(best_ridge_full) %>% last_fit(split =  data_split, metrics = custom_metric_set)
    
    best_ridge_gene <- tune_results_gene %>% extract_workflow_set_result("data_ridge") %>% select_best(metric = "roc_auc")
    validation_ridge_gene <- tune_results_gene %>% extract_workflow("data_ridge") %>% finalize_workflow(best_ridge_gene) %>% last_fit(split =  data_split, metrics = custom_metric_set)
    
    ### EN
    best_en_cov <- tune_results_cov %>% extract_workflow_set_result("data_en") %>% select_best(metric = "roc_auc")
    validation_en_cov <- tune_results_cov %>% extract_workflow("data_en") %>% finalize_workflow(best_en_cov) %>% last_fit(split =  data_split, metrics = custom_metric_set)
    
    best_en_full <- tune_results_full %>% extract_workflow_set_result("data_en") %>% select_best(metric = "roc_auc")
    validation_en_full <- tune_results_full %>% extract_workflow("data_en") %>% finalize_workflow(best_en_full) %>% last_fit(split =  data_split, metrics = custom_metric_set)
    
    best_en_gene <- tune_results_gene %>% extract_workflow_set_result("data_en") %>% select_best(metric = "roc_auc")
    validation_en_gene <- tune_results_gene %>% extract_workflow("data_en") %>% finalize_workflow(best_en_gene) %>% last_fit(split =  data_split, metrics = custom_metric_set)
    
    ## collect coefficients
    coefs <- rbind(
      coefs,
      validation_logreg_cov %>% extract_fit_parsnip() %>% tidy() %>% dplyr::select(term, estimate) %>% mutate(penalty = NA, model_type = "logreg", covariate_only = TRUE),
      validation_logreg_full %>% extract_fit_parsnip() %>% tidy() %>% dplyr::select(term, estimate) %>% mutate(penalty = NA, model_type = "logreg", covariate_only = FALSE),
      validation_logreg_gene %>% extract_fit_parsnip() %>% tidy() %>% dplyr::select(term, estimate) %>% mutate(penalty = NA, model_type = "logreg", covariate_only = FALSE),
      validation_ridge_cov %>% extract_fit_parsnip() %>% tidy() %>% mutate(model_type = "ridge", covariate_only = TRUE),
      validation_ridge_full %>% extract_fit_parsnip() %>% tidy() %>% mutate(model_type = "ridge", covariate_only = FALSE),
      validation_ridge_gene %>% extract_fit_parsnip() %>% tidy() %>% mutate(model_type = "ridge", covariate_only = FALSE),
      validation_en_cov %>% extract_fit_parsnip() %>% tidy() %>% mutate(model_type = "en", covariate_only = TRUE),
      validation_en_full %>% extract_fit_parsnip() %>% tidy() %>% mutate(model_type = "en", covariate_only = TRUE),
      validation_en_gene %>% extract_fit_parsnip() %>% tidy() %>% mutate(model_type = "en", covariate_only = FALSE)
    )
    
    ## collect predictions
    pred_cur_logreg_cov <- validation_logreg_cov %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_cov = .pred_Resilient, .pred_class_cov = .pred_class, truth = {{outcome_col}})
    pred_cur_logreg_full <- validation_logreg_full %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_full = .pred_Resilient, .pred_class_full = .pred_class, truth_full = {{outcome_col}})
    pred_cur_logreg_gene <- validation_logreg_gene %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_gene = .pred_Resilient, .pred_class_gene = .pred_class, truth_gene = {{outcome_col}})
    
    pred_cur_ridge_cov <- validation_ridge_cov %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_cov = .pred_Resilient,  .pred_class_cov = .pred_class, truth = {{outcome_col}})
    pred_cur_ridge_full <- validation_ridge_full %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_full = .pred_Resilient, .pred_class_full = .pred_class, truth_full = {{outcome_col}})
    pred_cur_ridge_gene <- validation_ridge_gene %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_gene = .pred_Resilient, .pred_class_gene = .pred_class, truth_gene = {{outcome_col}})
    
    pred_cur_en_cov <- validation_en_cov %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_cov = .pred_Resilient,  .pred_class_cov = .pred_class, truth = {{outcome_col}})
    pred_cur_en_full <- validation_en_full %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_full = .pred_Resilient, .pred_class_full = .pred_class, truth_full = {{outcome_col}})
    pred_cur_en_gene <- validation_en_gene %>% collect_predictions() %>% dplyr::select({{outcome_col}}, .pred_class, .pred_Resilient, .row) %>% rename(.pred_prob_gene = .pred_Resilient, .pred_class_gene = .pred_class, truth_gene = {{outcome_col}})
    
    pred_logreg <- pred_cur_logreg_cov %>% left_join(pred_cur_logreg_full, by = ".row") %>% left_join(pred_cur_logreg_gene, by = ".row") %>%  mutate(model_type = "logreg")
    pred_ridge <- pred_cur_ridge_cov %>% left_join(pred_cur_ridge_full, by = ".row") %>% left_join(pred_cur_ridge_gene, by = ".row") %>%  mutate(model_type = "ridge")
    pred_en <- pred_cur_en_cov %>% left_join(pred_cur_en_full, by = ".row") %>% left_join(pred_cur_en_gene, by = ".row") %>% mutate(model_type = "en")
    
    # Check that truth is identical
    if (!all(
      pred_logreg$truth == pred_logreg$truth_full & pred_ridge$truth == pred_ridge$truth_full & pred_en$truth == pred_en$truth_full &
      pred_logreg$truth == pred_logreg$truth_gene & pred_ridge$truth == pred_ridge$truth_gene & pred_en$truth == pred_en$truth_gene
    )) {
      stop(
        "The three last_fit objects do not appear to contain
       predictions for the same observations in the same order."
      )
    }
    
    preds <- rbind(
      preds,
      pred_logreg,
      pred_ridge,
      pred_en
    ) %>% 
      dplyr::select(-truth_full, -truth_gene) %>% 
       relocate(.row, .before = "truth")
    
    
    ## collect metrics
    ### get the point estimates
    point_logreg_cov <- cal_metrics(pred_logreg, prob_col = ".pred_prob_cov", class_col = ".pred_class_cov") %>% rename(.estimate_cov = .estimate)
    point_logreg_full <- cal_metrics(pred_logreg, prob_col = ".pred_prob_full", class_col = ".pred_class_full") %>% rename(.estimate_full = .estimate)
    point_logreg_gene <- cal_metrics(pred_logreg, prob_col = ".pred_prob_gene", class_col = ".pred_class_gene") %>% rename(.estimate_gene = .estimate)
    point_logreg <- point_logreg_cov %>% left_join(point_logreg_full, by = ".metric") %>% left_join(point_logreg_gene, by = ".metric") %>% mutate(model_type = "logreg")
    
    point_ridge_cov <- cal_metrics(pred_ridge, prob_col = ".pred_prob_cov", class_col = ".pred_class_cov") %>% rename(.estimate_cov = .estimate)
    point_ridge_full <- cal_metrics(pred_ridge, prob_col = ".pred_prob_full", class_col = ".pred_class_full") %>% rename(.estimate_full = .estimate)
    point_ridge_gene <- cal_metrics(pred_ridge, prob_col = ".pred_prob_gene", class_col = ".pred_class_gene") %>% rename(.estimate_gene = .estimate)
    point_ridge <- point_ridge_cov %>% left_join(point_ridge_full, by = ".metric") %>% left_join(point_ridge_gene, by = ".metric") %>% mutate(model_type = "ridge")
    
    point_en_cov <- cal_metrics(pred_en, prob_col = ".pred_prob_cov", class_col = ".pred_class_cov") %>% rename(.estimate_cov = .estimate)
    point_en_full <- cal_metrics(pred_en, prob_col = ".pred_prob_full", class_col = ".pred_class_full") %>% rename(.estimate_full = .estimate)
    point_en_gene <- cal_metrics(pred_en, prob_col = ".pred_prob_gene", class_col = ".pred_class_gene") %>% rename(.estimate_gene = .estimate)
    point_en <- point_en_cov %>% left_join(point_en_full, by = ".metric") %>% left_join(point_en_gene, by = ".metric") %>% mutate(model_type = "en")
    
    
    metrics <- rbind(
      metrics,
      point_logreg,
      point_ridge,
      point_en
    ) %>% 
      mutate(delta_metric = .estimate_full - .estimate_cov, .after = ".estimate_full")
    
    ## perform ROC AUC test
    auc_comp_logreg = roc_sig_test(dat = pred_logreg) %>% mutate(model_type = "logreg")
    auc_comp_ridge = roc_sig_test(dat = pred_ridge) %>% mutate(model_type = "ridge")
    auc_comp_en = roc_sig_test(dat = pred_en) %>% mutate(model_type = "en")
   
    auc_comp <- rbind(
      auc_comp,
      auc_comp_logreg,
      auc_comp_ridge,
      auc_comp_en
    )
  }
  
  # Final storage 
  list(
    metrics_all = metrics,
    preds_all   = preds,
    coef_all    = coefs,
    auc_comp    = auc_comp
  )
}

run_modeling <- function(
  train_data,
  test_data,
  outcome_cols, 
  gene_vars,
  covars,
  mode = c("separate", "all")
){
  mode <- match.arg(mode)
  
  # store all results by outcome
  results <- list()
  
  for (outcome in outcome_cols) {
    print(paste0("OUTCOME: ", outcome))
    
    results[[outcome]] <- list()
    
    ## ----------------------------------------------------------
    ## MODE 1: EACH GENE SEPARATELY
    ## ----------------------------------------------------------
    if (mode == "separate") {
      
      separate_list <- list()
      
      for (g in gene_vars) {

        print(paste0("Running gene: ", g))
        
        separate_list[[g]] <-
          modeling(
            train_data    = train_data,
            test_data     = test_data,
            outcome_col   = outcome,
            covar_col     = covars,
            gene_col      = g,
            mode          = mode
          )
      }
      
      results[[outcome]] <- separate_list
    }
    
    ## ----------------------------------------------------------
    ## MODE 2: ALL GENES AT ONCE
    ## ----------------------------------------------------------
    if (mode == "all") {

      print("Running: all genes together")
      
      results[[outcome]] <-
        modeling(
          train_data    = train_data,
          test_data     = test_data,
          outcome_col   = outcome,
          covar_col     = covars,
          gene_col      = gene_vars,
          mode          = mode
        )
    }
  }
  
  return(results)
}


```


| Column name      | Residual method to define          |
|------------------|------------------------------------|
| Res_X.Age_Y.Mem  | No centiloids, memory as outcome   |
| Res_X.Age_Y.LMDR | No centiloids, LMDR as outcome     |
| Res_X.CL_Y.Mem   | With centiloids, memory as outcome |
| Res_X.CL_Y.LMDR  | With centiloids, LMDR as outcome   |


# run modeling
```{r}
start_time = Sys.time()
m2a_sep <- run_modeling(
  train_data = mcsa_input,
  test_data = adni_input,
  outcome_cols = c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
  covars = c("Sex", "Educ", "Age", "APOE4"),
  gene_vars = nduf_bb$gene_name,
  mode = "separate"
)

m2a_all <- run_modeling(
  train_data = mcsa_input,
  test_data = adni_input,
  outcome_cols = c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
  covars = c("Sex", "Educ", "Age", "APOE4"),
  gene_vars = nduf_bb$gene_name,
  mode = "all"
)

a2m_sep <- run_modeling(
  train_data = adni_input,
  test_data = mcsa_input,
  outcome_cols = c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
  covars = c("Sex", "Educ", "Age", "APOE4"),
  gene_vars = nduf_bb$gene_name,
  mode = "separate"
)

a2m_all <- run_modeling(
  train_data = adni_input,
  test_data = mcsa_input,
  outcome_cols = c("Res_X.Age_Y.Mem", "Res_X.Age_Y.LMDR", "Res_X.CL_Y.Mem", "Res_X.CL_Y.LMDR"),
  covars = c("Sex", "Educ", "Age", "APOE4"),
  gene_vars = nduf_bb$gene_name,
  mode = "all"
)
save(m2a_sep, m2a_all, a2m_sep, a2m_all, file = paste0(outdir, "/results.RData"))
end_time = Sys.time()
message("Spent: ", end_time - start_time)
```

# organize the results
## MSCA to ADNI
```{r separate mode}
m2a_sep_metrics <- map_df(
  names(m2a_sep),
  function(outcome) {
    map_df(
      names(m2a_sep[[outcome]]),
      function(x) {
        m2a_sep[[outcome]][[x]]$metrics_all %>%
          mutate(Outcome = outcome, Gene = x)
      }
    )
  }
) %>% 
  arrange(Outcome, Gene) %>% 
  relocate(c(Outcome, Gene), .before = ".metric")
  
m2a_sep_auc_comp <- map_df(
  names(m2a_sep),
  function(outcome) {
    map_df(
      names(m2a_sep[[outcome]]),
      function(x) {
        m2a_sep[[outcome]][[x]]$auc_comp %>%
          mutate(Outcome = outcome, Gene = x)
      }
    )
  }
)
# process to get a dataframe focusing on AUC
m2a_auc <- m2a_sep_metrics %>% 
  filter(.metric == "roc_auc") %>% 
  left_join(m2a_sep_auc_comp, by = c("Outcome", "Gene", "model_type"))

rm(m2a_sep_auc_comp)

m2a_auc  
```

```{r all mode}
m2a_all_metrics <- map_df(
  names(m2a_all),
  function(outcome) {
      m2a_all[[outcome]]$metrics_all %>%
          mutate(Outcome = outcome)
  }
) %>% 
  mutate(Gene = "All 20 NDUFs") %>%  
  relocate(c(Outcome, Gene, model_type), .before = ".metric")

# combine the metrics dataframe
m2a_metrics <- m2a_sep_metrics %>% rbind(m2a_all_metrics)
  
m2a_all_auc_comp <- map_df(
  names(m2a_all),
  function(outcome) {
    m2a_all[[outcome]]$auc_comp %>%
          mutate(Outcome = outcome)
  }
)
# process to get a dataframe focusing on AUC
m2a_auc <- m2a_auc %>% 
  rbind(
    m2a_all_metrics %>% 
      filter(.metric == "roc_auc") %>% 
      left_join(m2a_all_auc_comp, by = c("Outcome", "model_type"))
    )

rm(m2a_sep_metrics, m2a_all_metrics, m2a_all_auc_comp)

m2a_metrics
m2a_auc
```


## ADNI to MCSA
```{r separate mode}
a2m_sep_metrics <- map_df(
  names(a2m_sep),
  function(outcome) {
    map_df(
      names(a2m_sep[[outcome]]),
      function(x) {
        a2m_sep[[outcome]][[x]]$metrics_all %>%
          mutate(Outcome = outcome, Gene = x)
      }
    )
  }
) %>% 
  arrange(Outcome, Gene) %>%
  relocate(c(Outcome, Gene), .before = ".metric")
  
a2m_sep_auc_comp <- map_df(
  names(a2m_sep),
  function(outcome) {
    map_df(
      names(a2m_sep[[outcome]]),
      function(x) {
        a2m_sep[[outcome]][[x]]$auc_comp %>%
          mutate(Outcome = outcome, Gene = x)
      }
    )
  }
)
# process to get a dataframe focusing on AUC
a2m_auc <- a2m_sep_metrics %>% 
  filter(.metric == "roc_auc") %>% 
  left_join(a2m_sep_auc_comp, by = c("Outcome", "Gene", "model_type"))

rm(a2m_sep_auc_comp)

a2m_auc
```

```{r all mode}
a2m_all_metrics <- map_df(
  names(a2m_all),
  function(outcome) {
      a2m_all[[outcome]]$metrics_all %>%
          mutate(Outcome = outcome)
  }
) %>% 
  mutate(Gene = "All 20 NDUFs") %>%  
  relocate(c(Outcome, Gene, model_type), .before = ".metric")

# combine the metrics dataframe
a2m_metrics <- a2m_sep_metrics %>% rbind(a2m_all_metrics)
  
a2m_all_auc_comp <- map_df(
  names(a2m_all),
  function(outcome) {
    a2m_all[[outcome]]$auc_comp %>%
          mutate(Outcome = outcome)
  }
)
# process to get a dataframe focusing on AUC
a2m_auc <- a2m_auc %>% 
  rbind(
    a2m_all_metrics %>% 
      filter(.metric == "roc_auc") %>% 
      left_join(a2m_all_auc_comp, by = c("Outcome", "model_type"))
    )

rm(a2m_sep_metrics, a2m_all_metrics, a2m_all_auc_comp)

a2m_metrics
a2m_auc
```

## save the dataframe
```{r}
m2a_metrics = m2a_metrics %>% 
  mutate(Outcome = case_when(
    Outcome == "Res_X.Age_Y.Mem" ~ "Memory (no CL)",
    Outcome == "Res_X.Age_Y.LMDR" ~ "LMDR (no CL)",
    Outcome == "Res_X.CL_Y.Mem" ~ "Memory (with CL)",
    Outcome == "Res_X.CL_Y.LMDR" ~ "LMDR (with CL)"
  ))
m2a_auc = m2a_auc %>% 
  mutate(Outcome = case_when(
    Outcome == "Res_X.Age_Y.Mem" ~ "Memory (no CL)",
    Outcome == "Res_X.Age_Y.LMDR" ~ "LMDR (no CL)",
    Outcome == "Res_X.CL_Y.Mem" ~ "Memory (with CL)",
    Outcome == "Res_X.CL_Y.LMDR" ~ "LMDR (with CL)"
  ))
a2m_metrics = a2m_metrics %>% 
  mutate(Outcome = case_when(
    Outcome == "Res_X.Age_Y.Mem" ~ "Memory (no CL)",
    Outcome == "Res_X.Age_Y.LMDR" ~ "LMDR (no CL)",
    Outcome == "Res_X.CL_Y.Mem" ~ "Memory (with CL)",
    Outcome == "Res_X.CL_Y.LMDR" ~ "LMDR (with CL)"
  ))
a2m_auc = a2m_auc %>% 
  mutate(Outcome = case_when(
    Outcome == "Res_X.Age_Y.Mem" ~ "Memory (no CL)",
    Outcome == "Res_X.Age_Y.LMDR" ~ "LMDR (no CL)",
    Outcome == "Res_X.CL_Y.Mem" ~ "Memory (with CL)",
    Outcome == "Res_X.CL_Y.LMDR" ~ "LMDR (with CL)"
  ))

write.xlsx(list(
  "Metrics_Train-MCSA_Test-ADNI" = m2a_metrics,
  "AUC_Train-MCSA_Test-ADNI" = m2a_auc,
  "Metrics_Train-ADNI_Test-ADNI" = a2m_metrics,
  "AUC_Train-ADNI_Test-MCSA" = a2m_auc
  ),
  paste0(outdir, "/metrics_auc.xlsx")
) # these tables are organized into Supplementary Table 14b
```

# Plot
## Distribution of CR scores
This is in Extended Data Fig.12d
```{r}
CRdf_mcsa = read.xlsx(paste0(rbt_v1_dir, "/2_1_3_BF/CRdf_2yrs.xlsx"), sheet = "MCSA") # from rebuttal/rbt_2_1_3_BF_TM.Rmd
CRdf_adni = read.xlsx(paste0(rbt_v1_dir, "/2_1_3_BF/CRdf_2yrs.xlsx"), sheet = "ADNI")

pdf(paste0(outdir, "/residuals_distribution.pdf"), width = 5.5, height = 3.5)
CRdf_mcsa %>% 
  dplyr::select(SubjectID, starts_with("CR_X.")) %>% 
  mutate(Cohort = "MCSA") %>% 
  rbind(CRdf_adni %>% dplyr::select(SubjectID, starts_with("CR_X.")) %>% mutate(Cohort = "ADNI")) %>% 
  pivot_longer(starts_with("CR_X."), names_to = "Outcome", values_to = "CR") %>% 
  mutate(Cohort = factor(Cohort, levels = c("MCSA", "ADNI"))) %>% 
  mutate(Outcome = case_when(
    Outcome == "CR_X.Age_Y.Mem" ~ "Memory (no CL)",
    Outcome == "CR_X.Age_Y.LMDR" ~ "LMDR (no CL)",
    Outcome == "CR_X.CL_Y.Mem" ~ "Memory (with CL)",
    Outcome == "CR_X.CL_Y.LMDR" ~ "LMDR (with CL)"
  )) %>% 
  ggplot(aes(x = CR)) +
  geom_histogram(bins = 8, color = "black", aes(fill = Cohort)) +
  facet_grid(Cohort ~ Outcome, scales = "free") +
  labs(x = "Residuals") +
  theme_bw() +
  theme(legend.position = "none")
dev.off()
```

## ROC curves
Plot for NDUFA9, NDUFV1, NDUFB8.
These are in Fig.3c and Extended Data Fig.12e
```{r LMDR no CL}
# Extract ROC data for plotting
roc_m2a_4p <- m2a_sep$Res_X.Age_Y.LMDR$NDUFA9$preds_all %>%
  roc_curve(truth = truth, .pred_prob_gene) %>%
  mutate(Gene = "NDUFA9", 
         AUC = subset(m2a_sep$Res_X.Age_Y.LMDR$NDUFA9$metrics_all, .metric == "roc_auc")$.estimate_gene,
         lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")")) %>% 
  rbind(
    m2a_sep$Res_X.Age_Y.LMDR$NDUFV1$preds_all %>%
      roc_curve(truth = truth, .pred_prob_gene) %>%
      mutate(Gene = "NDUFV1", 
             AUC = subset(m2a_sep$Res_X.Age_Y.LMDR$NDUFV1$metrics_all, .metric == "roc_auc")$.estimate_gene,
             lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")"))
    ) %>% 
  rbind(
    m2a_sep$Res_X.Age_Y.LMDR$NDUFB8$preds_all %>%
      roc_curve(truth = truth, .pred_prob_gene) %>%
      mutate(Gene = "NDUFB8", 
             AUC = subset(m2a_sep$Res_X.Age_Y.LMDR$NDUFB8$metrics_all, .metric == "roc_auc")$.estimate_gene,
             lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")"))
  )
roc_a2m_4p <- a2m_sep$Res_X.Age_Y.LMDR$NDUFA9$preds_all %>%
  roc_curve(truth = truth, .pred_prob_gene) %>%
  mutate(Gene = "NDUFA9", 
         AUC = subset(a2m_sep$Res_X.Age_Y.LMDR$NDUFA9$metrics_all, .metric == "roc_auc")$.estimate_gene,
         lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")")) %>% 
  rbind(
    a2m_sep$Res_X.Age_Y.LMDR$NDUFV1$preds_all %>%
      roc_curve(truth = truth, .pred_prob_gene) %>%
      mutate(Gene = "NDUFV1", 
             AUC = subset(a2m_sep$Res_X.Age_Y.LMDR$NDUFV1$metrics_all, .metric == "roc_auc")$.estimate_gene,
             lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")"))
  ) %>% 
  rbind(
    a2m_sep$Res_X.Age_Y.LMDR$NDUFB8$preds_all %>%
      roc_curve(truth = truth, .pred_prob_gene) %>%
      mutate(Gene = "NDUFB8", 
             AUC = subset(a2m_sep$Res_X.Age_Y.LMDR$NDUFB8$metrics_all, .metric == "roc_auc")$.estimate_gene,
             lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")"))
  )


p1 <- roc_m2a_4p %>% 
  mutate(Gene = factor(Gene, levels = c("NDUFA9", "NDUFV1", "NDUFB8"))) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, color=Gene)) +
  geom_path() +
  geom_abline(lty = 3) +
  scale_color_manual(
    labels = unique(roc_m2a_4p$lbl), 
    values = c("tomato","limegreen", "turquoise"), 
    name="Gene") +
  coord_equal() +
  labs(title="Train: MCSA --> Test: ADNI") +
  theme_bw(base_size = 7)
p2 <- roc_a2m_4p %>% 
  mutate(Gene = factor(Gene, levels = c("NDUFA9", "NDUFV1", "NDUFB8"))) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, color=Gene)) +
  geom_path() +
  geom_abline(lty = 3) +
  scale_color_manual(
    labels = unique(roc_a2m_4p$lbl), 
    values = c("tomato","limegreen", "turquoise"), 
    name="Gene") +
  coord_equal() +
  labs(title="Train: ADNI --> Test: MCSA") +
  theme_bw(base_size = 7)
```

```{r LMDR with CL}
# Extract ROC data for plotting
roc_m2a_4p_CL <- m2a_sep$Res_X.CL_Y.LMDR$NDUFA9$preds_all %>%
  roc_curve(truth = truth, .pred_prob_gene) %>%
  mutate(Gene = "NDUFA9", 
         AUC = subset(m2a_sep$Res_X.CL_Y.LMDR$NDUFA9$metrics_all, .metric == "roc_auc")$.estimate_gene,
         lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")")) %>% 
  rbind(
    m2a_sep$Res_X.CL_Y.LMDR$NDUFV1$preds_all %>%
      roc_curve(truth = truth, .pred_prob_gene) %>%
      mutate(Gene = "NDUFV1", 
             AUC = subset(m2a_sep$Res_X.CL_Y.LMDR$NDUFV1$metrics_all, .metric == "roc_auc")$.estimate_gene,
             lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")"))
    ) %>% 
  rbind(
    m2a_sep$Res_X.CL_Y.LMDR$NDUFB8$preds_all %>%
      roc_curve(truth = truth, .pred_prob_gene) %>%
      mutate(Gene = "NDUFB8", 
             AUC = subset(m2a_sep$Res_X.CL_Y.LMDR$NDUFB8$metrics_all, .metric == "roc_auc")$.estimate_gene,
             lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")"))
  )
roc_a2m_4p_CL <- a2m_sep$Res_X.CL_Y.LMDR$NDUFA9$preds_all %>%
  roc_curve(truth = truth, .pred_prob_gene) %>%
  mutate(Gene = "NDUFA9", 
         AUC = subset(a2m_sep$Res_X.CL_Y.LMDR$NDUFA9$metrics_all, .metric == "roc_auc")$.estimate_gene,
         lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")")) %>% 
  rbind(
    a2m_sep$Res_X.CL_Y.LMDR$NDUFV1$preds_all %>%
      roc_curve(truth = truth, .pred_prob_gene) %>%
      mutate(Gene = "NDUFV1", 
             AUC = subset(a2m_sep$Res_X.CL_Y.LMDR$NDUFV1$metrics_all, .metric == "roc_auc")$.estimate_gene,
             lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")"))
  ) %>% 
  rbind(
    a2m_sep$Res_X.CL_Y.LMDR$NDUFB8$preds_all %>%
      roc_curve(truth = truth, .pred_prob_gene) %>%
      mutate(Gene = "NDUFB8", 
             AUC = subset(a2m_sep$Res_X.CL_Y.LMDR$NDUFB8$metrics_all, .metric == "roc_auc")$.estimate_gene,
             lbl = paste0(Gene, " (AUC = ", round(AUC,2), ")"))
  )


p3 <- roc_m2a_4p_CL %>% 
  mutate(Gene = factor(Gene, levels = c("NDUFA9", "NDUFV1", "NDUFB8"))) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, color=Gene)) +
  geom_path() +
  geom_abline(lty = 3) +
  scale_color_manual(
    labels = unique(roc_m2a_4p_CL$lbl), 
    values = c("tomato","limegreen", "turquoise"), 
    name="Gene") +
  coord_equal() +
  labs(title="Train: MCSA --> Test: ADNI") +
  theme_bw(base_size = 7)
p4 <- roc_a2m_4p_CL %>% 
  mutate(Gene = factor(Gene, levels = c("NDUFA9", "NDUFV1", "NDUFB8"))) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, color=Gene)) +
  geom_path() +
  geom_abline(lty = 3) +
  scale_color_manual(
    labels = unique(roc_a2m_4p_CL$lbl), 
    values = c("tomato","limegreen", "turquoise"), 
    name="Gene") +
  coord_equal() +
  labs(title="Train: ADNI --> Test: MCSA") +
  theme_bw(base_size = 7)

pdf(paste0(outdir, "/roc_curves.pdf"), width = 8, height = 8)
(p1 + p2)/(p3 + p4)
dev.off()
```


```{r}
sessionInfo()
```



