---
title: "rbt_1_2"
output: html_document
date: "2026-07-30"
---

Author: Wei (Adelyn) Tsai; [tsai.wei\@mayo.edu](mailto:tsai.wei@mayo.edu){.email}.

If you are using this script, please cite our study.

This script addresses comment 2 about cognitive resilience in ROSMAP from reviewer 1.

```{r}
library(tidyverse) #v2.0.0
library(broom) #v1.0.12
library(broom.mixed) #v0.2.9
library(openxlsx) #v4.2.8
library(lmerTest) #v3.1
library(patchwork) #v1.3.2

source("./Codes/fns.R") #source a list of frequently used functions


# list of input directories where files are
indir <- "./indir"

brwgcna_dir <-"./outdir/br_wgcna"
nduf_assoc_dir <- "./outdir/nduf_assoc_de"

# output directory
outdir <- "./Rebuttal_2nd/outdir/1_2"
```

# File preparation
```{r}
comTrait_br <- readRDS(paste0(indir, "/comTrait_br.rds")) # a file with organized AMP-AD metadata
rosmap_id <- read_delim(paste0(indir,"/ROSMAP_id_map.txt"), delim = "\t") %>% mutate(projid = as.character(projid)) # a file mapping between RNAseq SampleID and projid

# Get cognitive scores, global pathology and education from RADC files. 
radc_cs <- read.xlsx(paste0(indir,"/dataset_886_cross-sectional_11-20-2025.xlsx"))
radc_long <- read.xlsx(paste0(indir,"/dataset_886_long_11-20-2025.xlsx"))

radc_use <- comTrait_br %>% 
  filter(Cohort == "ROSMAP") %>% 
  left_join(rosmap_id, by="SampleID") %>% 
  dplyr::select(SampleID, projid, everything()) %>% 
  left_join(radc_cs %>% 
              dplyr::select(projid, gpath, educ, cognep_demog_slope) %>%  #cognep_demog_slope = episodic memory domain score random slope
              dplyr::rename(GlobalPath = gpath), by="projid") %>% 
  left_join(radc_long %>%
              dplyr::select(projid, cogn_ep, age_at_visit) %>% #cogn_ep = episodic memory domain score
              group_by(projid) %>%
              filter(!is.na(cogn_ep)) %>% 
              slice_tail() # take the latest available cogn_ep; 
              )

sum(is.na(radc_use$cogn_ep)) # 17 people without cogn_ep
sum(is.na(radc_use$cognep_demog_slope)) # 36 people without cognep_demog_slope


# read in brain module eigengene file
consMEs_br <- read.csv(paste0(brwgcna_dir , "/MEs1_SP12_Br.csv")) # module eigegenes from consensus brain networks generated in 3_br_wgcna.R

# prepare CQN files
nduf_bb <- read.csv(paste0(nduf_assoc_dir,"/nduf_bb.csv")) # generated in 4_nduf_assoc_de.R; also in supplementary table 12
br_cqn2use <- readRDS(paste0(indir, "/br_cqn2use.rds")) # a list containing cqn of brain datasets. In each list is a dataframe where rows are genes and columns are donors
br_cqn_long_ls <- br_cqn2use %>% map(., ~(pivot_longer(.x, cols = c(4:ncol(.x)), names_to="SampleID", values_to="CQN")))
cqn_nduf <- br_cqn_long_ls$DLPFC %>% 
  filter(gene_id %in% nduf_bb$gene_id) %>% 
  pivot_wider(names_from = "gene_id", values_from = "CQN") %>% 
  dplyr::select(SampleID, any_of(nduf_bb$gene_id)) %>% 
  rename_with(
    ~ nduf_bb$gene_name[match(.x, nduf_bb$gene_id)],
    .cols = any_of(nduf_bb$gene_id)
  )

inputdf <- radc_use %>% 
  left_join(consMEs_br %>% dplyr::select(SubjectID, ME1, ME17, ME26), by = c("SampleID" = "SubjectID")) %>% 
  left_join(cqn_nduf) %>% 
  mutate(Batch = factor(Batch, levels=unique(Batch)),
         Source = factor(Source, levels=unique(Source))
         )

# center and scale the variables
standardize_FUN <- function(x) {
  
  x_med <- median(x, na.rm = TRUE)
  x_mad <- mad(x, na.rm = TRUE)
  x <- ((x - x_med) / x_mad)
  x
  
}
inputdf_scaled <- inputdf %>% 
  mutate_at(vars(Braak:GlobalPath), .funs = standardize_FUN) %>% 
  mutate_at(vars(ME1:NDUFS3), .funs = standardize_FUN)
```

# Functions for regression
```{r}
linear_fn <- function(data_df,
                     XVars,
                     XVars2 = NULL, 
                     YVars, 
                     base_covs,
                     random_var,
                     mode = c("linear", "linear mixed"),
                     interaction = FALSE){
  run_assoc <- function(xvar, yvar, xvar2 = NULL){
    
    if (interaction){
      message(xvar, " x ", xvar2, " and ", yvar)
      main_terms <- paste0(xvar, " * ", xvar2)
    } else {
      message(xvar, " and ", yvar)
      main_terms <- xvar
    }
    
    if (mode == "linear"){
      m <- lm(paste0(yvar, " ~ ", main_terms, " + ", paste0(base_covs, collapse = " + ")), data = data_df)
      res_df <- broom::tidy(m, conf.int = T)
      
    }
    else if (mode == "linear mixed"){
      m <- lmer(paste0(yvar, " ~ ", main_terms, " + ", paste0(base_covs, collapse = " + "), " + ", "(1|", random_var, ")"), data = data_df)
      res_df <- broom.mixed::tidy(m, conf.int = T)
    }
    

    tbl <- tibble(
      Model = list(m),
      SummDf = list(res_df)
    )
    return(tbl)
  }
  
  # Run for all combinations
  if (interaction){

    combos <- expand.grid(
      X = XVars,
      X2 = XVars2,
      Y = YVars,
      stringsAsFactors = FALSE
    ) %>%
      filter(X != X2)  # drop X*X pairs

    results <- combos %>%
      mutate(result = pmap(list(X, Y, X2),
                            ~ run_assoc(..1, ..2, ..3))) %>%
      unnest(result)

  } else {
    results <- expand.grid(
      X = XVars,
      Y = YVars,
      stringsAsFactors = FALSE
    ) %>%
      mutate(result = pmap(list(X, Y),
                            ~ run_assoc(..1, ..2))) %>%
      unnest(result)
  }
  
  return(results)
}
```



# Run associations

```{r run analyses - after scaling}
MTModsRes_scaled <- linear_fn(
  data_df = inputdf_scaled,
  XVars = c("Braak", "Thal", "CERAD", "GlobalPath"),
  YVars = c("cogn_ep", "cognep_demog_slope"), # cogn_ep = last episodic memory; cognep_demog_slope = episodic memory slope 
  XVars2 = c("ME1", "ME17", "ME26"),
  base_covs = c("AOD", "Sex_M1F0", "educ"),
  random_var = NULL,
  mode = "linear",
  interaction = TRUE
) 
MTModsSumm_all_scaled <- MTModsRes_scaled %>% 
  dplyr::select(X, X2, Y, SummDf) %>% 
  unnest(SummDf) 
MTMods_Summ_needed_scaled <- MTModsRes_scaled %>% 
  dplyr::select(X, X2, Y, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(X == term) %>% 
  group_by(X, Y) %>% # adjusting by the number of modules tested
  mutate(q.value = p.adjust(p.value, method = "fdr"), .after = "p.value") %>% 
  bind_rows(
    MTModsRes_scaled %>% 
      dplyr::select(X, X2, Y, SummDf) %>% 
      unnest(SummDf) %>% 
      filter(X2 == term) %>% 
      group_by(X, Y) %>% # adjusting by the number of modules tested
      mutate(q.value = p.adjust(p.value, method = "fdr"), .after = "p.value")
  ) %>% 
  bind_rows(
    MTModsRes_scaled %>% 
      dplyr::select(X, X2, Y, SummDf) %>% 
      unnest(SummDf) %>% 
      filter(grepl(":", term)) %>% 
      group_by(X, Y) %>% # adjusting by the number of modules tested
      mutate(q.value = p.adjust(p.value, method = "fdr"), .after = "p.value")
  ) %>% 
  arrange(Y, X, X2)
# after scaling, ME estimates become positive

NDUFRes_scaled <- linear_fn(
  data_df = inputdf_scaled,
  XVars = c("Braak", "Thal", "CERAD", "GlobalPath"),
  YVars = c("cogn_ep", "cognep_demog_slope"),
  XVars2 = intersect(nduf_bb$gene_name, names(inputdf_scaled)),
  base_covs = c("AOD", "Sex_M1F0", "RIN", "Source", "educ"),
  random_var = "Batch",
  mode = "linear mixed",
  interaction = TRUE
)
NDUFSumm_all_scaled <- NDUFRes_scaled %>% 
  dplyr::select(X, X2, Y, SummDf) %>% 
  unnest(SummDf) 
NDUFSumm_needed_scaled <- NDUFRes_scaled %>% 
  dplyr::select(X, X2, Y, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(X == term) %>% 
  group_by(X, Y) %>% 
  mutate(q.value = p.adjust(p.value, method = "fdr"), .after = "p.value") %>% # adjusting by number of genes tested (n=20)
  dplyr::select(-effect, -group) %>% 
  bind_rows(
    NDUFRes_scaled %>% 
      dplyr::select(X, X2, Y, SummDf) %>% 
      unnest(SummDf) %>% 
      filter(X2 == term) %>% 
      group_by(X, Y) %>% 
      mutate(q.value = p.adjust(p.value, method = "fdr"), .after = "p.value") %>% # adjusting by number of genes tested (n=20)
      dplyr::select(-effect, -group)
  ) %>% 
  bind_rows(
    NDUFRes_scaled %>% 
      dplyr::select(X, X2, Y, SummDf) %>% 
      unnest(SummDf) %>% 
      filter(grepl(":", term)) %>% 
      group_by(X, Y) %>% 
      mutate(q.value = p.adjust(p.value, method = "fdr"), .after = "p.value") %>% # adjusting by number of genes tested (n=20)
      dplyr::select(-effect, -group)
  ) %>% 
  arrange(Y, X, X2)

# save the summary dataframe
write.xlsx(
  list("MTModules_all" = MTModsSumm_all_scaled, 
       "MTModules_needed" = MTMods_Summ_needed_scaled,
       "NDUFs_all" = NDUFSumm_all_scaled,
       "NDUFs_needed" = NDUFSumm_needed_scaled),
  paste0(outdir, "/ROSMAP_resilience_Association.xlsx"))

# save the results
save(MTModsRes_scaled, NDUFRes_scaled, file = paste0(outdir, "/results.RData"))
```

# Plot 
p1 and p2 are Fig2.c; p3 and p4 are Extended Data Fig.11a
```{r plot mitochondrial modules results}
plotdf1 <- MTMods_Summ_needed_scaled %>% 
  mutate(X2 = str_remove(X2, "E"),
         X2 = str_c("Br_", X2),
         X = factor(X, levels = c("Braak", "Thal", "CERAD", "GlobalPath")),
         Y = ifelse(Y == "cogn_ep", "Last episodic memory", "Episodic memory slope"),
         Y = factor(Y, levels = c("Last episodic memory", "Episodic memory slope")),
         lbl = if_else(q.value<0.05, "\u2731", ""), 
         lbl = if_else(p.value<0.05 & q.value>0.05,"Δ", lbl))


p1 <- plotdf1 %>% 
  filter(grepl("^ME", term) & X != "CERAD" & Y == "Last episodic memory") %>% 
  ggplot(aes(x = X, y = X2)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt, family = "Arial") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  scale_y_discrete(limits = rev) +
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Association with\nlast episodic memory",
    y = "Brain mitochondrial consensus module",
    x = "Adjusted pathology",
    caption = "Δ:P<0.05\n\u2731:q<0.05") +
  guides(size = guide_legend(order = 1)) +
  theme_minimal(base_size = 6.5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5))
p2 <- plotdf1 %>% 
  filter(grepl("^ME", term) & X != "CERAD" & Y == "Episodic memory slope") %>% 
  ggplot(aes(x = X, y = X2)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt, family = "Arial") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  scale_y_discrete(limits = rev) +
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Association with\nepisodic memory slope",
    y = "Brain mitochondrial consensus module",
    x = "Adjusted pathology",
    caption = "Δ:P<0.05\n\u2731:q<0.05") +
  guides(size = guide_legend(order = 1)) +
  theme_minimal(base_size = 6.5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5))
p3 <- plotdf1 %>% 
  filter(grepl(":", term) & X != "CERAD" & Y == "Last episodic memory") %>% 
  ggplot(aes(x = X, y = X2)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt, family = "Arial") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  scale_y_discrete(limits = rev) +
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Association with\nlast episodic memory",
    subtitle = "Effect of interaction term\n(Br_ME x pathology)",
    y = "Brain mitochondrial consensus module",
    x = "Adjusted pathology",
    caption = "Δ:P<0.05\n\u2731:q<0.05") +
  guides(size = guide_legend(order = 1)) +
  theme_minimal(base_size = 6.5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
p4 <- plotdf1 %>% 
  filter(grepl(":", term) & X != "CERAD" & Y == "Episodic memory slope") %>% 
  ggplot(aes(x = X, y = X2)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt, family = "Arial") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  scale_y_discrete(limits = rev) +
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Association with\nepisodic memory slope",
    subtitle = "Effect of interaction term\n(Br_ME x pathology)",
    y = "Brain mitochondrial consensus module",
    x = "Adjusted pathology",
    caption = "Δ:P<0.05\n\u2731:q<0.05") +
  guides(size = guide_legend(order = 1)) +
  theme_minimal(base_size = 6.5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

cairo_pdf(paste0(outdir, "/MTModsCRCDR_hm_noCERAD.pdf"), width = 4, height = 5)
(p1+p2)/(p3+p4)
dev.off() 


```

p1 and p2 are Fig.4b; p3 and p4 are Extended Data Fig.11b
```{r plot NDUFs results}

plotdf2 <- NDUFSumm_needed_scaled %>% 
  mutate(X2 = str_remove(X2, "E"),
         X = factor(X, levels = c("Braak", "Thal", "CERAD", "GlobalPath")),
         Y = ifelse(Y == "cogn_ep", "Last episodic memory", "Episodic memory slope"),
         Y = factor(Y, levels = c("Last episodic memory", "Episodic memory slope")),
         lbl = if_else(q.value<0.05, "\u2731", ""), 
         lbl = if_else(p.value<0.05 & q.value>0.05,"Δ", lbl))

p1 <- plotdf2 %>% 
  filter(grepl("^NDUF", term) & X != "CERAD" & Y == "Last episodic memory") %>% 
  ggplot(aes(x = X, y = X2)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt, family = "Arial") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Association with\nlast episodic memory",
    y = "",
    x = "Adjusted pathology",
    caption = "Δ:P<0.05\n\u2731:q<0.05") +
  guides(size = guide_legend(order = 1)) +
  theme_minimal(base_size = 6.5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        axis.text.y = element_text(face = "italic"),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5))
p2 <- plotdf2 %>% 
  filter(grepl("^NDUF", term) & X != "CERAD" & Y == "Episodic memory slope") %>% 
  ggplot(aes(x = X, y = X2)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt, family = "Arial") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Association with\nepisodic memory slope",
    y = "",
    x = "Adjusted pathology",
    caption = "Δ:P<0.05\n\u2731:q<0.05") +
  #facet_grid(.~ type) +
  guides(size = guide_legend(order = 1)) +
  theme_minimal(base_size = 6.5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        axis.text.y = element_text(face = "italic"),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5))
p3 <- plotdf2 %>% 
  filter(grepl(":", term) & X != "CERAD" & Y == "Last episodic memory") %>% 
  ggplot(aes(x = X, y = X2)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt, family = "Arial") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Association with\nlast episodic memory",
    subtitle = "Effect of interaction term\n(NDUF x pathology)",
    y = "",
    x = "Adjusted pathology",
    caption = "Δ:P<0.05\n\u2731:q<0.05") +
  guides(size = guide_legend(order = 1)) +
  theme_minimal(base_size = 6.5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        axis.text.y = element_text(face = "italic"),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
p4 <- plotdf2 %>% 
  filter(grepl(":", term) & X != "CERAD" & Y == "Episodic memory slope") %>% 
  ggplot(aes(x = X, y = X2)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=5/.pt, family = "Arial") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  labs(
    title = "Association with\nepisodic memory slope",
    subtitle = "Effect of interaction term\n(NDUF x pathology)",
    y = "",
    x = "Adjusted pathology",
    caption = "Δ:P<0.05\n\u2731:q<0.05") +
  guides(size = guide_legend(order = 1)) +
  theme_minimal(base_size = 6.5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        axis.text.y = element_text(face = "italic"),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

cairo_pdf(paste0(outdir, "/NDUFsCRCDR_hm_noCERAD.pdf"), width = 4.5, height = 10)
(p1 + p2)/(p3 + p4)
dev.off()
```


This example plot is in second rebuttal document in Rebuttal Fig.1c
```{r visual example plot}
model = MTModsRes_scaled$Model[[9]]
tertiles <- quantile(inputdf_scaled$ME26, probs = c(0, 0.5, 1))
p1 <- sjPlot::plot_model(model,
             type = "pred",
             terms = c("Braak", paste0("ME26 [", paste(round(tertiles, 2), collapse = ","), "]"))) +
    scale_colour_brewer(palette = "Set1", aesthetics = c('colour', 'fill'),
                        labels = c("Minimum", "Medium (50%)", "Maximum")
                        )  +
    sjPlot::theme_sjplot() +
    labs(title = "Effect of Br_M26 expression on the\nrelationship between episodic memory & Braak",
         y = "last episodic memory",
         color = "Br_M26") 

pdf(paste0(outdir, "/visual_ex.pdf"), width = 6, height = 6)
p1
dev.off()
```


```{r}
sessionInfo()
```

