---
title: "rbt_1_4"
output: html_document
date: "2026-08-06"
---
Author: Wei (Adelyn) Tsai; [tsai.wei\@mayo.edu](mailto:tsai.wei@mayo.edu){.email}.

If you are using this script, please cite our study.

This script addresses comment 4 about blood cell composition from reviewer 1.
For additional cross-sectional and longitudinal associations results, they were addressed in the first rebuttal, in rebuttal/rbt_1_4_cbx_LM10.Rmd.

```{r prep}
library(tidyverse) #v2.0.0
library(patchwork) #v1.3.2
library(openxlsx) #v4.2.8
library(lmerTest) #v3.1
library(broom.mixed) #v0.2.9
library(broom) #v1.0.12
library(WGCNA) #v1.72
library(ggpubr) # v0.6.3
library(pbapply) #v1.7

set.seed(123)

# list of input/output directories
indir <- "./indir"
blwgcna_dir <- "./outdir/bl_wgcna"
nduf_assoc_dir <- "./outdir/nduf_assoc_de"
residir <- "./outdir/resids"
rbt_v1_dir <- "./Rebuttal/outdir" # directory to the results from the first rebuttal

# output directory
outdir <- "./Rebuttal_2nd/outdir/1_4"

source("./Codes/fns.R") #source a list of frequently used functions
```

# Processing residuals files
Residuals were generated using rbt_1_4_genResids_byCT.R and rbt_1_4_genResids_allCT.R
```{r}
# MCSA
mcsa_res_list <- list.files(path = paste0(outdir, "/resids/MCSA"), pattern = "\\.rds$", full.names = TRUE) %>% 
  set_names(basename(.) %>% str_remove("\\.rds$")) %>% 
  map(readRDS)

# ADNI
adni_res_list <- list.files(path = paste0(outdir, "/resids/ADNI"), pattern = "\\.rds$", full.names = TRUE) %>% 
  set_names(basename(.) %>% str_remove("\\.rds$")) %>% 
  map(readRDS)

consNet_object <- readRDS(paste0(blwgcna_dir, "/consNet_object_2023-08-15.rds")) # from 2_bl_wgcna.R
genes_use <- names(consNet_object$consNet$colors)
## filter the residuals to only those used to construct the original WGCNA
mcsa_res_use <- mcsa_res_list %>% 
  map(., ~ filter(.x, GeneId %in% genes_use)) %>% 
  map(., ~ pivot_wider(.x, names_from = "GeneId", values_from = ".resid")) %>% 
  map(., ~ column_to_rownames(.x, "ptnum")) %>% 
  map(., ~ data.matrix(.x)) 

adni_res_use <- adni_res_list %>% 
  map(., ~ column_to_rownames(.x, "SubjectID")) %>% 
  map(., ~ dplyr::select(.x, matches(genes_use))) %>% 
  map(., ~ data.matrix(.x))

comRes_ls <- list()
for(i in 1:length(mcsa_res_use)){
  name <- names(mcsa_res_use)[[i]]
  name <- sub(".*_", "", name)
  
  comRes_ls[[name]] <- rbind(mcsa_res_use[[i]], adni_res_use[[i]])
}
```

# Calculate module eigengenes
```{r}

ME_og <- read.csv(paste0(blwgcna_dir, "/MEs1_SP12.csv")) # from 2_bl_wgcna.R
ME_ls <- list()
for(i in names(comRes_ls)){

  ME_ls[[i]][["MCSA"]] <- moduleEigengenes(comRes_ls[[i]][1:105,], consNet_object$consNet$colors)$eigengenes
  ME_ls[[i]][["ADNI"]] <- moduleEigengenes(comRes_ls[[i]][106:196,], consNet_object$consNet$colors)$eigengenes
  
}

consME_ls <- ME_ls %>% map(reduce, rbind) %>% 
  map(., ~ rownames_to_column(.x, "SubjectID")) %>% 
  map(., ~ left_join(.x, ME_og %>% dplyr::select(SubjectID, ME3) %>% rename(ME3_no_adj = ME3)))

M3_corr_p_ls <- list()
for (i in names(consME_ls)){
  M3_corr_p_ls[[i]] <- 
    consME_ls[[i]] %>% 
    mutate(Cohort = c(rep("MCSA", 105), rep("ADNI", 91)),
           Cohort = factor(Cohort, levels = c("MCSA", "ADNI"))) %>% 
    ggplot(aes(x = ME3_no_adj, y = ME3)) +
    geom_point(aes(color = Cohort), size = 1.2, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
    ggpubr::stat_cor(method = "spearman", 
                     size = 5/.pt, 
                     label.sep = "\n") + 
    facet_grid(. ~ Cohort) +
    theme_bw(base_size = 6.7) +
    labs(
      title = paste0("Adjusted for ", i, " proportion"),
      x = "Bl_M3 (no cell proportion adjusted)",
      y = "Bl_M3 (Cell proportion adjusted)"
    ) +
    theme(legend.position = "none")
}
M3_corr_p <- wrap_plots(M3_corr_p_ls, nrow = 2)

pdf(paste0(outdir, "/M3_corr_p.pdf"), width = 8, height = 4)
M3_corr_p # this is in Extended Data Fig.13b
dev.off()
```

# Prepare files for running association
```{r}
comTrait <- readRDS(paste0(indir,"/comTrait.rds")) # a file having cross-sectional endophenotypes for MCSA (from row 1-105) and ADNI (from row 106-196); row names are donorID

M3_df <- comTrait %>% 
  rownames_to_column("SubjectID") %>% 
  dplyr::select(SubjectID, Sex, Educ, Age, Memory, LMDR, HippVol, Hipp_ICV, Mag_HippICV) %>% 
  left_join(ME_og %>% dplyr::select(SubjectID, ME3)) %>% # joining with the original ME 
  rename(None_ME3 = ME3) # rename the original ME

for (i in names(consME_ls)){
  M3_df <- M3_df %>% left_join(consME_ls[[i]] %>% dplyr::select(SubjectID, ME3) %>% rename("{i}_ME3" := ME3))
}
```


# Functions for regression
```{r}
linear_fn <- function(data_df,
                     XVars,
                     XVars2 = NULL, 
                     YVars, 
                     base_covs,
                     random_var,
                     mode = c("linear", "linear mixed"),
                     interaction = FALSE){
  run_assoc <- function(xvar, yvar, xvar2 = NULL){
    
    if (interaction){
      message(xvar, " x ", xvar2, " and ", yvar)
      main_terms <- paste0(xvar, " * ", xvar2)
    } else {
      message(xvar, " and ", yvar)
      main_terms <- xvar
    }
    
    if (mode == "linear"){
      m <- lm(paste0(yvar, " ~ ", main_terms, " + ", paste0(base_covs, collapse = " + ")), data = data_df)
      res_df <- broom::tidy(m, conf.int = T)
      
    }
    else if (mode == "linear mixed"){
      m <- lmer(paste0(yvar, " ~ ", main_terms, " + ", paste0(base_covs, collapse = " + "), " + ", "(1|", random_var, ")"), data = data_df)
      res_df <- broom.mixed::tidy(m, conf.int = T)
    }
    

    tbl <- tibble(
      Model = list(m),
      SummDf = list(res_df)
    )
    return(tbl)
  }
  
  # Run for all combinations
  if (interaction){

    combos <- expand.grid(
      X = XVars,
      X2 = XVars2,
      Y = YVars,
      stringsAsFactors = FALSE
    ) %>%
      filter(X != X2)  # drop X*X pairs

    results <- combos %>%
      mutate(result = pmap(list(X, Y, X2),
                            ~ run_assoc(..1, ..2, ..3))) %>%
      unnest(result)

  } else {
    results <- expand.grid(
      X = XVars,
      Y = YVars,
      stringsAsFactors = FALSE
    ) %>%
      mutate(result = pmap(list(X, Y),
                            ~ run_assoc(..1, ..2))) %>%
      unnest(result)
  }
  
  return(results)
}
```

# Run association
```{r}
yvariables <- c("None_ME3", paste0(names(consME_ls), "_ME3"))

mcsa_cog <- linear_fn(data_df = M3_df[1:105, ],
                      XVars = c("Memory", "LMDR"),
                      YVars = yvariables,
                      base_covs = c("Age", "Sex", "Educ"),
                      mode = "linear"
                      )
mcsa_hv <- linear_fn(data_df = M3_df[1:105, ],
                      XVars = c("HippVol"),
                      YVars = yvariables,
                      base_covs = c("Age", "Sex", "Educ", "Hipp_ICV"),
                      mode = "linear"
                      )

adni_cog <- linear_fn(data_df = M3_df[106:196, ],
                      XVars = c("Memory", "LMDR"),
                      YVars = yvariables,
                      base_covs = c("Age", "Sex", "Educ"),
                      mode = "linear"
                      )
adni_hv <- linear_fn(data_df = M3_df[106:196, ],
                      XVars = c("HippVol"),
                      YVars = yvariables,
                      base_covs = c("Age", "Sex", "Educ", "Hipp_ICV", "Mag_HippICV"),
                      mode = "linear"
                      )
save(mcsa_cog, mcsa_hv, adni_cog, adni_hv, file = paste0(outdir, "/CBX_CPAdj_M3_Assoc.RData"))
```

# Cross-sectional
## Join with previous results
```{r}
load(paste0(rbt_v1_dir, "/1_4/CBX_LM10/CBX_CPAdj_Assoc.RData"), verbose = T) # from rebuttal/rbt_1_4_cbx_LM10.Rmd

# need to rename the columns of the current results to rbind with the previous results 
mcsa_m3 <- mcsa_cog %>% 
  rbind(mcsa_hv) %>% 
  separate_wider_delim(
    cols = Y,
    delim = "_",
    names = c("ADJUST", "YVar")
  ) %>% 
  rename(XVar = X) %>% 
  mutate(YVar = "Bl_M3",
         ADJUST = ifelse(ADJUST == "allCT", "All", ADJUST))
adni_m3 <- adni_cog %>% 
  rbind(adni_hv) %>% 
  separate_wider_delim(
    cols = Y,
    delim = "_",
    names = c("ADJUST", "YVar")
  ) %>% 
  rename(XVar = X) %>% 
  mutate(YVar = "Bl_M3",
         ADJUST = ifelse(ADJUST == "allCT", "All", ADJUST))

mcsa_all <- rbind(mcsa_m3, mcsa_assoc) # mcsa_assoc loaded from previous results
adni_all <- rbind(adni_m3, adni_assoc) # adni_assoc loaded from previous results

mcsa_assoc_df <- mcsa_all %>% 
  dplyr::select(XVar, YVar, ADJUST, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, XVar)) %>% 
  group_by(XVar, ADJUST) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for number of genes/ME x adjustment tested (number of genes/ME = 21, adjustment = no adjustment + 8 cell types = 9, so overall = 21*9)
  ungroup() %>% 
  mutate(XVar = factor(XVar, levels = c("Memory", "LMDR", "HippVol"))) %>% 
  arrange(YVar, XVar) %>% 
  group_by(YVar, XVar) %>% 
  mutate(pct_attenuation = (estimate[1] - estimate)/estimate[1]*100, .after = "estimate") %>% 
  mutate(pct_attenuation = ifelse(pct_attenuation == pct_attenuation[1], NA, pct_attenuation))
adni_assoc_df <- adni_all %>% 
  dplyr::select(XVar, YVar, ADJUST, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, XVar)) %>% 
  group_by(XVar, ADJUST) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for number of genes/ME x adjustment tested (number of genes/ME = 21, adjustment = no adjustment + 8 cell types = 9, so overall = 21*9)
  ungroup() %>% 
  mutate(XVar = factor(XVar, levels = c("Memory", "LMDR", "HippVol"))) %>% 
  arrange(YVar, XVar) %>% 
  group_by(YVar, XVar) %>% 
  mutate(pct_attenuation = (estimate[1] - estimate)/estimate[1]*100, .after = "estimate") %>% 
  mutate(pct_attenuation = ifelse(pct_attenuation == pct_attenuation[1], NA, pct_attenuation))

```

## Meta-analysis
```{r}
meta_df = inner_join(mcsa_assoc_df, adni_assoc_df, by = c("XVar","YVar", "ADJUST", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(rn = paste0(XVar, "_", YVar, "_", ADJUST)) %>% 
  column_to_rownames("rn")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "rn")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "rn") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "rn")

# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_CP_final = meta_df %>%
  group_by(XVar, ADJUST) %>%
  mutate(
    q_local_random = p.adjust(p_random, method = "fdr"),
    q_local_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(XVar) %>% 
  mutate(
    q_global_random = p.adjust(p_random, method = "fdr"),
    q_global_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for numer of genes x adjustment tested (21*9)
  ungroup() %>% 
  mutate(model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_local_meta = ifelse(model_meta =="Fixed", q_local_fixed, q_local_random),
         q_global_meta = ifelse(model_meta =="Fixed", q_global_fixed, q_global_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random)
         ) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"))) %>% 
  dplyr::select(-rn, -term) %>% 
  group_by(YVar, XVar) %>% 
  mutate(pct_attenuation_meta = (beta_meta[1] - beta_meta)/beta_meta[1]*100, .after = "beta_meta") %>% 
  mutate(pct_attenuation_meta = ifelse(pct_attenuation_meta == pct_attenuation_meta[1], NA, pct_attenuation_meta))

write.xlsx(meta_CP_final, paste0(outdir, "/CBX_meta_CPAdj_final.xlsx")) # the organized table of this is in Supplementary Table 13a

rm(meta_df, fix_res, rand_res, het_res, meta_raw, eff_idx1, eff_idx2, err_idx1, err_idx2)
```

```{r plot the results with or without adjustment}
# meta-analyzed results
plot_df <- meta_CP_final %>%
  dplyr::select(XVar:estimate_MCSA, p.value_MCSA:q_global_MCSA, estimate_ADNI, p.value_ADNI:q_global_ADNI, beta_meta, p_meta:q_global_meta) %>% 
  dplyr::rename(
    estimate_meta = beta_meta,
    p.value_meta = p_meta,
    ) %>%
  pivot_longer(
    cols = ends_with(c("_MCSA" , "_ADNI" , "_meta")),
    names_to = "name",
    values_to = "value"
  ) %>%
  separate(
    name,
    into = c("stat", "Cohort"),
    sep = "_(?=[^_]+$)",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
  mutate(XVar=factor(XVar, levels=c("Memory", "LMDR", "HippVol")),
         ADJUST = factor(ADJUST, levels = c("None", "All", "Bcells", "CD4T", "CD8T", "NK", "Monocytes", "Mast", "Neutrophils")),
         Cohort = ifelse(Cohort=="meta", "Meta-analyzed", Cohort),
         Cohort = factor(Cohort, levels=c("MCSA", "ADNI","Meta-analyzed")),
         lbl = if_else(q_global<0.05, "✱✱",  ""), 
         lbl = if_else(q_local<0.05 & q_global > 0.05, "✱", lbl), 
         lbl = if_else(p.value<0.05 & q_local>0.05,"Δ", lbl)
         )

cairo_pdf(paste0(outdir, "/CBX_meta_assoc_adjustbyCP.pdf"), width = 10, height = 6.5)
plot_df %>% 
  filter(Cohort == "Meta-analyzed") %>% 
  ggplot(aes(x=ADJUST, y=YVar)) +
  geom_point(shape=21, color="black", aes(fill=estimate, size = -log(p.value))) +
  geom_text(aes(label=lbl), size=9/.pt) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name=expression(beta))+
  #scale_size_continuous(range = c(2,10)) + 
  scale_y_discrete(limits = rev) +
  labs(
    x = "Adjustment",
    caption = "Δ:P<0.05; ✱:local q<0.05; ✱✱:global q<0.05") +
  facet_grid(.~XVar) +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic"),
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.title.y = element_blank(),
        strip.background = element_rect(colour = "black"),
        legend.key.size = unit(0.1,"in"),
        text = element_text(family = "ArialMT"))
dev.off() # This is Extended Data Fig.14a
```


# Longitudinal
## Calculate % attenuation
```{r}
load(paste0(rbt_v1_dir, "/1_4/CBX_LM10/CBX_CPAdj_LongAssoc.RData"), verbose = T) # from rebuttal/rbt_1_4_cbx_LM10.Rmd

mcsa_longassoc_df <- mcsa_longassoc %>% 
  dplyr::select(XVar, YVar, ADJUST, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, ":")) %>% 
  group_by(YVar, ADJUST) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes tested (n=20)
  ungroup() %>% 
  group_by(YVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for number of genes x adjustment tested (number of genes = 20, adjustment = no adjustment + 8 cell types = 9, so overall = 20*9)
  ungroup() %>% 
  dplyr::select(-effect, -group) %>% 
  mutate(YVar = factor(YVar, levels = c("Memory", "LMDR", "HippVol"))) %>% 
  arrange(XVar, YVar) %>% 
  group_by(XVar, YVar) %>% 
  mutate(pct_attenuation = (estimate[1] - estimate)/estimate[1]*100, .after = "estimate") %>% 
  mutate(pct_attenuation = ifelse(pct_attenuation == pct_attenuation[1], NA, pct_attenuation))

adni_longassoc_df <- adni_longassoc %>% 
  dplyr::select(XVar, YVar, ADJUST, SummDf) %>% 
  unnest(SummDf) %>% 
  filter(str_detect(term, ":")) %>% 
  group_by(YVar, ADJUST) %>% 
  mutate(q_local = p.adjust(p.value, "fdr"), .after = "p.value") %>% # this adjusts for number of genes tested (n=20)
  ungroup() %>% 
  group_by(YVar) %>% 
  mutate(q_global = p.adjust(p.value, "fdr"), .after = "q_local") %>% # this adjusts for number of genes x adjustment tested (number of genes = 20, adjustment = no adjustment + 8 cell types = 9, so overall = 20*9)
  ungroup() %>%
  dplyr::select(-effect, -group) %>% 
  mutate(YVar = factor(YVar, levels = c("Memory", "LMDR", "HippVol"))) %>% 
  arrange(XVar, YVar) %>% 
  group_by(XVar, YVar) %>% 
  mutate(pct_attenuation = (estimate[1] - estimate)/estimate[1]*100, .after = "estimate") %>% 
  mutate(pct_attenuation = ifelse(pct_attenuation == pct_attenuation[1], NA, pct_attenuation))
```

## conduct meta-analysis then calculate % attenuation
```{r}
meta_df = inner_join(mcsa_longassoc_df, adni_longassoc_df, by = c("XVar","YVar", "ADJUST", "term"), suffix = c("_MCSA", "_ADNI")) %>% 
  mutate(rn = paste0(XVar, "_", YVar, "_", ADJUST)) %>% 
  column_to_rownames("rn")

# this indicates column indices corresponding to estimates and standard error of MCSA and ADNI
eff_idx1 = which(colnames(meta_df) == "estimate_MCSA")
eff_idx2 = which(colnames(meta_df) == "estimate_ADNI")
err_idx1 = which(colnames(meta_df) == "std.error_MCSA")
err_idx2 = which(colnames(meta_df) == "std.error_ADNI")

# run meta-analysis using the function meta_gen_fn defined above
meta_raw = meta_df %>% pbapply(1, meta_gen_fn, eff_idx1 = eff_idx1, eff_idx2 = eff_idx2, err_idx1 = err_idx1, err_idx2 = err_idx2)

# Get all fixed-effect results
fix_res = meta_raw %>%
  pbsapply(function(x) x$fixed) %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~paste0(.x, "_fixed")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>% 
  mutate_all(.funs = ~unname(.x))

# Get all random-effect results
rand_res = meta_raw %>%
  pbsapply(function(x) x$random) %>%
  t() %>%
  data.frame() %>%
  rename_with(~paste0(.x, "_random")) %>%
  rownames_to_column(var = "rn") %>%
  mutate_all(.funs = ~unlist(.x)) %>%
  mutate_all(.funs = ~unname(.x))

# Get heterogenity meassurments Q, I2, tau2
het_res = meta_raw %>%
  map(~.x[c("Q","df.Q","pval.Q","I2", "lower.I2", "upper.I2", "tau2", "se.tau2", "H", "lower.H", "upper.H")]) %>%
  bind_rows(.id = "rn")

# Join the results to the original regression coeffcients
meta_df = meta_df %>%
  rownames_to_column(var = "rn") %>% 
  inner_join(fix_res) %>%
  inner_join(rand_res) %>% 
  rename_with(~sub("^TE", "beta", .x), everything()) %>%
  rename_with(~sub("^seTE", "se", .x), everything()) %>%
  inner_join(het_res, by = "rn")

# final meta-analyze dataframe, including adjusting for multiple test correction, determining meta-analyzed beta, p and q-values based on heterogeneity measures I2
meta_CP_long_final = meta_df %>%
  group_by(YVar, ADJUST) %>%
  mutate(
    q_local_random = p.adjust(p_random, method = "fdr"),
    q_local_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for number of genes/ME tested (n=21)
  ungroup() %>% 
  group_by(YVar) %>% 
  mutate(
    q_global_random = p.adjust(p_random, method = "fdr"),
    q_global_fixed = p.adjust(p_fixed, method = "fdr")
  ) %>% # this adjusts for numer of genes x adjustment tested (21*9)
  ungroup() %>% 
  mutate(model_meta = ifelse(I2 < 0.25, "Fixed", "Random"),
         beta_meta = ifelse(model_meta =="Fixed", beta_fixed, beta_random),
         se_meta = ifelse(model_meta =="Fixed", se_fixed, se_random),
         p_meta = ifelse(model_meta =="Fixed", p_fixed, p_random),
         q_local_meta = ifelse(model_meta =="Fixed", q_local_fixed, q_local_random),
         q_global_meta = ifelse(model_meta =="Fixed", q_global_fixed, q_global_random),
         CI_L_meta = ifelse(model_meta == "Fixed", lower_fixed, lower_random),
         CI_H_meta = ifelse(model_meta =="Fixed", upper_fixed, upper_random)
  ) %>%
  dplyr::select(-c(ends_with("_random"), ends_with("_fixed"))) %>% 
  dplyr::select(-rn, -term) %>% 
  group_by(YVar, XVar) %>% 
  mutate(pct_attenuation_meta = (beta_meta[1] - beta_meta)/beta_meta[1]*100, .after = "beta_meta") %>% 
  mutate(pct_attenuation_meta = ifelse(pct_attenuation_meta == pct_attenuation_meta[1], NA, pct_attenuation_meta))

rm(meta_df, fix_res, rand_res, het_res, meta_raw, eff_idx1, eff_idx2, err_idx1, err_idx2)

write.xlsx(meta_CP_long_final, paste0(outdir, "/CBX_meta_CPAdj_long_final.xlsx")) # the organized table of this is in Supplementary Table 13c
```

